---
title: "DLB Cohort Analysis — CSF α-synuclein SAA"
subtitle: "Clinical, biofluid and imaging profile by seed-amplification assay status"
author: "E. Bernardi"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: true
    code_folding: hide
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.align = "center", dpi = 150)

library(readxl);  library(dplyr);   library(tidyr);   library(lubridate)
library(stringr); library(purrr);   library(tibble)
library(gtsummary); library(gt);    library(ggplot2); library(knitr)
library(kableExtra)

theme_set(theme_minimal(base_size = 12))
set.seed(20260826)
```

```{r config}
## ---------------------------------------------------------------------------
## PATHS
## ---------------------------------------------------------------------------
desktop_path <- "~/Desktop/Ricerca/Lewy_R/SAA"
results_path <- file.path(desktop_path, "Results")
dir.create(results_path, recursive = TRUE, showWarnings = FALSE)

## Keep every figure as a file as well. html_document is self-contained by
## default, so plots are embedded in the HTML as base64 and rmarkdown then
## deletes the temporary figure folder -- leaving nothing to attach to an email.
## Pointing fig.path inside Results survives that cleanup.
fig_dir <- file.path(results_path, "figures")
dir.create(fig_dir, recursive = TRUE, showWarnings = FALSE)
knitr::opts_chunk$set(fig.path = paste0(fig_dir, "/"))

f_clin  <- file.path(desktop_path, "datafreeze_lewy.xlsx")
f_bio   <- file.path(desktop_path, "BIOFLUIDS_CLINICAL.xlsx")


## ---------------------------------------------------------------------------
## MERGE OPTIONS
## ---------------------------------------------------------------------------
MAX_GAP_DAYS <- 365   # visit <-> sample; pairs beyond this are flagged

## Restrict the analysis cohort to probable DLB (Inigo). Set to FALSE to keep
## every Lewy diagnosis; the diagnosis breakdown is printed either way.
PROBABLE_ONLY <- TRUE
## "MCI-LB probable" is probable Lewy body disease at the prodromal stage, not
## DLB dementia. TRUE keeps those patients as well; FALSE restricts to dementia.
## iRBD, possible DLB and MCI-LB possible stay out of the cohort either way.
INCLUDE_MCI_LB_PROBABLE <- TRUE

## ---------------------------------------------------------------------------
## DATA-QUALITY GUARDS
## Values outside these ranges are set to NA and reported in section 2.
## Kept as a regression check: an earlier run produced Mayo = 21 (SD 62) on a
## 0-4 scale, which turned out to come from the baseline-date row selection
## rather than from sentinel codes. If this table is ever non-empty, the source
## column is wrong.
## ---------------------------------------------------------------------------
SENTINELS <- c(-9, -99, -999, 888, 999, 9999, -1)

PLAUSIBLE <- tribble(
  ~var,                    ~lo, ~hi,
  "edad",                    18, 110,
  "EDUC",                     0,  30,
  "puntuaciones",             0,  30,   # MoCA total
  "visuoespacialp",           0,   5,
  "identificacionp",          0,   3,
  "atencionp",                0,   6,
  "lenguajep",                0,   3,
  "abstraccionp",             0,   2,
  "orientacionp",             0,   6,
  "mis",                      0,  15,   # MoCA memory index
  "mayo_total_score",         0,   4,   # 4 dichotomous items
  "updrs1a_total_score",      0,  24,
  "updrs1b_total_score",      0,  28,
  "updrs2_total_score",       0,  52,
  "updrs3_total_score",       0, 132,
  "hoehn_yahr",               0,   5,
  "sppb_total_score",         0,  12,
  "sppb_subtotal1",           0,   4,
  "sppb_subtotal2",           0,   4,
  "sppb_subtotal3",           0,   4,
  "mta",                      0,   4,
  "fazekas",                  0,   3,
  "MMSE",                     0,  30,
  "GDS",                      1,   7,
  "gds_visit",                1,   7
)

```

# Data loading and merge

```{r load}
## ---------------------------------------------------------------------------
## guess_max = Inf is essential. readxl types each column from the first 1000
## rows; in a long REDCap export the visit-level columns (UPDRS, MoCA, SPPB)
## are empty on the leading enrolment rows and would be typed <logical>,
## silently discarding every value below row 1000.
## ---------------------------------------------------------------------------
read_x <- function(p) read_excel(p, guess_max = Inf, .name_repair = "minimal")

clin_raw <- read_x(f_clin)
bio_raw  <- read_x(f_bio)

## drop blank / duplicated column names produced by Excel
tidy_names <- function(df) {
  df <- df[, names(df) != "" & !is.na(names(df)), drop = FALSE]
  df[, !duplicated(names(df)), drop = FALSE]
}
clin_raw <- tidy_names(clin_raw); bio_raw <- tidy_names(bio_raw)

## ---- helpers --------------------------------------------------------------
norm_id <- function(x) {                       # 0012345 == 12345
  x <- toupper(trimws(as.character(x)))
  x[x %in% c("", "NA", "NULL", "-", "0")] <- NA
  x <- gsub("[^0-9A-Z]", "", x)
  num <- !is.na(x) & grepl("^[0-9]+$", x)
  x[num] <- sub("^0+", "", x[num])
  x[!is.na(x) & x == ""] <- NA
  x
}

as_date_safe <- function(x) {                  # Date / POSIXct / serial / text
  if (inherits(x, "Date"))   return(x)
  if (inherits(x, "POSIXt")) return(as.Date(x))
  if (is.numeric(x)) {
    out <- as.Date(rep(NA, length(x)))
    ok <- !is.na(x) & x > 10000 & x < 60000
    out[ok] <- as.Date(x[ok], origin = "1899-12-30"); return(out)
  }
  x <- trimws(as.character(x))
  x[x %in% c("", "NA", "NaN", "NULL", "-", "#N/A")] <- NA
  out <- as.Date(rep(NA, length(x)))
  ser <- !is.na(x) & grepl("^[0-9]{5}(\\.[0-9]+)?$", x)
  out[ser] <- as.Date(as.numeric(x[ser]), origin = "1899-12-30")
  for (f in c("%Y-%m-%d","%d/%m/%Y","%d-%m-%Y","%Y/%m/%d","%d.%m.%Y","%Y%m%d")) {
    todo <- !is.na(x) & is.na(out)
    if (!any(todo)) break
    out[todo] <- as.Date(substr(x[todo], 1, 10), format = f)
  }
  out[!is.na(out) & (out < as.Date("1900-01-01") | out > Sys.Date() + 365)] <- NA
  out
}

num_safe <- function(x) {                      # "<200", "1,5", " 12 "
  if (is.numeric(x)) return(x)
  v <- gsub("\\s", "", gsub("[<>=]", "", trimws(as.character(x))))
  v <- ifelse(grepl(",", v) & !grepl("\\.", v), gsub(",", ".", v), gsub(",", "", v))
  suppressWarnings(as.numeric(v))
}

has_col <- function(df, x) x %in% names(df)
`%||%` <- function(a, b) if (is.null(a)) b else a

## First non-missing value across a list of candidate columns, coalesced
## row-wise. The join suffixes clinical/biofluid name clashes, so each candidate
## is also tried with _clin and _bio appended.
coalesce_cols <- function(df, candidates, numeric = FALSE) {
  nms <- unlist(lapply(candidates, function(x)
    c(x, paste0(x, "_clin"), paste0(x, "_bio"))))
  nms <- nms[nms %in% names(df)]
  if (!length(nms))
    return(if (numeric) rep(NA_real_, nrow(df)) else rep(NA_character_, nrow(df)))
  out <- if (numeric) rep(NA_real_, nrow(df)) else rep(NA_character_, nrow(df))
  for (nm in nms) {
    v <- if (numeric) num_safe(df[[nm]]) else as.character(df[[nm]])
    if (!numeric) v[!is.na(v) & trimws(v) == ""] <- NA
    out <- dplyr::coalesce(out, v)
  }
  out
}
```

```{r clinical-reshape}
## ---------------------------------------------------------------------------
## The REDCap export is LONG: subject-level fields sit on the enrolment event,
## visit-level fields on repeating instances. Collapsing without carrying the
## subject-level fields down loses either the core symptoms or the scores.
## ---------------------------------------------------------------------------
nhc_col <- grep("^nhc$", names(clin_raw), ignore.case = TRUE, value = TRUE)[1]
stopifnot(!is.na(nhc_col))

clin <- clin_raw %>%
  mutate(NHC = norm_id(.data[[nhc_col]])) %>%
  filter(!is.na(NHC))

## ---------------------------------------------------------------------------
## Per-visit date.
## visit_date is filled on only a minority of rows, so dating visits by it alone
## discards most assessments. Each instrument in this export carries its own
## date (fecha_moca, fechaupdrs, fecha_sppb ...), and those are used as
## fallbacks, with the baseline date last.
##
## NOT date fields, despite the names: `fecha`, `mes`, `any`, `dia_semana`,
## `lugar`, `localidad` are MoCA orientation ITEMS scored 0/1.
## ---------------------------------------------------------------------------
VISIT_DATE_FIELDS <- c("visit_date", "fecha_moca", "fechaupdrs", "fecha_updrs_ia",
                       "fecha_sppb", "fecha_eq5d", "fechahad", "fecha_zarit",
                       "fecha_seguimiento", "fecha_um", "visit_date_base_lewy")

date_coverage <- tibble(
  field = VISIT_DATE_FIELDS,
  present = VISIT_DATE_FIELDS %in% names(clin),
  n_dated = vapply(VISIT_DATE_FIELDS, function(f)
    if (f %in% names(clin)) sum(!is.na(as_date_safe(clin[[f]]))) else 0L, integer(1)))

clin$visit_any <- {
  out <- as.Date(rep(NA, nrow(clin)))
  for (f in VISIT_DATE_FIELDS)
    if (f %in% names(clin)) out <- coalesce(out, as_date_safe(clin[[f]]))
  out
}
clin$visit_d <- as_date_safe(if (has_col(clin, "visit_date")) clin$visit_date else NA)

## Columns that belong to a VISIT (never carried across rows)
visit_pat <- paste0("^(visit_date|redcap_event_instance|neurologist|visitdiagnosis|",
  "gds_visit|medication|hoehn_yahr|diagnostico_visita|updrs|sppb|mayo|zarit|",
  "lethargic|sleep_day|disorganized|staring|fecha|puntuaciones|resultado|mis$|",
  "visuoespacial|identificacion|atencion|lenguaje|abstraccion|orientacion|",
  "at_|len_|directos|inversos|letras|serie7|repeticion|fluidez|num_palab|",
  "rostro|seda|templo|clavel|rojo|speech|facial|rigidity|tapping|handmovements|",
  "pronation|legagility|arising|gait|freezing|postural|globalspont|kinetictremor|",
  "resttremor|constancy|dyskinesia|posture|movilidad|autocuidado|actividades|",
  "dolor_|ansiedad|depresion|termometro|a_[0-9]|d_[0-9]|resumen_|seguimiento)")
visit_cols   <- grep(visit_pat, names(clin), value = TRUE)
subject_cols <- setdiff(names(clin), c(visit_cols, "NHC", "visit_d",
                                       "visit_fallback", "visit_any"))

clin <- clin %>%
  group_by(NHC) %>%
  mutate(across(all_of(subject_cols),
                ~ if (all(is.na(.x))) .x else coalesce(.x, .x[which(!is.na(.x))[1]]))) %>%
  ungroup()

clin_visits <- clin %>% filter(!is.na(visit_any)) %>%
  distinct(NHC, visit_any, .keep_all = TRUE)
clin_nodate <- clin %>% filter(!NHC %in% clin_visits$NHC) %>%
  distinct(NHC, .keep_all = TRUE)
clin_use <- bind_rows(clin_visits, clin_nodate)
```

```{r biofluid-merge}
bio <- bio_raw %>%
  mutate(NHC  = norm_id(.data[["NHC"]]),
         NHCb = if (has_col(bio_raw, "NHC2")) norm_id(.data[["NHC2"]]) else NA_character_) %>%
  mutate(NHC = coalesce(NHC, NHCb)) %>%
  filter(!is.na(NHC)) %>%
  distinct(across(everything()), .keep_all = TRUE)

bio$sample_d <- as_date_safe(bio$SAMPLEDATE)

join_args <- list(x = clin_use, y = bio, by = "NHC", suffix = c("_clin", "_bio"))
if (utils::packageVersion("dplyr") >= "1.1.0")
  join_args$relationship <- "many-to-many"

merged_df <- suppressWarnings(do.call(inner_join, join_args)) %>%
  mutate(date_diff = abs(as.numeric(sample_d - visit_any)),
         .rank = ifelse(is.na(date_diff), Inf, date_diff)) %>%
  group_by(NHC) %>%
  arrange(.rank, visit_any, sample_d, .by_group = TRUE) %>%
  slice(1) %>% ungroup() %>%
  mutate(gap_over_limit = !is.na(date_diff) & date_diff > MAX_GAP_DAYS)

merge_note <- sprintf(
  paste("Clinical: %s subjects / %s dated visits. Biofluids: %s subjects / %s samples.",
        "Merged: %s subjects. Visit-sample gap: median %s d [IQR %s-%s], %s%% within 90 d,",
        "%s pair(s) beyond %s d."),
  n_distinct(clin$NHC), nrow(clin_visits), n_distinct(bio$NHC), nrow(bio),
  nrow(merged_df),
  round(median(merged_df$date_diff, na.rm = TRUE)),
  round(quantile(merged_df$date_diff, .25, na.rm = TRUE)),
  round(quantile(merged_df$date_diff, .75, na.rm = TRUE)),
  round(100 * mean(merged_df$date_diff <= 90, na.rm = TRUE)),
  sum(merged_df$gap_over_limit, na.rm = TRUE), MAX_GAP_DAYS)
cat(merge_note)
```

```{r date-coverage}
kable(date_coverage %>%
        mutate(`% of rows` = round(100 * n_dated / nrow(clin), 1)),
      caption = paste("Which date fields are actually populated in the clinical",
        "export. visit_date alone would date only a fraction of the rows;",
        "visit_any coalesces down this list in order.")) %>%
  kable_styling(full_width = FALSE, bootstrap_options = c("striped","condensed"))
cat(sprintf("Rows with any usable date: %d of %d (%.0f%%).",
            sum(!is.na(clin$visit_any)), nrow(clin),
            100 * mean(!is.na(clin$visit_any))))
```

```{r gap-plot, fig.width=6.5, fig.height=3}
ggplot(merged_df %>% filter(!is.na(date_diff)), aes(date_diff)) +
  geom_histogram(bins = 40, fill = "#4C72B0", colour = "white") +
  geom_vline(xintercept = c(90, MAX_GAP_DAYS), linetype = c("dashed", "dotted"),
             colour = c("#D55E00", "grey40")) +
  labs(title = "Interval between clinical visit and biofluid sampling",
       x = "|visit date − sample date| (days)", y = "Subjects",
       caption = "Dashed = 90 d; dotted = analysis limit")
```

# Data quality

```{r derive}
## ---- safe recoders --------------------------------------------------------
## ANCHORED patterns. The original used str_detect(x, "pos|1"), which matches
## any string containing a "1" (e.g. "10", "0.15", a plate code).
recode_saa <- function(x) {
  v <- tolower(trimws(as.character(x)))
  out <- rep(NA_character_, length(v))
  out[grepl("^(0|neg|negative|negativo|negativa)$", v)] <- "Negative"
  out[grepl("^(1|pos|positive|positivo|positiva)$", v)] <- "Positive"
  out[grepl("^(2|3|9|ind|indet|indeterminate|inconclusive|dudoso|equivocal)$", v)] <-
    "Indeterminate"
  factor(out, levels = c("Negative", "Positive", "Indeterminate"))
}

recode_yesno <- function(x) {
  v <- tolower(trimws(as.character(x)))
  v[v %in% c("", "na", "null", "-", "9", "99", "unknown", "desconocido", "nd")] <- NA
  out <- rep(NA_character_, length(v))
  out[grepl("^(0|no|n|absent|ausente|negative|negativo|normal|false)$", v)] <- "No"
  out[grepl("^(1|2|yes|y|si|s[ií]|present|presente|positive|positivo|abnormal|anormal|patologico|true)$", v)] <- "Yes"
  factor(out, levels = c("No", "Yes"))
}

## Free-text/multi-level fields that must be collapsed rather than shown raw.
## The previous run printed DaT-SCAN and PSG with their full level sets
## (including "Scheduled", "NA", "RBD+SAHS+PLM"), producing an uninterpretable
## chi-squared across administrative categories.
recode_datscan <- function(x) {
  v <- tolower(trimws(as.character(x)))
  out <- rep(NA_character_, length(v))
  out[grepl("normal", v) & !grepl("abnormal|anormal|patolog", v)] <- "Normal"
  out[grepl("abnormal|anormal|patolog|altered|alterado", v)] <- "Abnormal"
  factor(out, levels = c("Normal", "Abnormal"))
}
recode_psg <- function(x) {                 # RBD present on PSG, yes/no
  v <- tolower(trimws(as.character(x)))
  out <- rep(NA_character_, length(v))
  out[grepl("no rbd|sin rbd", v)] <- "No"
  out[grepl("rbd", v) & !grepl("no rbd|sin rbd", v)] <- "Yes"
  out[grepl("scheduled|pending|programad|^na$|^$", v)] <- NA
  factor(out, levels = c("No", "Yes"))
}

## A/T status. The previous anchored version silently matched nothing on this
## export -- A_clean and AT_profile then vanished from every table because
## keep_usable() drops all-missing variables without comment. Substring matching
## is used instead, and the raw-to-recoded mapping is printed below so the
## result can be checked rather than assumed.
recode_status <- function(x, letter) {
  v <- gsub("\\s+", "", tolower(trimws(as.character(x))))
  v[v %in% c("", "na", "nan", "null", "nd", "unknown", "desconocido", "?")] <- NA
  out <- rep(NA_character_, length(v))
  lo <- tolower(letter)
  neg <- !is.na(v) & (grepl("neg", v) | v %in% c("0", "false", "-", paste0(lo, "-")))
  pos <- !is.na(v) & (grepl("pos", v) | v %in% c("1", "true", "+", paste0(lo, "+")))
  out[neg] <- paste0(letter, "-")
  out[pos] <- paste0(letter, "+")      # positive wins if both somehow match
  factor(out, levels = paste0(letter, c("-", "+")))
}

clamp <- function(x, lo, hi, label) {
  v <- num_safe(x)
  v[v %in% SENTINELS] <- NA
  bad <- !is.na(v) & (v < lo | v > hi)
  if (any(bad)) qc_log[[label]] <<- sum(bad)
  v[bad] <- NA
  v
}

qc_log <- list()

analytic_df <- merged_df
analytic_df$saa3_raw <- if (has_col(analytic_df, "CSF_ALPHASYN_SAA03"))
  recode_saa(analytic_df$CSF_ALPHASYN_SAA03) else factor(rep(NA, nrow(analytic_df)),
                                                         levels = c("Negative","Positive"))

## apply plausibility limits
for (i in seq_len(nrow(PLAUSIBLE))) {
  vn <- PLAUSIBLE$var[i]
  if (has_col(analytic_df, vn))
    analytic_df[[vn]] <- clamp(analytic_df[[vn]], PLAUSIBLE$lo[i],
                               PLAUSIBLE$hi[i], vn)
}

analytic_df <- analytic_df %>%
  mutate(
    SEX = factor(ifelse(grepl("^(f|female|mujer|femenino)$", tolower(trimws(SEX))),
                        "Female",
                 ifelse(grepl("^(m|male|hombre|var|masculino)", tolower(trimws(SEX))),
                        "Male", NA)), levels = c("Female", "Male")),

    saa_local_pos  = recode_saa(CSF_ALPHASYN_SAA02),
    saa_munich_pos = recode_saa(CSF_ALPHASYN_SAA01),
    saa_third      = saa3_raw,

    ## Own-site result preferred, Munich as fallback. The original rule coded
    ## ANY positive as Positive, which silently resolved every discordant pair
    ## in favour of Positive. Kept for reference; nothing is grouped by it.
    saa_any_pos = factor(coalesce(as.character(saa_local_pos),
                                  as.character(saa_munich_pos)),
                         levels = c("Negative", "Positive", "Indeterminate")),

    A_clean = recode_status(ASTATUS, "A"),
    T_clean = recode_status(TSTATUS, "T"),
    AT_profile = factor(if_else(is.na(A_clean) | is.na(T_clean), NA_character_,
                                paste0(A_clean, " / ", T_clean)),
                        levels = c("A- / T-", "A- / T+", "A+ / T-", "A+ / T+")),

    ## APOE: genotype string -> carrier. Guarded against a dosage-coded column.
    apoe4_carrier = {
      g <- gsub("[^0-9]", "", as.character(APOE))
      factor(ifelse(is.na(g) | nchar(g) < 2, NA,
                    ifelse(grepl("4", g), "Yes", "No")), levels = c("No", "Yes"))
    },

    hallucinations = recode_yesno(hallucinations),
    fluctuations   = recode_yesno(fluctuations),
    rbdclinical    = recode_yesno(rbdclinical),
    parkinsonism   = recode_yesno(parkinsonism),
    hyposmia       = recode_yesno(hyposmia),
    dysautonomia   = recode_yesno(dysautonomia),
    repeatedfalls  = recode_yesno(repeatedfalls),
    hypersomnia    = recode_yesno(hypersomnia),
    neurolepticsens= recode_yesno(neurolepticsens),
    datscan_bin    = recode_datscan(datscan),
    psg_rbd        = recode_psg(psg),

    ## age at sampling beats a stored age from an unrelated timepoint
    age_at_sample = {
      a <- as.numeric(sample_d - as_date_safe(fnac)) / 365.25
      a[!is.na(a) & (a < 18 | a > 110)] <- NA
      coalesce(a, num_safe(edad), num_safe(AGE))
    },

    n_core = rowSums(cbind(hallucinations == "Yes", fluctuations == "Yes",
                           rbdclinical == "Yes", parkinsonism == "Yes"), na.rm = TRUE),
    n_core = ifelse(rowSums(!is.na(cbind(hallucinations, fluctuations,
                                         rbdclinical, parkinsonism))) == 0, NA, n_core),

    ## UPDRS total: NA unless all four parts are present (partial sums are not
    ## comparable across subjects)
    total_updrs = ifelse(
      is.na(updrs1a_total_score) | is.na(updrs1b_total_score) |
      is.na(updrs2_total_score)  | is.na(updrs3_total_score), NA,
      updrs1a_total_score + updrs1b_total_score +
      updrs2_total_score + updrs3_total_score)
  )
```

```{r dlb-and-combinations}
## ---------------------------------------------------------------------------
## Added on Inigo's feedback: DLB diagnostic certainty, GDS, and combined
## criteria counts.
## ---------------------------------------------------------------------------

## Diagnostic certainty. Raw levels are printed below -- check them before
## trusting the recode, the wording varies between fields.
## ---------------------------------------------------------------------------
## Diagnosis. Two problems in the previous version, both silent:
##  1. coalesce took the first NON-MISSING value, and `diagnosis` holds
##     "Diagnostico CONFIRMADO"/"PENDIENTE" for most rows -- non-missing but
##     uninformative, so they blocked the fallback to dxlewy_last and left the
##     variable almost entirely NA.
##  2. "MCI-LBprob" matched "prob" and became Probable DLB, and "MCI-AD" fell
##     through to the prodromal branch although it is an Alzheimer diagnosis.
## Each field is therefore recoded FIRST and the recoded values are coalesced,
## so an uninformative entry yields NA and the next field is tried.
## ---------------------------------------------------------------------------
DX_FIELDS <- c("dxlewy_last", "diagnosis", "dx_estudio_last_visit",
               "diagnostico_visita", "visitdiagnosis", "dxbaselinelewy")

recode_dx <- function(x) {
  v <- tolower(trimws(as.character(x)))
  v <- gsub("[^a-z]", "", v)              # "DLB prob" and "DLBprob" alike
  out <- rep(NA_character_, length(v))
  is_lb  <- grepl("dlb|lewy|lb", v)
  is_mci <- grepl("mci|dcl|prodrom", v)
  ## MCI-AD is not a Lewy diagnosis, and must be caught before the mci branch
  out[grepl("mciad|dclad", v)]                        <- "MCI-AD"
  out[is.na(out) & is_mci & is_lb & grepl("prob", v)] <- "MCI-LB probable"
  out[is.na(out) & is_mci & is_lb]                    <- "MCI-LB possible"
  out[is.na(out) & is_lb & grepl("prob", v)]          <- "Probable DLB"
  out[is.na(out) & is_lb & grepl("pos", v)]           <- "Possible DLB"
  out[is.na(out) & grepl("irbd|isolatedrbd", v)]      <- "iRBD"
  out[is.na(out) & grepl("^ad$|alzheimer", v)]        <- "AD"
  ## "Diagnostico CONFIRMADO/PENDIENTE", "Otros", "Cognitively normal" -> NA
  factor(out, levels = c("iRBD", "MCI-LB possible", "MCI-LB probable",
                         "Possible DLB", "Probable DLB", "MCI-AD", "AD"))
}

dx_recoded <- lapply(intersect(DX_FIELDS, names(analytic_df)),
                     function(f) as.character(recode_dx(analytic_df[[f]])))
analytic_df$dlb_certainty <- factor(Reduce(dplyr::coalesce, dx_recoded),
  levels = c("iRBD", "MCI-LB possible", "MCI-LB probable",
             "Possible DLB", "Probable DLB", "MCI-AD", "AD"))
dx_raw <- coalesce_cols(analytic_df, DX_FIELDS)
analytic_df$dx_raw <- dx_raw

## Cognitive stage. Within the analysis cohort this is exactly the diagnosis
## (MCI-LB probable = prodromal, probable DLB = dementia); it is derived from
## the diagnosis rather than from GDS-FAST because GDS is missing for many
## patients and its cut-point between stages is not applied consistently.
analytic_df$cog_stage <- factor(dplyr::case_when(
  analytic_df$dlb_certainty %in% c("MCI-LB possible", "MCI-LB probable",
                                   "MCI-AD")            ~ "MCI",
  analytic_df$dlb_certainty %in% c("Possible DLB", "Probable DLB",
                                   "AD")                ~ "Dementia",
  TRUE                                                  ~ NA_character_),
  levels = c("MCI", "Dementia"))

## GDS-FAST
analytic_df$gds_fast <- {
  g <- coalesce_cols(analytic_df, c("gds_visit", "GDS", "gds_fast_last_visit"),
                     numeric = TRUE)
  g[!is.na(g) & (g < 1 | g > 7)] <- NA
  factor(g, levels = 1:7, labels = paste("GDS", 1:7))   # ordinal stage, not a score
}

## MIBG, needed for the biomarker count
analytic_df$mibg <- recode_yesno(coalesce_cols(analytic_df, c("mibg")))

## A/T in three groups: discordant profiles pooled as intermediate.
analytic_df$at_group <- factor(dplyr::case_when(
  is.na(analytic_df$A_clean) | is.na(analytic_df$T_clean) ~ NA_character_,
  analytic_df$A_clean == "A-" & analytic_df$T_clean == "T-" ~ "AT-negative",
  analytic_df$A_clean == "A+" & analytic_df$T_clean == "T+" ~ "AT-positive",
  TRUE ~ "AT-intermediate"),
  levels = c("AT-negative", "AT-intermediate", "AT-positive"))

## ---- combined categories --------------------------------------------------
yes <- function(x) !is.na(x) & x == "Yes"
abn <- function(x) !is.na(x) & x == "Abnormal"
seen <- function(...) rowSums(!is.na(cbind(...))) > 0

analytic_df <- analytic_df %>%
  mutate(
    ## RBD: clinical, PSG-confirmed, either, both
    rbd_any  = factor(ifelse(seen(rbdclinical, psg_rbd),
                 ifelse(yes(rbdclinical) | yes(psg_rbd), "Yes", "No"), NA),
                 levels = c("No", "Yes")),
    rbd_both = factor(ifelse(!is.na(rbdclinical) & !is.na(psg_rbd),
                 ifelse(yes(rbdclinical) & yes(psg_rbd), "Yes", "No"), NA),
                 levels = c("No", "Yes")),
    ## parkinsonism with dopaminergic imaging
    park_dat_any = factor(ifelse(seen(parkinsonism, datscan_bin),
                 ifelse(yes(parkinsonism) | abn(datscan_bin), "Yes", "No"), NA),
                 levels = c("No", "Yes")),
    park_dat_both = factor(ifelse(!is.na(parkinsonism) & !is.na(datscan_bin),
                 ifelse(yes(parkinsonism) & abn(datscan_bin), "Yes", "No"), NA),
                 levels = c("No", "Yes")),
    ## Indicative biomarkers present (DaT, PSG, MIBG), 0-3.
    ## COMPLETE CASES ONLY. Counting a test that was never done as "negative"
    ## makes the count partly a measure of how much workup a patient received,
    ## and workup intensity is not independent of the group.
    n_indicative = ifelse(!is.na(datscan_bin) & !is.na(psg_rbd) & !is.na(mibg),
      rowSums(cbind(abn(datscan_bin), yes(psg_rbd), yes(mibg))), NA_real_),
    ## same rule for the number of tests actually performed, which is worth
    ## checking on its own: if it differs by group, every count below inherits it
    n_tests_done = rowSums(cbind(!is.na(datscan_bin), !is.na(psg_rbd), !is.na(mibg))),
    ## core clinical + indicative biomarkers, 0-7, complete cases only
    n_core_plus = ifelse(is.na(n_core) | is.na(n_indicative), NA_real_,
      n_core + n_indicative))

cat(sprintf(paste("Diagnostic certainty recoded for %d of %d subjects.",
                  "Cognitive stage in %d (MCI %d, dementia %d). MIBG in %d.\n"),
  sum(!is.na(analytic_df$dlb_certainty)), nrow(analytic_df),
  sum(!is.na(analytic_df$cog_stage)),
  sum(analytic_df$cog_stage == "MCI", na.rm = TRUE),
  sum(analytic_df$cog_stage == "Dementia", na.rm = TRUE),
  sum(!is.na(analytic_df$mibg))))

dx_check <- tibble(Raw = dx_raw,
                   Recoded = as.character(analytic_df$dlb_certainty)) %>%
  filter(!is.na(Raw)) %>% count(Raw, Recoded, sort = TRUE) %>% head(25)
kable(dx_check, caption = paste("Diagnosis field: raw value to recoded certainty.",
  "Rows with Recoded = NA are not counted as possible or probable DLB.")) %>%
  kable_styling(full_width = FALSE, bootstrap_options = c("striped","condensed"))
```

```{r pd-phenotype}
## ---------------------------------------------------------------------------
## Motor phenotype (Jankovic): mean tremor / mean akinetic-rigid, from the
## MDS-UPDRS III items. Absent from the original script.
## ---------------------------------------------------------------------------
TREM <- c("posturaltremorr","posturaltremorl","kinetictremorr","kinetictremorl",
          "resttremorrue","resttremorlue","resttremorrle","resttremorlle",
          "resttremorlipjaw","constancy")
AKRI <- c("speech","facialexpression","rigidityneck","rigidityrue","rigiditylue",
          "rigidityrle","rigiditylle","tappingrue","tappinglue","handmovementsr",
          "handmovementsl","pronationsupinationr","pronationsupinationl",
          "tappingrle","tappinglle","legagilityr","legagilityl","arising","gait",
          "freezing","posturalstability","posture","globalspontaneityl")

item_matrix <- function(df, items) {
  got <- intersect(items, names(df))
  if (!length(got)) return(list(m = NULL, used = character()))
  m <- sapply(got, function(cc) {
    v <- num_safe(df[[cc]]); v[!is.na(v) & (v < 0 | v > 4)] <- NA; v
  })
  list(m = as.matrix(m), used = got)
}
tr <- item_matrix(analytic_df, TREM); ak <- item_matrix(analytic_df, AKRI)

mean_min <- function(m, k) {
  if (is.null(m)) return(rep(NA_real_, nrow(analytic_df)))
  out <- rowMeans(m, na.rm = TRUE); out[rowSums(!is.na(m)) < k] <- NA_real_; out
}
analytic_df$tremor_score  <- mean_min(tr$m, max(3, length(tr$used) %/% 2))
analytic_df$akrigid_score <- mean_min(ak$m, max(5, length(ak$used) %/% 2))
analytic_df <- analytic_df %>%
  mutate(td_ar_ratio = ifelse(!is.na(akrigid_score) & akrigid_score > 0,
                              tremor_score / akrigid_score, NA_real_),
         pd_phenotype = factor(case_when(
           is.na(td_ar_ratio)   ~ NA_character_,
           td_ar_ratio >= 1.0   ~ "Tremor-dominant",
           td_ar_ratio <= 0.8   ~ "Akinetic-rigid",
           TRUE                 ~ "Mixed"),
           levels = c("Akinetic-rigid", "Mixed", "Tremor-dominant")))

cat(sprintf("UPDRS-III items found: %d/%d tremor, %d/%d akinetic-rigid. Phenotype classifiable in %d subjects.",
            length(tr$used), length(TREM), length(ak$used), length(AKRI),
            sum(!is.na(analytic_df$pd_phenotype))))
```

```{r qc-report}
qc_tbl <- tibble(Variable = names(qc_log),
                 `Implausible values set to NA` = unlist(qc_log))
if (nrow(qc_tbl)) {
  kable(qc_tbl, caption = paste("Out-of-range values removed. Any non-zero row",
        "means the source column contains sentinel codes or the wrong field.")) %>%
    kable_styling(full_width = FALSE, bootstrap_options = c("striped","condensed"))
} else cat("No out-of-range values detected.")

saa_raw_check <- bind_rows(
  tibble(Assay = "SAA02 (Sant Pau)", Raw = as.character(merged_df$CSF_ALPHASYN_SAA02)),
  tibble(Assay = "SAA01 (Munich)",   Raw = as.character(merged_df$CSF_ALPHASYN_SAA01))) %>%
  filter(!is.na(Raw)) %>% count(Assay, Raw, sort = TRUE)
kable(saa_raw_check, caption = "Raw SAA values before recoding — verify every level maps as intended.") %>%
  kable_styling(full_width = FALSE, bootstrap_options = c("striped","condensed"))

at_raw_check <- bind_rows(
  if (has_col(merged_df, "ASTATUS"))
    tibble(Field = "ASTATUS", Raw = as.character(merged_df$ASTATUS),
           Recoded = as.character(recode_status(merged_df$ASTATUS, "A"))) else NULL,
  if (has_col(merged_df, "TSTATUS"))
    tibble(Field = "TSTATUS", Raw = as.character(merged_df$TSTATUS),
           Recoded = as.character(recode_status(merged_df$TSTATUS, "T"))) else NULL) %>%
  filter(!is.na(Raw)) %>% count(Field, Raw, Recoded, sort = TRUE)
if (nrow(at_raw_check))
  kable(at_raw_check, caption = paste("Amyloid / tau status: raw value to recoded",
    "value. Any row with Recoded = NA is being lost.")) %>%
    kable_styling(full_width = FALSE, bootstrap_options = c("striped","condensed"))
```

```{r cohorts}
## Indeterminate results are excluded from group comparisons, not silently merged.
n_indet <- sum(analytic_df$saa_any_pos == "Indeterminate", na.rm = TRUE)

df_saa2 <- analytic_df %>% filter(saa_local_pos %in% c("Negative","Positive")) %>%
  mutate(saa_local_pos = droplevels(saa_local_pos))

## Diagnosis composition of the SAA cohort, before any restriction.
dx_before <- df_saa2 %>% count(dlb_certainty, saa_local_pos) %>%
  pivot_wider(names_from = saa_local_pos, values_from = n, values_fill = 0)
print(kable(dx_before, caption = "Diagnoses in the Sant Pau SAA cohort before restriction.") %>%
  kable_styling(full_width = FALSE, bootstrap_options = c("striped","condensed")))

## Patients the recoder could not classify: list their raw diagnosis so any
## probable DLB hidden behind unrecognised wording can be recovered rather than
## silently dropped by the restriction.
unclassified <- df_saa2 %>% filter(is.na(dlb_certainty)) %>%
  count(dx_raw, saa_local_pos, name = "n")
if (nrow(unclassified)) {
  print(kable(unclassified, caption = paste("Cohort patients with no recoded",
    "diagnosis. Check these before accepting the restriction: if any is",
    "probable DLB, add its wording to recode_dx().")) %>%
    kable_styling(full_width = FALSE, bootstrap_options = c("striped","condensed")))
} else cat("Every patient in the cohort has a recoded diagnosis.\n")

if (PROBABLE_ONLY) {
  n_pre <- nrow(df_saa2)
  keep_dx <- c("Probable DLB",
               if (INCLUDE_MCI_LB_PROBABLE) "MCI-LB probable")
  df_saa2 <- df_saa2 %>% filter(dlb_certainty %in% keep_dx) %>%
    mutate(saa_local_pos = droplevels(saa_local_pos),
           dlb_certainty = droplevels(dlb_certainty))
  cat(sprintf("Restricted to %s: %d of %d patients retained (%d SAA+, %d SAA-).\n",
              paste(keep_dx, collapse = " and "), nrow(df_saa2), n_pre,
              sum(df_saa2$saa_local_pos == "Positive"),
              sum(df_saa2$saa_local_pos == "Negative")))
}
## Sant Pau only. saa_any_pos is still derived above but nothing is grouped by
## it; the concordance section it served has been removed.

saveRDS(list(analytic_df = analytic_df, df_saa2 = df_saa2),
        file.path(results_path, "analytic_df.rds"))

cat(sprintf("Sant Pau (SAA02): n = %d (%d SAA+ / %d SAA-). Indeterminate excluded: %d.",
  nrow(df_saa2), sum(df_saa2$saa_local_pos == "Positive"),
  sum(df_saa2$saa_local_pos == "Negative"), n_indet))
```

# Clinical demographics

```{r table-helpers}
## ---- effect sizes ---------------------------------------------------------
smd_value <- function(x, g) {
  g <- droplevels(factor(g)); if (nlevels(g) != 2) return(NA_real_)
  i1 <- g == levels(g)[1]; i2 <- g == levels(g)[2]
  if (is.numeric(x)) {
    a <- x[i1]; a <- a[!is.na(a)]; b <- x[i2]; b <- b[!is.na(b)]
    if (length(a) < 2 || length(b) < 2) return(NA_real_)
    den <- sqrt((var(a) + var(b)) / 2)
    return(if (!is.finite(den) || den == 0) NA_real_ else (mean(a) - mean(b)) / den)
  }
  f <- droplevels(factor(x)); k <- nlevels(f); if (k < 2) return(NA_real_)
  p1 <- as.numeric(prop.table(table(f[i1])))[-k]
  p2 <- as.numeric(prop.table(table(f[i2])))[-k]
  if (anyNA(p1) || anyNA(p2)) return(NA_real_)
  if (k == 2) {
    den <- sqrt((p1*(1-p1) + p2*(1-p2)) / 2)
    return(if (!is.finite(den) || den == 0) NA_real_ else (p1 - p2) / den)
  }
  S <- ((diag(p1, k-1) - outer(p1,p1)) + (diag(p2, k-1) - outer(p2,p2))) / 2
  inv <- try(solve(S), silent = TRUE); if (inherits(inv,"try-error")) return(NA_real_)
  as.numeric(sqrt(max(0, t(p1-p2) %*% inv %*% (p1-p2))))
}

fn_smd <- function(data, variable, by, ...) {
  v <- smd_value(data[[variable]], data[[by]])
  if (is.na(v)) NA_character_ else sprintf("%.2f", v)
}

## Mean difference (continuous) or risk difference (binary), with 95% CI.
## Supplied as a custom stat rather than via add_difference(), which would
## collide with add_p() over the p.value column.
fn_md <- function(data, variable, by, ...) {
  x <- data[[variable]]; g <- droplevels(factor(data[[by]]))
  if (nlevels(g) != 2) return(NA_character_)
  i1 <- g == levels(g)[1]; i2 <- g == levels(g)[2]
  if (is.numeric(x)) {
    a <- x[i1]; b <- x[i2]
    if (sum(!is.na(a)) < 2 || sum(!is.na(b)) < 2) return(NA_character_)
    tt <- try(t.test(a, b), silent = TRUE)
    if (inherits(tt, "try-error")) return(NA_character_)
    return(sprintf("%.2f (%.2f, %.2f)", unname(tt$estimate[1] - tt$estimate[2]),
                   tt$conf.int[1], tt$conf.int[2]))
  }
  f <- droplevels(factor(x)); if (nlevels(f) != 2) return(NA_character_)
  hi <- levels(f)[2]
  x1 <- sum(f[i1] == hi, na.rm = TRUE); n1 <- sum(!is.na(f[i1]))
  x2 <- sum(f[i2] == hi, na.rm = TRUE); n2 <- sum(!is.na(f[i2]))
  if (n1 < 2 || n2 < 2) return(NA_character_)
  pt <- try(prop.test(c(x1, x2), c(n1, n2), correct = FALSE), silent = TRUE)
  if (inherits(pt, "try-error")) return(NA_character_)
  sprintf("%.1f%% (%.1f, %.1f)", 100*(x1/n1 - x2/n2),
          100*pt$conf.int[1], 100*pt$conf.int[2])
}

## A tbl_summary module: n, median [IQR] or n(%), non-parametric p, difference, SMD.
## Continuous tests are rank-based; the previous script used Welch t-tests on
## ordinal (MTA, Fazekas) and heavily skewed (GFAP, NfL) variables.
mod_cont <- function(data, vars, labels, group_var,
                     stat = "{median} [{p25}, {p75}]") {
  data %>% select(all_of(vars), all_of(group_var)) %>%
    tbl_summary(by = all_of(group_var), missing = "no",
                type = list(everything() ~ "continuous"),
                label = labels,
                statistic = list(everything() ~ stat),
                digits = list(everything() ~ 1)) %>%
    add_n() %>%
    add_p(test = everything() ~ "wilcox.test",
          pvalue_fun = ~ style_pvalue(.x, digits = 3)) %>%
    add_stat(fns = everything() ~ fn_md) %>%
    modify_header(add_stat_1 = "**Diff. (95% CI)**") %>%
    add_stat(fns = everything() ~ fn_smd) %>%
    modify_header(add_stat_2 = "**SMD**") %>%
    bold_p(t = 0.05)
}

mod_cat <- function(data, vars, labels, group_var) {
  data %>% select(all_of(vars), all_of(group_var)) %>%
    tbl_summary(by = all_of(group_var), missing = "no", label = labels) %>%
    add_n() %>%
    add_p(test = everything() ~ "fisher.test",
          pvalue_fun = ~ style_pvalue(.x, digits = 3)) %>%
    add_stat(fns = everything() ~ fn_md) %>%
    modify_header(add_stat_1 = "**Diff. (95% CI)**") %>%
    add_stat(fns = everything() ~ fn_smd) %>%
    modify_header(add_stat_2 = "**SMD**") %>%
    bold_p(t = 0.05)
}

keep_usable <- function(data, vars, min_n = 5) {
  vars[vapply(vars, function(v) {
    if (!has_col(data, v)) return(FALSE)
    x <- data[[v]]
    sum(!is.na(x)) >= min_n && length(unique(x[!is.na(x)])) > 1
  }, logical(1))]
}

format_gt_table <- function(tbl) {
  as_gt(tbl) %>%
    tab_style(style = list(cell_fill(color = "#E0E0E0"), cell_text(weight = "bold")),
              locations = cells_row_groups())
}
```

```{r clinical-table}
LAB <- c(
  SEX="Female sex", apoe4_carrier="APOE ε4 carrier", age_at_sample="Age at sampling (y)",
  EDUC="Education (y)", puntuaciones="MoCA total", visuoespacialp="  Visuospatial/executive",
  identificacionp="  Naming", atencionp="  Attention", lenguajep="  Language",
  abstraccionp="  Abstraction", orientacionp="  Orientation", mis="  Memory index",
  total_updrs="MDS-UPDRS total (IA+IB+II+III)", updrs1a_total_score="  Part IA",
  updrs1b_total_score="  Part IB", updrs2_total_score="  Part II",
  updrs3_total_score="  Part III", hoehn_yahr="Hoehn & Yahr",
  pd_phenotype="Motor phenotype", td_ar_ratio="Tremor / akinetic-rigid ratio",
  hallucinations="Visual hallucinations", fluctuations="Cognitive fluctuations",
  rbdclinical="Clinical RBD", parkinsonism="Parkinsonism", n_core="Number of core features",
  hyposmia="Hyposmia", dysautonomia="Dysautonomia", repeatedfalls="Repeated falls",
  hypersomnia="Hypersomnia", neurolepticsens="Neuroleptic sensitivity",
  mayo_total_score="Mayo Fluctuations Scale", sppb_total_score="SPPB total",
  datscan_bin="DaT-SCAN abnormal", psg_rbd="PSG compatible with RBD",
  A_clean="Amyloid status", AT_profile="A/T profile", at_group="A/T profile",
  mta="MTA score", fazekas="Fazekas score",
  dlb_certainty="Diagnosis", gds_fast="GDS-FAST stage", mibg="MIBG abnormal",
  cog_stage="Cognitive stage",
  rbd_any="RBD (clinical or PSG)", rbd_both="RBD (clinical and PSG)",
  park_dat_any="Parkinsonism or abnormal DaT", park_dat_both="Parkinsonism and abnormal DaT",
  n_indicative="Indicative biomarkers positive (0-3)",
  n_tests_done="Indicative tests performed (0-3)",
  n_core_plus="Core + biomarkers (0-7)")

lab_list <- function(v) setNames(as.list(unname(LAB[v])), v) %>%
  { imap(., function(l, n) as.formula(paste0("`", n, "` ~ '", l, "'"))) } %>% unname()

build_clinical <- function(data, group_var) {
  blocks <- list(
    list(t="cat",  v=c("SEX","apoe4_carrier"),                      h="Demographics"),
    list(t="cont", v=c("age_at_sample","EDUC"),                      h="Demographics"),
    list(t="cat",  v=c("dlb_certainty", "cog_stage"),                h="Diagnosis"),
    list(t="cont", v=c("puntuaciones","visuoespacialp","identificacionp","atencionp",
                       "lenguajep","abstraccionp","orientacionp","mis"),
                                                                     h="Cognition (MoCA)"),
    list(t="cont", v=c("total_updrs","updrs1a_total_score","updrs1b_total_score",
                       "updrs2_total_score","updrs3_total_score","hoehn_yahr"),
                                                                     h="Motor (MDS-UPDRS)"),
    list(t="cat",  v="pd_phenotype",                                 h="Motor phenotype"),
    list(t="cat",  v=c("hallucinations","fluctuations","rbdclinical","parkinsonism"),
                                                                     h="Core features"),
    list(t="cont", v="n_core",                                       h="Core features"),
    list(t="cat",  v=c("hyposmia","dysautonomia","repeatedfalls","hypersomnia",
                       "neurolepticsens"),                           h="Supportive features"),
    list(t="cont", v=c("mayo_total_score","sppb_total_score"),        h="Scales"),
    list(t="cat",  v=c("datscan_bin","psg_rbd","mibg","at_group"),   h="Biomarkers & imaging"),
    list(t="cat",  v=c("rbd_any","rbd_both","park_dat_any","park_dat_both"),
                                                                     h="Combined categories"),
    list(t="cont", v=c("n_indicative","n_tests_done","n_core_plus"),
                                                                     h="Combined counts"),
    list(t="cont", v=c("mta","fazekas"),                             h="Biomarkers & imaging"))

  built <- list(); heads <- character()
  for (b in blocks) {
    vv <- keep_usable(data, b$v)
    if (!length(vv)) next
    tb <- if (b$t == "cont")
            mod_cont(data, vv, lab_list(vv), group_var,
                     stat = if (identical(b$h, "Combined counts"))
                       "{mean} ({sd}) / {median} [{p25}, {p75}]"
                     else "{median} [{p25}, {p75}]")
          else mod_cat (data, vv, lab_list(vv), group_var)
    built <- c(built, list(tb)); heads <- c(heads, b$h)
  }
  tbl_stack(built, group_header = heads) %>%
    modify_header(label ~ "**Variable**") %>% bold_labels()
}
```

## Sant Pau cohort (SAA02)

```{r tbl-saa2}
gt_demo_saa2 <- build_clinical(df_saa2, "saa_local_pos") %>%
  format_gt_table() %>%
  tab_source_note("Continuous: median [IQR], Wilcoxon rank-sum. Categorical: n (%), Fisher exact. Difference = Negative − Positive (Welch). SMD = standardized mean difference. N = subjects with the variable.")
gtsave(gt_demo_saa2, file.path(results_path, "Clinical_SantPau.html"))
gt_demo_saa2
```

# Combining criteria

Following Inigo's suggestion, two constructions rather than an exhaustive
screen:

- **Graded counts** — how many core clinical criteria, and how many criteria
  plus biomarkers, a patient has. Tested at every threshold, so the question
  "does having at least N criteria identify SAA positivity" gets a direct
  answer.
- **A clinical feature paired with its objective test** — RBD with PSG,
  parkinsonism with DaT-SCAN, dysautonomia with MIBG. Kept as four categories
  (neither, clinical only, test only, both) rather than collapsed to an AND,
  because which of the two is positive is itself informative.

Counts require every component to have been assessed. Treating a test that was
never done as a negative result would make the count partly a measure of how
much workup a patient received.

```{r combo-helpers}
MIN_RULE_N  <- 15
MIN_PER_ARM <- 5

rule_metrics <- function(rule, saa, label) {
  ok <- !is.na(rule) & !is.na(saa)
  if (sum(ok) < MIN_RULE_N) return(NULL)
  r <- rule[ok]; g <- droplevels(saa[ok])
  if (nlevels(g) != 2) return(NULL)
  pos <- g == "Positive"
  tp <- sum(r & pos); fp <- sum(r & !pos); fn <- sum(!r & pos); tn <- sum(!r & !pos)
  if ((tp + fn) < MIN_PER_ARM || (tn + fp) < MIN_PER_ARM) return(NULL)
  sens <- tp / (tp + fn); spec <- tn / (tn + fp)
  or <- ((tp + .5) * (tn + .5)) / ((fp + .5) * (fn + .5))
  se <- sqrt(1/(tp+.5) + 1/(fp+.5) + 1/(fn+.5) + 1/(tn+.5))
  tibble(Rule = label, n = sum(ok), `n SAA+` = sum(pos),
         Sens = round(100 * sens), Spec = round(100 * spec),
         PPV = round(100 * tp / max(1, tp + fp)),
         NPV = round(100 * tn / max(1, tn + fn)),
         Youden = round(sens + spec - 1, 2),
         AUC = round((sens + spec) / 2, 3),
         OR = round(or, 2),
         OR_CI = sprintf("%.2f-%.2f", exp(log(or) - 1.96*se), exp(log(or) + 1.96*se)),
         p = tryCatch(fisher.test(matrix(c(tp, fn, fp, tn), 2))$p.value,
                      error = function(e) NA_real_))
}
```

## Graded counts, at every threshold

```{r count-thresholds}
COUNTS <- c(n_core = "Core clinical criteria (0-4)",
            n_indicative = "Indicative biomarkers (0-3)",
            n_core_plus = "Core + biomarkers (0-7)")

thr_tab <- map_dfr(names(COUNTS), function(v) {
  if (!has_col(df_saa2, v)) return(NULL)
  x <- num_safe(df_saa2[[v]])
  ks <- sort(unique(x[!is.na(x)]))
  ks <- ks[ks > min(ks)]                     # ">= the minimum" is everyone
  map_dfr(ks, function(k) {
    m <- rule_metrics(x >= k, df_saa2$saa_local_pos,
                      sprintf("%s >= %g", unname(COUNTS[v]), k))
    if (is.null(m)) return(NULL)
    bind_cols(tibble(Score = unname(COUNTS[v]), Threshold = k), m %>% select(-Rule))
  })
})

if (nrow(thr_tab)) {
  write.csv(thr_tab, file.path(results_path, "Count_thresholds.csv"), row.names = FALSE)
  thr_tab %>%
    transmute(Score, `Cut-off` = paste0(">= ", Threshold), n, `n SAA+`,
              Sens, Spec, PPV, NPV, Youden,
              OR = sprintf("%.2f (%s)", OR, OR_CI), p = signif(p, 3)) %>%
    kable(caption = paste("Every usable cut-off of each count. A cut-off is",
      "skipped when fewer than", MIN_RULE_N, "patients have the full count or an",
      "arm falls below", MIN_PER_ARM, "patients.")) %>%
    kable_styling(full_width = FALSE, bootstrap_options = c("striped","condensed"),
                  font_size = 11)
} else cat("No count had enough complete cases to test at any threshold.")
```

```{r count-roc, fig.width=7, fig.height=5.5}
if (requireNamespace("pROC", quietly = TRUE)) {
  roc_list <- list(); auc_rows <- list()
  for (v in names(COUNTS)) {
    if (!has_col(df_saa2, v)) next
    d <- df_saa2 %>% mutate(x = num_safe(.data[[v]])) %>%
      filter(!is.na(x), !is.na(saa_local_pos))
    if (nrow(d) < 20 || n_distinct(d$saa_local_pos) < 2 || n_distinct(d$x) < 3) next
    r <- pROC::roc(d$saa_local_pos, d$x, levels = c("Negative", "Positive"),
                   direction = "<", quiet = TRUE)
    ci <- as.numeric(pROC::ci.auc(r))
    roc_list[[v]] <- tibble(score = unname(COUNTS[v]),
                            spec = rev(r$specificities), sens = rev(r$sensitivities),
                            thr  = rev(r$thresholds))
    auc_rows[[v]] <- tibble(Score = unname(COUNTS[v]), n = nrow(d),
      `n SAA+` = sum(d$saa_local_pos == "Positive"),
      AUC = round(as.numeric(pROC::auc(r)), 3),
      `95% CI` = sprintf("%.2f-%.2f", ci[1], ci[3]))
  }
  if (length(auc_rows))
    print(kable(bind_rows(auc_rows), caption =
      "Discrimination of each count. AUC 0.5 means the count carries no information.") %>%
      kable_styling(full_width = FALSE, bootstrap_options = c("striped","condensed")))
  rc <- bind_rows(roc_list)
  if (nrow(rc)) {
    ## Label the operating points with their cut-off. ggrepel when available,
    ## otherwise alternate the offset so labels do not sit on top of each other.
    pts <- rc %>% filter(is.finite(thr)) %>% group_by(score) %>%
      arrange(thr, .by_group = TRUE) %>%
      mutate(nudge = ifelse(dplyr::row_number() %% 2 == 0, 5, -5)) %>% ungroup()
    g <- ggplot(rc, aes(100 * (1 - spec), 100 * sens, colour = score)) +
      geom_abline(slope = 1, intercept = 0, linetype = "dashed", colour = "grey65") +
      geom_line(linewidth = .9) +
      geom_point(data = pts, size = 1.8) +
      coord_equal(xlim = c(-5, 105), ylim = c(-5, 105)) +
      labs(x = "100 - specificity (%)", y = "Sensitivity (%)", colour = NULL,
           title = "Counts of criteria as predictors of SAA positivity",
           caption = "Points mark each cut-off; labels give the minimum number of criteria.") +
      theme(legend.position = "bottom")
    g <- if (requireNamespace("ggrepel", quietly = TRUE))
      g + ggrepel::geom_text_repel(data = pts, aes(label = paste0(">=", ceiling(thr))),
            size = 3, show.legend = FALSE, min.segment.length = 0,
            box.padding = .5, max.overlaps = Inf, seed = 1)
    else
      g + geom_text(data = pts, aes(label = paste0(">=", ceiling(thr))), size = 3,
            nudge_y = pts$nudge, show.legend = FALSE)
    print(g)
  }
} else cat("pROC is not installed; install.packages('pROC') for the AUCs and curves.")
```

## Clinical feature paired with its objective test

```{r domain-pairs}
## Four categories, not an AND: which of the pair is positive matters. NA unless
## both were assessed, so every patient in the table contributed both.
pair_factor <- function(cl, te, cl_lab, te_lab) {
  ok <- !is.na(cl) & !is.na(te)
  out <- rep(NA_character_, length(cl))
  out[ok] <- ifelse(cl[ok] & te[ok], "Both",
             ifelse(cl[ok] & !te[ok], paste(cl_lab, "only"),
             ifelse(!cl[ok] & te[ok], paste(te_lab, "only"), "Neither")))
  factor(out, levels = c("Neither", paste(cl_lab, "only"), paste(te_lab, "only"), "Both"))
}

df_saa2 <- df_saa2 %>% mutate(
  pair_rbd  = pair_factor(rbdclinical == "Yes",  psg_rbd == "Yes",     "Clinical", "PSG"),
  pair_park = pair_factor(parkinsonism == "Yes", datscan_bin == "Abnormal",
                          "Parkinsonism", "DaT"),
  pair_auto = pair_factor(dysautonomia == "Yes", mibg == "Yes",        "Dysautonomia", "MIBG"))

PAIRS <- c(pair_rbd = "RBD: clinical vs PSG",
           pair_park = "Parkinsonism vs DaT-SCAN",
           pair_auto = "Dysautonomia vs MIBG")

pair_cross <- map_dfr(names(PAIRS), function(v) {
  d <- df_saa2 %>% filter(!is.na(.data[[v]]), !is.na(saa_local_pos))
  if (!nrow(d)) return(NULL)
  tb <- table(droplevels(d[[v]]), droplevels(d$saa_local_pos))
  ## both SAA groups must be present or the percentage column is undefined
  if (nrow(tb) < 2 || ncol(tb) < 2) return(NULL)
  p <- tryCatch(fisher.test(tb, simulate.p.value = TRUE)$p.value,
                error = function(e) NA_real_)
  as.data.frame.matrix(tb) %>% tibble::rownames_to_column("Category") %>%
    mutate(Pair = unname(PAIRS[v]), n = sum(tb), p = signif(p, 3),
           `% SAA+ in category` = round(100 * Positive / (Positive + Negative)))
})

if (nrow(pair_cross)) {
  write.csv(pair_cross, file.path(results_path, "Domain_pairs.csv"), row.names = FALSE)
  pair_cross %>% select(Pair, Category, Negative, Positive,
                        `% SAA+ in category`, n, p) %>%
    kable(caption = paste("Each clinical feature against its objective test.",
      "Only patients with both assessed. p is a Fisher test across the four",
      "categories within each pair.")) %>%
    kable_styling(full_width = FALSE, bootstrap_options = c("striped","condensed"))
} else cat("No pair had both components assessed in enough patients.")
```

```{r domain-binaries}
## The same pairs collapsed to a single rule: both components positive.
## `&` does not propagate NA the way this rule needs: FALSE & NA is FALSE, so a
## patient who is clinically RBD-negative and never had a PSG came out as
## rule-negative rather than unevaluable. That put 23 patients without a PSG
## into the denominator as negatives and made the rule look stronger than it is
## (n = 42, p = 0.038 instead of n = 19, p = 0.17). Both components must have
## been assessed, the same complete-case rule used for the counts.
and_assessed <- function(a, b) ifelse(is.na(a) | is.na(b), NA, a & b)

pair_rules <- bind_rows(
  rule_metrics(and_assessed(df_saa2$rbdclinical == "Yes", df_saa2$psg_rbd == "Yes"),
               df_saa2$saa_local_pos, "RBD clinical AND PSG compatible"),
  rule_metrics(and_assessed(df_saa2$parkinsonism == "Yes",
                            df_saa2$datscan_bin == "Abnormal"),
               df_saa2$saa_local_pos, "Parkinsonism AND DaT"),
  rule_metrics(and_assessed(df_saa2$dysautonomia == "Yes", df_saa2$mibg == "Yes"),
               df_saa2$saa_local_pos, "Dysautonomia AND MIBG"))

if (nrow(pair_rules)) {
  pair_rules %>%
    transmute(Rule, n, `n SAA+`, Sens, Spec, PPV, NPV, Youden, AUC,
              OR = sprintf("%.2f (%s)", OR, OR_CI), p = signif(p, 3)) %>%
    kable(caption = paste("Both components positive. n is the number of patients",
      "who had both assessed, which is why it is much smaller than the cohort.")) %>%
    kable_styling(full_width = FALSE, bootstrap_options = c("striped","condensed"))
} else cat("No pair reached the minimum evaluable sample.")
```

# Tables for the abstract

AD/PD accepts JPG images up to 500 KB and 600 by 800 pixels, and images do not
count towards the 280-word limit. Two are written to the Results folder.

At 600 pixels wide a table has to stay narrow, so these carry only the columns
that earn their place. Check them at full size before submitting: if a row is
unreadable, drop it rather than shrinking the font further.

```{r abstract-tables}
## AD/PD caps uploads at 600 x 800 px, which is small for a 20-row table. Two
## versions are written from the same layout: the capped JPG for submission, and
## a high-resolution PNG for reading and for emailing to co-authors. PNG because
## JPEG smears text edges on flat backgrounds.
IMG_W <- 600; IMG_H <- 800; IMG_DPI <- 100
HIRES_SCALE <- 3

## Compact summary rows: median [IQR] for numeric, n (%) per level for factors.
sum_rows <- function(dat, v, label, group = "saa_local_pos") {
  if (!has_col(dat, v)) return(NULL)
  x <- dat[[v]]; g <- dat[[group]]
  ok <- !is.na(x) & !is.na(g); if (sum(ok) < 3) return(NULL)
  x <- x[ok]; g <- droplevels(g[ok])
  lv <- levels(g)
  if (is.numeric(x)) {
    f <- function(k) sprintf("%.1f [%.1f, %.1f]", median(x[g == k]),
                             quantile(x[g == k], .25), quantile(x[g == k], .75))
    p <- tryCatch(wilcox.test(x ~ g, exact = FALSE)$p.value, error = function(e) NA)
    tibble(Variable = label, Level = "", `SAA-` = f(lv[1]), `SAA+` = f(lv[2]),
           p = ifelse(is.na(p), "", format.pval(p, digits = 2, eps = .001)))
  } else {
    f <- droplevels(factor(x))
    p <- tryCatch(fisher.test(table(f, g), simulate.p.value = TRUE)$p.value,
                  error = function(e) NA)
    pc <- function(k, l) sprintf("%d (%.0f%%)", sum(f[g == k] == l),
                                 100 * mean(f[g == k] == l))
    if (nlevels(f) == 2) {
      ## A binary variable needs one row, not two: the second level implies the
      ## first. At 600 px wide the table has no room for redundant rows.
      l <- levels(f)[2]
      tibble(Variable = paste0(label, ", ", l), Level = "",
             `SAA-` = pc(lv[1], l), `SAA+` = pc(lv[2], l),
             p = ifelse(is.na(p), "", format.pval(p, digits = 2, eps = .001)))
    } else {
      map_dfr(seq_along(levels(f)), function(i) {
        l <- levels(f)[i]
        tibble(Variable = if (i == 1) label else "", Level = l,
               `SAA-` = pc(lv[1], l), `SAA+` = pc(lv[2], l),
               p = if (i == 1 && !is.na(p)) format.pval(p, digits = 2, eps = .001) else "")
      })
    }
  }
}

T1_VARS <- list(
  c("age_at_sample", "Age, years"), c("SEX", "Sex"),
  c("EDUC", "Education, years"), c("apoe4_carrier", "APOE e4 carrier"),
  c("cog_stage", "Cognitive stage"), c("puntuaciones", "MoCA total"),
  c("updrs3_total_score", "MDS-UPDRS III"),
  c("fluctuations", "Fluctuations"), c("hallucinations", "Visual hallucinations"),
  c("rbdclinical", "RBD, clinical"), c("parkinsonism", "Parkinsonism"),
  c("datscan_bin", "DaT-SCAN"), c("psg_rbd", "PSG compatible with RBD"),
  c("mibg", "MIBG"), c("at_group", "A/T profile"))

tab1 <- map_dfr(T1_VARS, function(z) sum_rows(df_saa2, z[1], z[2]))

n_neg <- sum(df_saa2$saa_local_pos == "Negative")
n_pos <- sum(df_saa2$saa_local_pos == "Positive")
names(tab1)[3:4] <- c(sprintf("SAA- (n=%d)", n_neg), sprintf("SAA+ (n=%d)", n_pos))

## ---------------------------------------------------------------------------
## Table images.
## The previous version handed a tableGrob to annotation_custom(), which draws
## the grob at its natural size and centres it, leaving most of a 600 x 800
## canvas empty and the text unreadably small. This lays the table out on
## explicit coordinates instead, so the rows fill the height and the columns
## fill the width whatever the row count.
## ---------------------------------------------------------------------------
save_table_jpg <- function(df, file, title = NULL, align = NULL,
                           col_x = NULL, base_pt = NULL) {
  ## df[] preserves the column names; as.data.frame(lapply(...)) sanitises them
  ## and turns "SAA- (n=37)" into "SAA...n.37."
  df <- as.data.frame(df, stringsAsFactors = FALSE, check.names = FALSE)
  df[] <- lapply(df, as.character)
  nr <- nrow(df); nc <- ncol(df)
  if (!nr) return(invisible(NULL))
  if (is.null(align)) align <- c("l", rep("r", nc - 1))

  ## Column x positions from the widest cell in each column.
  if (is.null(col_x)) {
    w <- vapply(seq_len(nc), function(j)
      max(nchar(c(names(df)[j], df[[j]])), na.rm = TRUE), numeric(1))
    w <- w + 2                      # gutter
    col_x <- c(0, cumsum(w)[-nc]) / sum(w)
    col_x <- 0.02 + col_x * 0.96
    right <- col_x + (w / sum(w)) * 0.96 - 0.012
  } else right <- col_x
  xs <- ifelse(align == "l", col_x, right)

  ## Rows share the height below the title; the font follows from the row
  ## pitch, capped so a short table does not end up with giant text.
  top <- if (is.null(title)) 0.97 else 0.90
  pitch <- (top - 0.04) / (nr + 1)
  ## The font must satisfy BOTH constraints: the row pitch, and the total width
  ## of the widest cells. Sizing on height alone let long cells such as
  ## "78.8 [72.3, 81.6]" run into the next column.
  pt_h <- pitch * IMG_H / IMG_DPI * 72 * 0.55
  chars <- if (exists("w")) sum(w) else sum(nchar(names(df)) + 2)
  pt_w <- (0.96 * IMG_W / IMG_DPI) * 72 / (chars * 0.5)
  pt <- if (!is.null(base_pt)) base_pt else min(11, max(5, pt_h, na.rm = TRUE),
                                                pt_w, na.rm = TRUE)

  cells <- do.call(rbind, lapply(seq_len(nr), function(i)
    data.frame(x = xs, y = top - i * pitch, lab = unlist(df[i, ], use.names = FALSE),
               hj = ifelse(align == "l", 0, 1), stringsAsFactors = FALSE)))
  head <- data.frame(x = xs, y = top + pitch * 0.35, lab = names(df),
                     hj = ifelse(align == "l", 0, 1), stringsAsFactors = FALSE)

  p <- ggplot() +
    geom_text(data = head, aes(x, y, label = lab, hjust = hj),
              size = pt / 2.845, fontface = "bold", colour = "grey30") +
    geom_text(data = cells, aes(x, y, label = lab, hjust = hj),
              size = pt / 2.845, colour = "grey10") +
    geom_hline(yintercept = top - pitch * 0.45, linewidth = .5, colour = "grey20") +
    scale_x_continuous(limits = c(0, 1), expand = c(0, 0)) +
    scale_y_continuous(limits = c(0, 1), expand = c(0, 0)) +
    theme_void() +
    ## theme_void() leaves the background transparent; PNG keeps the alpha
    ## channel and viewers render it black, so paint it explicitly.
    theme(plot.margin = margin(4, 4, 4, 4),
          plot.background = element_rect(fill = "white", colour = NA),
          panel.background = element_rect(fill = "white", colour = NA))
  if (!is.null(title))
    p <- p + annotate("text", x = 0.02, y = 0.965, label = title, hjust = 0,
                      size = (pt + 1.5) / 2.845, fontface = "bold", colour = "grey10")

  ## submission copy, capped at 600 x 800 px
  ggsave(file.path(results_path, file), p, device = "jpeg",
         width = IMG_W / IMG_DPI, height = IMG_H / IMG_DPI, dpi = IMG_DPI,
         units = "in", quality = 100, bg = "white")
  kb <- round(file.size(file.path(results_path, file)) / 1024)

  ## readable copy: same physical layout, more pixels
  hires <- sub("\\.jpg$", "_hires.png", file)
  ggsave(file.path(results_path, hires), p, device = "png",
         width = IMG_W / IMG_DPI, height = IMG_H / IMG_DPI,
         dpi = IMG_DPI * HIRES_SCALE, units = "in", bg = "white")

  cat(sprintf("%s: %d rows, %.1f pt text, %d x %d px, %d KB %s\n  %s: %d x %d px\n",
              file, nr, pt, IMG_W, IMG_H, kb,
              if (kb > 500) "-- OVER the 500 KB limit" else "(within limits)",
              hires, IMG_W * HIRES_SCALE, IMG_H * HIRES_SCALE))
}

cat(sprintf("Table 1: %d rows. At %d px tall that is about %.0f px per row%s\n",
    nrow(tab1), IMG_H, IMG_H / (nrow(tab1) + 3),
    if (IMG_H / (nrow(tab1) + 3) < 20) " -- likely too cramped, drop rows." else "."))
save_table_jpg(tab1, "Abstract_Table1.jpg",
               "Table 1. Cohort by CSF alpha-synuclein SAA status")
kable(tab1, caption = "Table 1 as exported.") %>%
  kable_styling(full_width = FALSE, bootstrap_options = c("striped","condensed"),
                font_size = 11)
```

```{r abstract-table2}
## Table 2 stacks the count thresholds on the paired clinical/objective rules.
## One Rule column rather than Measure plus Rule: nine columns do not fit in
## 600 px, and the measure is short enough to fold into the rule text.
short <- c("Core clinical criteria (0-4)" = "Core criteria",
           "Indicative biomarkers (0-3)" = "Biomarkers",
           "Core + biomarkers (0-7)" = "Core + biomarkers")
t2a <- if (exists("thr_tab") && nrow(thr_tab))
  thr_tab %>% transmute(
    Rule = paste0(ifelse(is.na(short[Score]), Score, short[Score]), " >= ", Threshold),
    n, Sens, Spec, PPV, NPV, OR = sprintf("%.1f", OR),
    p = format.pval(p, digits = 2, eps = .001)) else NULL
t2b <- if (exists("pair_rules") && nrow(pair_rules))
  pair_rules %>% transmute(Rule = gsub(" compatible", "", Rule),
                           n, Sens, Spec, PPV, NPV, OR = sprintf("%.1f", OR),
                           p = format.pval(p, digits = 2, eps = .001)) else NULL
tab2 <- bind_rows(t2a, t2b)

if (nrow(tab2)) {
  cat(sprintf("Table 2: %d rows, about %.0f px per row%s\n",
      nrow(tab2), IMG_H / (nrow(tab2) + 3),
      if (IMG_H / (nrow(tab2) + 3) < 20) " -- likely too cramped, drop rows." else "."))
  save_table_jpg(tab2, "Abstract_Table2.jpg",
                 "Table 2. Counts of criteria and paired clinical/objective rules")
  kable(tab2, caption = "Table 2 as exported.") %>%
    kable_styling(full_width = FALSE, bootstrap_options = c("striped","condensed"),
                  font_size = 11)
} else cat("Neither the threshold table nor the paired rules produced rows.")
```

## Numbers quoted in the abstract

Every figure the abstract text cites, printed together so it can be checked
against the tables above without hunting through the document.

```{r abstract-numbers}
## ---------------------------------------------------------------------------
## Nothing new is computed here; this is a transcription aid. If a number in
## the abstract does not appear below, it does not come from this analysis.
## ---------------------------------------------------------------------------
d <- df_saa2
pos <- d$saa_local_pos == "Positive"

pct <- function(k, n) sprintf("%d/%d (%.0f%%)", k, n, 100 * k / n)
med <- function(x) { x <- num_safe(x); x <- x[!is.na(x)]
  sprintf("%.1f [%.1f, %.1f]", median(x), quantile(x, .25), quantile(x, .75)) }

cat("COHORT\n")
cat(sprintf("  n = %d (%d SAA+, %d SAA-)\n", nrow(d), sum(pos), sum(!pos)))
if (has_col(d, "SEX")) {
  f <- d$SEX == "Female"
  cat(sprintf("  Women: %s\n", pct(sum(f, na.rm = TRUE), sum(!is.na(f)))))
}
cat(sprintf("  Age at sampling, whole cohort, median [IQR]: %s\n",
            med(d$age_at_sample)))

cat("\nSTAGE\n")
st <- table(d$cog_stage, d$saa_local_pos)
print(st)
if (nrow(st) == 2 && ncol(st) == 2)
  cat(sprintf("  Positivity: MCI %s vs dementia %s, Fisher p = %.3f\n",
      pct(st["MCI","Positive"], sum(st["MCI",])),
      pct(st["Dementia","Positive"], sum(st["Dementia",])),
      fisher.test(st)$p.value))

cat("\nFEATURES CITED\n")
for (v in c("rbdclinical", "datscan_bin", "psg_rbd", "hyposmia")) {
  if (!has_col(d, v)) next
  lv <- if (v == "datscan_bin") "Abnormal" else "Yes"
  ok <- !is.na(d[[v]]) & !is.na(d$saa_local_pos)
  tb <- table(droplevels(factor(d[[v]][ok])), droplevels(d$saa_local_pos[ok]))
  ## a feature present in every patient has one row and no test to run
  pv <- if (nrow(tb) > 1 && ncol(tb) > 1) fisher.test(tb)$p.value else NA_real_
  cat(sprintf("  %-13s SAA+ %s | SAA- %s | Fisher p = %s\n", v,
      pct(sum(d[[v]][ok & pos] == lv), sum(ok & pos)),
      pct(sum(d[[v]][ok & !pos] == lv), sum(ok & !pos)),
      ifelse(is.na(pv), "not testable", sprintf("%.3f", pv))))
}
cat("\nOdds ratios and CIs are in the rule tables above; the core-criteria AUC",
    "is in the count-discrimination table.\n")
```

# Cognitive progression (MMSE and MoCA)

Longitudinal scores, so this uses **all** visits rather than the single row per
subject the rest of the report is built on. MoCA comes from the REDCap visits,
MMSE from the biofluids follow-ups.

```{r progression-build}
## Each instrument is dated by ITS OWN date field. Dating MoCA by visit_date
## discarded most assessments, which is why the first run found only 8 subjects
## with two MoCAs.
saa_lookup <- df_saa2 %>% select(NHC, saa_local_pos, SEX, age_at_sample)

score_history <- function(df, score_col, date_cols, label, lo, hi) {
  if (!has_col(df, score_col))
    return(tibble(NHC = character(), date = as.Date(character()),
                  score = numeric(), test = character()))
  d <- as.Date(rep(NA, nrow(df)))
  for (f in date_cols) if (f %in% names(df)) d <- coalesce(d, as_date_safe(df[[f]]))
  tibble(NHC = df$NHC, date = d, score = num_safe(df[[score_col]]), test = label) %>%
    filter(!is.na(NHC), !is.na(date), !is.na(score), score >= lo, score <= hi)
}

moca_long  <- score_history(clin, "puntuaciones",
                            c("fecha_moca", "visit_date", "visit_date_base_lewy"),
                            "MoCA", 0, 30)
updrs_long <- score_history(clin, "updrs3_total_score",
                            c("fechaupdrs", "visit_date"), "MDS-UPDRS III", 0, 132)
## MMSE dating: FUPDATE may be a subject-level "last follow-up" rather than a
## per-assessment date, which would collapse every MMSE of a patient onto one
## date. Pick the field that actually varies within subject, and show the
## comparison rather than assuming.
mmse_date_check <- map_dfr(c("FUPDATE", "DATA_CAPTURE", "SAMPLEDATE"), function(f) {
  if (!has_col(bio, f) || !has_col(bio, "MMSE")) return(NULL)
  d <- tibble(NHC = bio$NHC, date = as_date_safe(bio[[f]]),
              score = num_safe(bio$MMSE)) %>%
    filter(!is.na(NHC), !is.na(date), !is.na(score))
  if (!nrow(d)) return(NULL)
  per <- d %>% group_by(NHC) %>%
    summarise(dates = n_distinct(date), pairs = n_distinct(paste(date, score)),
              .groups = "drop")
  tibble(field = f, rows = nrow(d), subjects = nrow(per),
         `subjects with >=2 dates` = sum(per$dates >= 2),
         `subjects with >=2 date-score pairs` = sum(per$pairs >= 2))
})
if (nrow(mmse_date_check))
  print(kable(mmse_date_check, caption = paste("MMSE date fields compared.",
    "A field that gives few subjects with two distinct dates is a subject-level",
    "date, not a per-assessment one.")) %>%
    kable_styling(full_width = FALSE, bootstrap_options = c("striped","condensed")))

best_mmse_date <- if (nrow(mmse_date_check))
  mmse_date_check$field[which.max(mmse_date_check$`subjects with >=2 dates`)] else "SAMPLEDATE"
mmse_long  <- score_history(bio, "MMSE", best_mmse_date, "MMSE", 0, 30)

## What each source contributed, before any modelling.
src <- bind_rows(
  tibble(test = "MoCA",          rows = nrow(moca_long),  subjects = n_distinct(moca_long$NHC)),
  tibble(test = "MDS-UPDRS III", rows = nrow(updrs_long), subjects = n_distinct(updrs_long$NHC)),
  tibble(test = "MMSE",          rows = nrow(mmse_long),  subjects = n_distinct(mmse_long$NHC)))
kable(src, caption = "Dated assessments found in the whole export, before restricting to the SAA cohort.") %>%
  kable_styling(full_width = FALSE, bootstrap_options = c("striped","condensed"))

long <- bind_rows(moca_long, updrs_long, mmse_long) %>%
  distinct(NHC, test, date, .keep_all = TRUE) %>%
  inner_join(saa_lookup, by = "NHC") %>%
  group_by(NHC, test) %>%
  arrange(date, .by_group = TRUE) %>%
  mutate(t_years = as.numeric(date - min(date)) / 365.25,
         n_obs = dplyr::n(),
         span = max(t_years)) %>%
  ungroup()

avail <- long %>% group_by(test) %>%
  summarise(subjects = n_distinct(NHC),
            `with >=2 visits` = n_distinct(NHC[n_obs >= 2]),
            `SAA+ with >=2` = n_distinct(NHC[n_obs >= 2 & saa_local_pos == "Positive"]),
            `SAA- with >=2` = n_distinct(NHC[n_obs >= 2 & saa_local_pos == "Negative"]),
            `median visits` = median(n_obs),
            `median follow-up (y)` = round(median(span[n_obs >= 2]), 2),
            .groups = "drop")
kable(avail, caption = "Longitudinal data available in the Sant Pau SAA cohort.") %>%
  kable_styling(full_width = FALSE, bootstrap_options = c("striped","condensed"))
```

```{r progression-model}
## Two estimates, because neither alone is convincing at this size:
## a per-subject slope (simple, robust, uses only subjects with >=2 visits) and
## a mixed model (uses everyone, borrows strength, needs lme4).
prog_slopes <- long %>% filter(n_obs >= 2) %>%
  group_by(NHC, test, saa_local_pos) %>%
  summarise(slope = tryCatch(unname(coef(lm(score ~ t_years))[2]),
                             error = function(e) NA_real_),
            .groups = "drop") %>%
  filter(is.finite(slope))

if (nrow(prog_slopes)) {
  slope_tab <- prog_slopes %>% group_by(test, saa_local_pos) %>%
    summarise(n = dplyr::n(),
              `median slope (pts/y)` = round(median(slope), 2),
              IQR = sprintf("[%.2f, %.2f]", quantile(slope, .25), quantile(slope, .75)),
              .groups = "drop")
  slope_p <- prog_slopes %>% group_by(test) %>%
    summarise(p = tryCatch(signif(wilcox.test(slope ~ saa_local_pos,
                                              exact = FALSE)$p.value, 3),
                           error = function(e) NA_real_),
              SMD = round(smd_value(slope, saa_local_pos), 2), .groups = "drop")
  kable(left_join(slope_tab, slope_p, by = "test"),
        caption = paste("Per-subject annual change. Negative = decline.",
                        "Only subjects with at least two timepoints.")) %>%
    kable_styling(full_width = FALSE, bootstrap_options = c("striped","condensed"))
  write.csv(prog_slopes, file.path(results_path, "Progression_slopes.csv"),
            row.names = FALSE)
} else cat("No subject has two or more timepoints for either test.")
```

```{r progression-lmm}
if (!requireNamespace("lme4", quietly = TRUE)) {
  cat("lme4 is not installed, so the mixed model is skipped.",
      "install.packages('lme4') to enable it.")
} else {
  lmm_res <- map_dfr(unique(long$test), function(tt) {
    d <- long %>% filter(test == tt) %>% drop_na(score, t_years, saa_local_pos,
                                                 SEX, age_at_sample)
    if (n_distinct(d$NHC) < 10 || sum(d$n_obs >= 2) < 6) return(NULL)
    m <- try(lme4::lmer(score ~ t_years * saa_local_pos + age_at_sample + SEX +
                          (1 | NHC), data = d, REML = TRUE), silent = TRUE)
    if (inherits(m, "try-error")) return(NULL)
    sm <- summary(m)$coefficients
    rn <- grep("^t_years:saa_local_pos", rownames(sm), value = TRUE)[1]
    if (is.na(rn)) return(NULL)
    tibble(Test = tt, subjects = n_distinct(d$NHC), observations = nrow(d),
           `slope difference (pts/y)` = round(unname(sm[rn, "Estimate"]), 3),
           SE = round(unname(sm[rn, "Std. Error"]), 3),
           t = round(unname(sm[rn, "t value"]), 2))
  })
  if (nrow(lmm_res))
    kable(lmm_res, caption = paste("Mixed model, score ~ time x SAA + age + sex",
      "+ (1|subject). The term shown is the difference in slope, SAA+ minus",
      "SAA-. |t| > 2 is roughly p < 0.05.")) %>%
      kable_styling(full_width = FALSE, bootstrap_options = c("striped","condensed"))
  else cat("Not enough longitudinal data to fit the mixed model.")
}
```

```{r progression-plot, fig.width=9.5, fig.height=4.2}
pd <- long %>% filter(n_obs >= 2)
if (nrow(pd) >= 10) {
  ## free_y let each panel autoscale to whatever few points it had, so MoCA and
  ## MMSE - both 0-30 - came out on different axes and looked incomparable.
  ## Each test is pinned to its own real range instead.
  RANGES <- tibble(test = c("MoCA", "MMSE", "MDS-UPDRS III"),
                   lo = c(0, 0, 0), hi = c(30, 30, 132)) %>%
    filter(test %in% unique(pd$test))
  frame <- bind_rows(
    RANGES %>% transmute(test, t_years = 0, score = lo),
    RANGES %>% transmute(test, t_years = 0, score = hi))

  ggplot(pd, aes(t_years, score)) +
    geom_blank(data = frame, aes(t_years, score), inherit.aes = FALSE) +
    geom_line(aes(group = NHC, colour = saa_local_pos), alpha = .45) +
    geom_point(aes(colour = saa_local_pos), size = 1.3, alpha = .75) +
    geom_smooth(aes(group = saa_local_pos, colour = saa_local_pos),
                method = "lm", formula = y ~ x, se = TRUE, linewidth = 1.1) +
    facet_wrap(~ test, scales = "free_y") +
    scale_colour_manual(values = c(Negative = "#E69F00", Positive = "#56B4E9")) +
    labs(x = "Years from first assessment", y = "Score", colour = "SAA",
         title = "Trajectories by SAA status",
         caption = "Axes span each instrument's full range: MoCA and MMSE 0-30, MDS-UPDRS III 0-132.") +
    theme(legend.position = "bottom")
} else cat("Too few repeated assessments to plot.")
```

# UPDRS III and physical performance (SPPB)

```{r updrs-sppb}
## The original showed a scatter plot only. Correlations are quantified here,
## per group, with a Fisher r-to-z test of whether they differ.
cor_one <- function(dat, lab) {
  x <- num_safe(dat$updrs3_total_score); y <- num_safe(dat$sppb_total_score)
  ok <- !is.na(x) & !is.na(y); n <- sum(ok)
  if (n < 5) return(tibble(Group = lab, n = n, rho = NA, p = NA, Pearson_r = NA))
  sp <- suppressWarnings(cor.test(x[ok], y[ok], method = "spearman", exact = FALSE))
  pe <- suppressWarnings(cor.test(x[ok], y[ok], method = "pearson"))
  tibble(Group = lab, n = n, rho = round(unname(sp$estimate), 3),
         p = signif(sp$p.value, 3), Pearson_r = round(unname(pe$estimate), 3))
}
cor_res <- bind_rows(
  cor_one(df_saa2 %>% filter(saa_local_pos == "Positive"), "SAA positive"),
  cor_one(df_saa2 %>% filter(saa_local_pos == "Negative"), "SAA negative"),
  cor_one(df_saa2, "All (combined)"),
  cor_one(df_saa2, "All (Sant Pau)"))

a <- cor_res[1,]; b <- cor_res[2,]
fisher_line <- if (!is.na(a$rho) && !is.na(b$rho) && a$n > 3 && b$n > 3) {
  z <- (atanh(a$rho) - atanh(b$rho)) / sqrt(1/(a$n-3) + 1/(b$n-3))
  sprintf("Fisher r-to-z, SAA+ vs SAA−: z = %.2f, p = %.3f", z, 2*pnorm(-abs(z)))
} else "Too few complete pairs to compare the two correlations."

kable(cor_res, caption = paste("UPDRS III vs SPPB.", fisher_line)) %>%
  kable_styling(full_width = FALSE, bootstrap_options = c("striped","condensed"))
write.csv(cor_res, file.path(results_path, "Correlation_UPDRS_SPPB.csv"), row.names = FALSE)
```

```{r updrs-sppb-plot, fig.width=7.5, fig.height=5}
pd <- df_saa2 %>%
  mutate(u = num_safe(updrs3_total_score), s = num_safe(sppb_total_score)) %>%
  filter(!is.na(u), !is.na(s))
if (nrow(pd) >= 5) {
  ggplot(pd, aes(u, s, colour = saa_local_pos)) +
    geom_point(alpha = .75, size = 2.6) +
    geom_smooth(method = "lm", se = TRUE, formula = y ~ x) +
    scale_colour_manual(values = c(Negative = "#E69F00", Positive = "#56B4E9")) +
    labs(title = "UPDRS III vs SPPB by SAA status",
         subtitle = sprintf("n = %d complete pairs", nrow(pd)),
         x = "MDS-UPDRS III", y = "SPPB total", colour = "SAA status") +
    theme(legend.position = "bottom")
} else {
  cat(sprintf("Only %d subjects have both UPDRS III and SPPB — no plot drawn.", nrow(pd)))
}
```

# Summary of methodological choices

```{r summary-notes, results='asis'}
cat(sprintf("
- **Scope.** Sant Pau (SAA02) only, clinical variables and A/T status. Other
  biofluid biomarkers and the MRI analyses are parked; they are recoverable from
  the git history of this file.
- **Merge.** Each clinical visit paired with the biofluid sample closest in time
  (`visit_date`, not the baseline date). %s
- **Long-format REDCap.** Subject-level fields carried down within NHC before
  collapsing to one row per subject, so core symptoms and visit scores coexist.
  The progression section deliberately uses all visits instead.
- **Reading.** `guess_max = Inf`, without which readxl types the visit-level
  columns from the empty enrolment rows and discards their data.
- **Recoding.** Anchored SAA patterns and substring A/T patterns; every raw level
  is printed above for checking. Indeterminate SAA excluded (n = %d).
- **Plausibility limits.** Values outside each scale range are set to NA and
  counted; the guard finds nothing in this dataset and remains as a regression
  check.
- **Tests.** Rank-based for continuous, Fisher for categorical, with SMD next to
  the mean or risk difference. No multiplicity adjustment: the comparisons are
  pre-specified and direction-led.
", merge_note, n_indet))
```

# References

Found with Undermind and read against the results above. The note on each line
is what it says about our finding, not a summary of the paper.

**Clinical features**

1. Coughlin DG, MacLeod KR, Middleton JS, et al. *Association of CSF
   alpha-Synuclein Seeding Amplification Assay Results With Clinical Features of
   Possible and Probable Dementia With Lewy Bodies.* Neurology 2024;103(3):e209656.
   doi:10.1212/WNL.0000000000209656
   — n = 191 DLB. RBD in 86% of SAA+ vs 62% of SAA- (p < 0.0001, OR 3.63):
   our RBD result replicates with a smaller effect. Hallucinations 50% vs 47%
   (p = 0.73): our null replicates theirs. MDS-UPDRS III 33.8 vs 25.6
   (p = 0.001): **we do not replicate this** — our UPDRS III and parkinsonism
   prevalence are flat. DaT-SPECT was not analysed against SAA status, so our
   DaT result has no direct published comparator.

**AD co-pathology**

2. Honig LS, Kang MS, Lamoureux J, Marder K. *Cerebrospinal fluid seed
   aggregation assay and Alzheimer biomarkers in a Dementia with Lewy Body
   cohort.* Alzheimers Dement 2025;21(Suppl 2):e107531. doi:10.1002/alz70856_107531
   — No difference in amyloid or tau biomarker status between SAA+ and SAA-.
   Our A/T null replicates this.

3. Burton EJ, Barber R, Mukaetova-Ladinska EB, et al. *Medial temporal lobe
   atrophy on MRI differentiates Alzheimer's disease from dementia with Lewy
   bodies and vascular cognitive impairment.* Brain 2009;132(1):195-203.
   doi:10.1093/brain/awn298
   — The classic MTA result, DLB vs AD rather than SAA-stratified. Our visual
   MTA finding (higher in SAA-) sits in this tradition.

**Where this leaves us.** RBD and hallucinations replicate. Parkinsonism does
not. DaT-SCAN against SAA status appears not to have been reported, and is the
most novel of the clinical results.

# Reproducibility

```{r session}
sessionInfo()
```