---
title: "DLB Cohort Analysis — CSF α-synuclein SAA"
subtitle: "Clinical, biofluid and imaging profile by seed-amplification assay status"
author: "E. Bernardi"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: true
    code_folding: hide
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.align = "center", dpi = 150)

library(readxl);  library(dplyr);   library(tidyr);   library(lubridate)
library(stringr); library(purrr);   library(tibble)
library(gtsummary); library(gt);    library(ggplot2); library(knitr)
library(kableExtra)

theme_set(theme_minimal(base_size = 12))
set.seed(20260826)
```

```{r config}
## ---------------------------------------------------------------------------
## PATHS
## ---------------------------------------------------------------------------
desktop_path <- "~/Desktop/Ricerca/Lewy_R/SAA"
results_path <- file.path(desktop_path, "Results")
dir.create(results_path, recursive = TRUE, showWarnings = FALSE)

f_clin  <- file.path(desktop_path, "datafreeze_lewy.xlsx")
f_bio   <- file.path(desktop_path, "BIOFLUIDS_CLINICAL.xlsx")
f_cat12 <- file.path(desktop_path, "CAT12_Hammers_DK_SPIN.xlsx")
## Optional: FreeSurfer normative Z from the visual-hallucination pipeline.
## If present, the imaging section adds a cross-pipeline check on the same
## subjects; if absent, that subsection is skipped and everything else runs.

have_cat12 <- file.exists(f_cat12)
## Default so that chunk-level eval conditions resolve even when the imaging
## file is absent and the chunk that sets this never runs.
imaging_ok  <- FALSE
restrict_ok <- FALSE

## ---------------------------------------------------------------------------
## MERGE OPTIONS
## ---------------------------------------------------------------------------
MAX_GAP_DAYS <- 365   # visit <-> sample; pairs beyond this are flagged
MAX_MRI_GAP  <- 730   # sample <-> MRI
## Image-quality threshold, set in CAT12's native rating (LOWER is better,
## 1 = A+). A rating of 3 is the accepted cut-off here. CAT12 may export this
## either as the rating or as the percentage grade (higher better); the scale is
## detected on load and this threshold is converted to match.
IQR_MAX_RATING <- 3
IQR_MIN        <- 105 - 10 * IQR_MAX_RATING   # = 75% on the percentage scale
MIN_N_MODEL  <- 15    # minimum complete cases to fit an imaging model

## ---------------------------------------------------------------------------
## DATA-QUALITY GUARDS
## Values outside these ranges are set to NA and reported in section 2.
## Kept as a regression check: an earlier run produced Mayo = 21 (SD 62) on a
## 0-4 scale, which turned out to come from the baseline-date row selection
## rather than from sentinel codes. If this table is ever non-empty, the source
## column is wrong.
## ---------------------------------------------------------------------------
SENTINELS <- c(-9, -99, -999, 888, 999, 9999, -1)

PLAUSIBLE <- tribble(
  ~var,                    ~lo, ~hi,
  "edad",                    18, 110,
  "EDUC",                     0,  30,
  "puntuaciones",             0,  30,   # MoCA total
  "visuoespacialp",           0,   5,
  "identificacionp",          0,   3,
  "atencionp",                0,   6,
  "lenguajep",                0,   3,
  "abstraccionp",             0,   2,
  "orientacionp",             0,   6,
  "mis",                      0,  15,   # MoCA memory index
  "mayo_total_score",         0,   4,   # 4 dichotomous items
  "updrs1a_total_score",      0,  24,
  "updrs1b_total_score",      0,  28,
  "updrs2_total_score",       0,  52,
  "updrs3_total_score",       0, 132,
  "hoehn_yahr",               0,   5,
  "sppb_total_score",         0,  12,
  "sppb_subtotal1",           0,   4,
  "sppb_subtotal2",           0,   4,
  "sppb_subtotal3",           0,   4,
  "mta",                      0,   4,
  "fazekas",                  0,   3,
  "MMSE",                     0,  30,
  "GDS",                      1,   7,
  "gds_visit",                1,   7
)

## ---------------------------------------------------------------------------
## ANALYTE PANEL — only assays with real coverage are tabulated.
## The previous run printed "NA (NA)" rows for sTREM2, SNAP-25, NPTX2, TDP-43
## and two plasma H0010 assays (n <= 1). MIN_PER_GROUP removes them.
## ---------------------------------------------------------------------------
MIN_PER_GROUP <- 8
## assays reported on a log scale (right-skewed concentrations)
LOG_SCALE <- "NFL|NFLIGHT|GFAP|YKL40|STREM2|TREM2|TAU|TDP43|SNAP25|NPTX2|ABETA|AB4|AB1"
```

# Data loading and merge

```{r load}
## ---------------------------------------------------------------------------
## guess_max = Inf is essential. readxl types each column from the first 1000
## rows; in a long REDCap export the visit-level columns (UPDRS, MoCA, SPPB)
## are empty on the leading enrolment rows and would be typed <logical>,
## silently discarding every value below row 1000.
## ---------------------------------------------------------------------------
read_x <- function(p) read_excel(p, guess_max = Inf, .name_repair = "minimal")

clin_raw <- read_x(f_clin)
bio_raw  <- read_x(f_bio)

## drop blank / duplicated column names produced by Excel
tidy_names <- function(df) {
  df <- df[, names(df) != "" & !is.na(names(df)), drop = FALSE]
  df[, !duplicated(names(df)), drop = FALSE]
}
clin_raw <- tidy_names(clin_raw); bio_raw <- tidy_names(bio_raw)

## ---- helpers --------------------------------------------------------------
norm_id <- function(x) {                       # 0012345 == 12345
  x <- toupper(trimws(as.character(x)))
  x[x %in% c("", "NA", "NULL", "-", "0")] <- NA
  x <- gsub("[^0-9A-Z]", "", x)
  num <- !is.na(x) & grepl("^[0-9]+$", x)
  x[num] <- sub("^0+", "", x[num])
  x[!is.na(x) & x == ""] <- NA
  x
}

as_date_safe <- function(x) {                  # Date / POSIXct / serial / text
  if (inherits(x, "Date"))   return(x)
  if (inherits(x, "POSIXt")) return(as.Date(x))
  if (is.numeric(x)) {
    out <- as.Date(rep(NA, length(x)))
    ok <- !is.na(x) & x > 10000 & x < 60000
    out[ok] <- as.Date(x[ok], origin = "1899-12-30"); return(out)
  }
  x <- trimws(as.character(x))
  x[x %in% c("", "NA", "NaN", "NULL", "-", "#N/A")] <- NA
  out <- as.Date(rep(NA, length(x)))
  ser <- !is.na(x) & grepl("^[0-9]{5}(\\.[0-9]+)?$", x)
  out[ser] <- as.Date(as.numeric(x[ser]), origin = "1899-12-30")
  for (f in c("%Y-%m-%d","%d/%m/%Y","%d-%m-%Y","%Y/%m/%d","%d.%m.%Y","%Y%m%d")) {
    todo <- !is.na(x) & is.na(out)
    if (!any(todo)) break
    out[todo] <- as.Date(substr(x[todo], 1, 10), format = f)
  }
  out[!is.na(out) & (out < as.Date("1900-01-01") | out > Sys.Date() + 365)] <- NA
  out
}

num_safe <- function(x) {                      # "<200", "1,5", " 12 "
  if (is.numeric(x)) return(x)
  v <- gsub("\\s", "", gsub("[<>=]", "", trimws(as.character(x))))
  v <- ifelse(grepl(",", v) & !grepl("\\.", v), gsub(",", ".", v), gsub(",", "", v))
  suppressWarnings(as.numeric(v))
}

has_col <- function(df, x) x %in% names(df)
```

```{r clinical-reshape}
## ---------------------------------------------------------------------------
## The REDCap export is LONG: subject-level fields sit on the enrolment event,
## visit-level fields on repeating instances. Collapsing without carrying the
## subject-level fields down loses either the core symptoms or the scores.
## ---------------------------------------------------------------------------
nhc_col <- grep("^nhc$", names(clin_raw), ignore.case = TRUE, value = TRUE)[1]
stopifnot(!is.na(nhc_col))

clin <- clin_raw %>%
  mutate(NHC = norm_id(.data[[nhc_col]])) %>%
  filter(!is.na(NHC))

## Per-visit date. THE ORIGINAL SCRIPT USED visit_date_base_lewy, which is the
## baseline date (constant per subject) -- the "closest visit" rule was
## therefore never applied. visit_date is the per-instance column.
clin$visit_d <- as_date_safe(if (has_col(clin, "visit_date")) clin$visit_date else NA)
clin$visit_fallback <- as_date_safe(
  if (has_col(clin, "visit_date_base_lewy")) clin$visit_date_base_lewy else NA)
clin$visit_any <- coalesce(clin$visit_d, clin$visit_fallback)

## Columns that belong to a VISIT (never carried across rows)
visit_pat <- paste0("^(visit_date|redcap_event_instance|neurologist|visitdiagnosis|",
  "gds_visit|medication|hoehn_yahr|diagnostico_visita|updrs|sppb|mayo|zarit|",
  "lethargic|sleep_day|disorganized|staring|fecha|puntuaciones|resultado|mis$|",
  "visuoespacial|identificacion|atencion|lenguaje|abstraccion|orientacion|",
  "at_|len_|directos|inversos|letras|serie7|repeticion|fluidez|num_palab|",
  "rostro|seda|templo|clavel|rojo|speech|facial|rigidity|tapping|handmovements|",
  "pronation|legagility|arising|gait|freezing|postural|globalspont|kinetictremor|",
  "resttremor|constancy|dyskinesia|posture|movilidad|autocuidado|actividades|",
  "dolor_|ansiedad|depresion|termometro|a_[0-9]|d_[0-9]|resumen_|seguimiento)")
visit_cols   <- grep(visit_pat, names(clin), value = TRUE)
subject_cols <- setdiff(names(clin), c(visit_cols, "NHC", "visit_d",
                                       "visit_fallback", "visit_any"))

clin <- clin %>%
  group_by(NHC) %>%
  mutate(across(all_of(subject_cols),
                ~ if (all(is.na(.x))) .x else coalesce(.x, .x[which(!is.na(.x))[1]]))) %>%
  ungroup()

clin_visits <- clin %>% filter(!is.na(visit_any)) %>%
  distinct(NHC, visit_any, .keep_all = TRUE)
clin_nodate <- clin %>% filter(!NHC %in% clin_visits$NHC) %>%
  distinct(NHC, .keep_all = TRUE)
clin_use <- bind_rows(clin_visits, clin_nodate)
```

```{r biofluid-merge}
bio <- bio_raw %>%
  mutate(NHC  = norm_id(.data[["NHC"]]),
         NHCb = if (has_col(bio_raw, "NHC2")) norm_id(.data[["NHC2"]]) else NA_character_) %>%
  mutate(NHC = coalesce(NHC, NHCb)) %>%
  filter(!is.na(NHC)) %>%
  distinct(across(everything()), .keep_all = TRUE)

bio$sample_d <- as_date_safe(bio$SAMPLEDATE)

join_args <- list(x = clin_use, y = bio, by = "NHC", suffix = c("_clin", "_bio"))
if (utils::packageVersion("dplyr") >= "1.1.0")
  join_args$relationship <- "many-to-many"

merged_df <- suppressWarnings(do.call(inner_join, join_args)) %>%
  mutate(date_diff = abs(as.numeric(sample_d - visit_any)),
         .rank = ifelse(is.na(date_diff), Inf, date_diff)) %>%
  group_by(NHC) %>%
  arrange(.rank, visit_any, sample_d, .by_group = TRUE) %>%
  slice(1) %>% ungroup() %>%
  mutate(gap_over_limit = !is.na(date_diff) & date_diff > MAX_GAP_DAYS)

merge_note <- sprintf(
  paste("Clinical: %s subjects / %s dated visits. Biofluids: %s subjects / %s samples.",
        "Merged: %s subjects. Visit-sample gap: median %s d [IQR %s-%s], %s%% within 90 d,",
        "%s pair(s) beyond %s d."),
  n_distinct(clin$NHC), nrow(clin_visits), n_distinct(bio$NHC), nrow(bio),
  nrow(merged_df),
  round(median(merged_df$date_diff, na.rm = TRUE)),
  round(quantile(merged_df$date_diff, .25, na.rm = TRUE)),
  round(quantile(merged_df$date_diff, .75, na.rm = TRUE)),
  round(100 * mean(merged_df$date_diff <= 90, na.rm = TRUE)),
  sum(merged_df$gap_over_limit, na.rm = TRUE), MAX_GAP_DAYS)
cat(merge_note)
```

```{r gap-plot, fig.width=6.5, fig.height=3}
ggplot(merged_df %>% filter(!is.na(date_diff)), aes(date_diff)) +
  geom_histogram(bins = 40, fill = "#4C72B0", colour = "white") +
  geom_vline(xintercept = c(90, MAX_GAP_DAYS), linetype = c("dashed", "dotted"),
             colour = c("#D55E00", "grey40")) +
  labs(title = "Interval between clinical visit and biofluid sampling",
       x = "|visit date − sample date| (days)", y = "Subjects",
       caption = "Dashed = 90 d; dotted = analysis limit")
```

# Data quality

```{r derive}
## ---- safe recoders --------------------------------------------------------
## ANCHORED patterns. The original used str_detect(x, "pos|1"), which matches
## any string containing a "1" (e.g. "10", "0.15", a plate code).
recode_saa <- function(x) {
  v <- tolower(trimws(as.character(x)))
  out <- rep(NA_character_, length(v))
  out[grepl("^(0|neg|negative|negativo|negativa)$", v)] <- "Negative"
  out[grepl("^(1|pos|positive|positivo|positiva)$", v)] <- "Positive"
  out[grepl("^(2|3|9|ind|indet|indeterminate|inconclusive|dudoso|equivocal)$", v)] <-
    "Indeterminate"
  factor(out, levels = c("Negative", "Positive", "Indeterminate"))
}

recode_yesno <- function(x) {
  v <- tolower(trimws(as.character(x)))
  v[v %in% c("", "na", "null", "-", "9", "99", "unknown", "desconocido", "nd")] <- NA
  out <- rep(NA_character_, length(v))
  out[grepl("^(0|no|n|absent|ausente|negative|negativo|normal|false)$", v)] <- "No"
  out[grepl("^(1|2|yes|y|si|s[ií]|present|presente|positive|positivo|abnormal|anormal|patologico|true)$", v)] <- "Yes"
  factor(out, levels = c("No", "Yes"))
}

## Free-text/multi-level fields that must be collapsed rather than shown raw.
## The previous run printed DaT-SCAN and PSG with their full level sets
## (including "Scheduled", "NA", "RBD+SAHS+PLM"), producing an uninterpretable
## chi-squared across administrative categories.
recode_datscan <- function(x) {
  v <- tolower(trimws(as.character(x)))
  out <- rep(NA_character_, length(v))
  out[grepl("normal", v) & !grepl("abnormal|anormal|patolog", v)] <- "Normal"
  out[grepl("abnormal|anormal|patolog|altered|alterado", v)] <- "Abnormal"
  factor(out, levels = c("Normal", "Abnormal"))
}
recode_psg <- function(x) {                 # RBD present on PSG, yes/no
  v <- tolower(trimws(as.character(x)))
  out <- rep(NA_character_, length(v))
  out[grepl("no rbd|sin rbd", v)] <- "No"
  out[grepl("rbd", v) & !grepl("no rbd|sin rbd", v)] <- "Yes"
  out[grepl("scheduled|pending|programad|^na$|^$", v)] <- NA
  factor(out, levels = c("No", "Yes"))
}

## A/T status. The previous anchored version silently matched nothing on this
## export -- A_clean and AT_profile then vanished from every table because
## keep_usable() drops all-missing variables without comment. Substring matching
## is used instead, and the raw-to-recoded mapping is printed below so the
## result can be checked rather than assumed.
recode_status <- function(x, letter) {
  v <- gsub("\\s+", "", tolower(trimws(as.character(x))))
  v[v %in% c("", "na", "nan", "null", "nd", "unknown", "desconocido", "?")] <- NA
  out <- rep(NA_character_, length(v))
  lo <- tolower(letter)
  neg <- !is.na(v) & (grepl("neg", v) | v %in% c("0", "false", "-", paste0(lo, "-")))
  pos <- !is.na(v) & (grepl("pos", v) | v %in% c("1", "true", "+", paste0(lo, "+")))
  out[neg] <- paste0(letter, "-")
  out[pos] <- paste0(letter, "+")      # positive wins if both somehow match
  factor(out, levels = paste0(letter, c("-", "+")))
}

clamp <- function(x, lo, hi, label) {
  v <- num_safe(x)
  v[v %in% SENTINELS] <- NA
  bad <- !is.na(v) & (v < lo | v > hi)
  if (any(bad)) qc_log[[label]] <<- sum(bad)
  v[bad] <- NA
  v
}

qc_log <- list()

analytic_df <- merged_df
analytic_df$saa3_raw <- if (has_col(analytic_df, "CSF_ALPHASYN_SAA03"))
  recode_saa(analytic_df$CSF_ALPHASYN_SAA03) else factor(rep(NA, nrow(analytic_df)),
                                                         levels = c("Negative","Positive"))

## apply plausibility limits
for (i in seq_len(nrow(PLAUSIBLE))) {
  vn <- PLAUSIBLE$var[i]
  if (has_col(analytic_df, vn))
    analytic_df[[vn]] <- clamp(analytic_df[[vn]], PLAUSIBLE$lo[i],
                               PLAUSIBLE$hi[i], vn)
}

analytic_df <- analytic_df %>%
  mutate(
    SEX = factor(ifelse(grepl("^(f|female|mujer|femenino)$", tolower(trimws(SEX))),
                        "Female",
                 ifelse(grepl("^(m|male|hombre|var|masculino)", tolower(trimws(SEX))),
                        "Male", NA)), levels = c("Female", "Male")),

    saa_local_pos  = recode_saa(CSF_ALPHASYN_SAA02),
    saa_munich_pos = recode_saa(CSF_ALPHASYN_SAA01),
    saa_third      = saa3_raw,

    ## Own-site result preferred, Munich as fallback. The original rule coded
    ## ANY positive as Positive, which silently resolved every discordant pair
    ## in favour of Positive -- incompatible with the concordance analysis.
    saa_any_pos = factor(coalesce(as.character(saa_local_pos),
                                  as.character(saa_munich_pos)),
                         levels = c("Negative", "Positive", "Indeterminate")),

    A_clean = recode_status(ASTATUS, "A"),
    T_clean = recode_status(TSTATUS, "T"),
    AT_profile = factor(if_else(is.na(A_clean) | is.na(T_clean), NA_character_,
                                paste0(A_clean, " / ", T_clean)),
                        levels = c("A- / T-", "A- / T+", "A+ / T-", "A+ / T+")),

    ## APOE: genotype string -> carrier. Guarded against a dosage-coded column.
    apoe4_carrier = {
      g <- gsub("[^0-9]", "", as.character(APOE))
      factor(ifelse(is.na(g) | nchar(g) < 2, NA,
                    ifelse(grepl("4", g), "Yes", "No")), levels = c("No", "Yes"))
    },

    hallucinations = recode_yesno(hallucinations),
    fluctuations   = recode_yesno(fluctuations),
    rbdclinical    = recode_yesno(rbdclinical),
    parkinsonism   = recode_yesno(parkinsonism),
    hyposmia       = recode_yesno(hyposmia),
    dysautonomia   = recode_yesno(dysautonomia),
    repeatedfalls  = recode_yesno(repeatedfalls),
    hypersomnia    = recode_yesno(hypersomnia),
    neurolepticsens= recode_yesno(neurolepticsens),
    datscan_bin    = recode_datscan(datscan),
    psg_rbd        = recode_psg(psg),

    ## age at sampling beats a stored age from an unrelated timepoint
    age_at_sample = {
      a <- as.numeric(sample_d - as_date_safe(fnac)) / 365.25
      a[!is.na(a) & (a < 18 | a > 110)] <- NA
      coalesce(a, num_safe(edad), num_safe(AGE))
    },

    n_core = rowSums(cbind(hallucinations == "Yes", fluctuations == "Yes",
                           rbdclinical == "Yes", parkinsonism == "Yes"), na.rm = TRUE),
    n_core = ifelse(rowSums(!is.na(cbind(hallucinations, fluctuations,
                                         rbdclinical, parkinsonism))) == 0, NA, n_core),

    ## UPDRS total: NA unless all four parts are present (partial sums are not
    ## comparable across subjects)
    total_updrs = ifelse(
      is.na(updrs1a_total_score) | is.na(updrs1b_total_score) |
      is.na(updrs2_total_score)  | is.na(updrs3_total_score), NA,
      updrs1a_total_score + updrs1b_total_score +
      updrs2_total_score + updrs3_total_score)
  )
```

```{r pd-phenotype}
## ---------------------------------------------------------------------------
## Motor phenotype (Jankovic): mean tremor / mean akinetic-rigid, from the
## MDS-UPDRS III items. Absent from the original script.
## ---------------------------------------------------------------------------
TREM <- c("posturaltremorr","posturaltremorl","kinetictremorr","kinetictremorl",
          "resttremorrue","resttremorlue","resttremorrle","resttremorlle",
          "resttremorlipjaw","constancy")
AKRI <- c("speech","facialexpression","rigidityneck","rigidityrue","rigiditylue",
          "rigidityrle","rigiditylle","tappingrue","tappinglue","handmovementsr",
          "handmovementsl","pronationsupinationr","pronationsupinationl",
          "tappingrle","tappinglle","legagilityr","legagilityl","arising","gait",
          "freezing","posturalstability","posture","globalspontaneityl")

item_matrix <- function(df, items) {
  got <- intersect(items, names(df))
  if (!length(got)) return(list(m = NULL, used = character()))
  m <- sapply(got, function(cc) {
    v <- num_safe(df[[cc]]); v[!is.na(v) & (v < 0 | v > 4)] <- NA; v
  })
  list(m = as.matrix(m), used = got)
}
tr <- item_matrix(analytic_df, TREM); ak <- item_matrix(analytic_df, AKRI)

mean_min <- function(m, k) {
  if (is.null(m)) return(rep(NA_real_, nrow(analytic_df)))
  out <- rowMeans(m, na.rm = TRUE); out[rowSums(!is.na(m)) < k] <- NA_real_; out
}
analytic_df$tremor_score  <- mean_min(tr$m, max(3, length(tr$used) %/% 2))
analytic_df$akrigid_score <- mean_min(ak$m, max(5, length(ak$used) %/% 2))
analytic_df <- analytic_df %>%
  mutate(td_ar_ratio = ifelse(!is.na(akrigid_score) & akrigid_score > 0,
                              tremor_score / akrigid_score, NA_real_),
         pd_phenotype = factor(case_when(
           is.na(td_ar_ratio)   ~ NA_character_,
           td_ar_ratio >= 1.0   ~ "Tremor-dominant",
           td_ar_ratio <= 0.8   ~ "Akinetic-rigid",
           TRUE                 ~ "Mixed"),
           levels = c("Akinetic-rigid", "Mixed", "Tremor-dominant")))

cat(sprintf("UPDRS-III items found: %d/%d tremor, %d/%d akinetic-rigid. Phenotype classifiable in %d subjects.",
            length(tr$used), length(TREM), length(ak$used), length(AKRI),
            sum(!is.na(analytic_df$pd_phenotype))))
```

```{r qc-report}
qc_tbl <- tibble(Variable = names(qc_log),
                 `Implausible values set to NA` = unlist(qc_log))
if (nrow(qc_tbl)) {
  kable(qc_tbl, caption = paste("Out-of-range values removed. Any non-zero row",
        "means the source column contains sentinel codes or the wrong field.")) %>%
    kable_styling(full_width = FALSE, bootstrap_options = c("striped","condensed"))
} else cat("No out-of-range values detected.")

saa_raw_check <- bind_rows(
  tibble(Assay = "SAA02 (Sant Pau)", Raw = as.character(merged_df$CSF_ALPHASYN_SAA02)),
  tibble(Assay = "SAA01 (Munich)",   Raw = as.character(merged_df$CSF_ALPHASYN_SAA01))) %>%
  filter(!is.na(Raw)) %>% count(Assay, Raw, sort = TRUE)
kable(saa_raw_check, caption = "Raw SAA values before recoding — verify every level maps as intended.") %>%
  kable_styling(full_width = FALSE, bootstrap_options = c("striped","condensed"))

at_raw_check <- bind_rows(
  if (has_col(merged_df, "ASTATUS"))
    tibble(Field = "ASTATUS", Raw = as.character(merged_df$ASTATUS),
           Recoded = as.character(recode_status(merged_df$ASTATUS, "A"))) else NULL,
  if (has_col(merged_df, "TSTATUS"))
    tibble(Field = "TSTATUS", Raw = as.character(merged_df$TSTATUS),
           Recoded = as.character(recode_status(merged_df$TSTATUS, "T"))) else NULL) %>%
  filter(!is.na(Raw)) %>% count(Field, Raw, Recoded, sort = TRUE)
if (nrow(at_raw_check))
  kable(at_raw_check, caption = paste("Amyloid / tau status: raw value to recoded",
    "value. Any row with Recoded = NA is being lost.")) %>%
    kable_styling(full_width = FALSE, bootstrap_options = c("striped","condensed"))
```

```{r cohorts}
## Indeterminate results are excluded from group comparisons, not silently merged.
n_indet <- sum(analytic_df$saa_any_pos == "Indeterminate", na.rm = TRUE)

df_saa2 <- analytic_df %>% filter(saa_local_pos %in% c("Negative","Positive")) %>%
  mutate(saa_local_pos = droplevels(saa_local_pos))
df_both <- analytic_df %>% filter(saa_any_pos %in% c("Negative","Positive")) %>%
  mutate(saa_any_pos = droplevels(saa_any_pos))

saveRDS(list(analytic_df = analytic_df, df_saa2 = df_saa2, df_both = df_both),
        file.path(results_path, "analytic_df.rds"))

cat(sprintf("Sant Pau (SAA02): n = %d (%d+ / %d-).\nCombined (SAA01+SAA02): n = %d (%d+ / %d-).\nIndeterminate excluded: %d.",
  nrow(df_saa2), sum(df_saa2$saa_local_pos=="Positive"), sum(df_saa2$saa_local_pos=="Negative"),
  nrow(df_both), sum(df_both$saa_any_pos=="Positive"), sum(df_both$saa_any_pos=="Negative"),
  n_indet))
```

# Clinical demographics

```{r table-helpers}
## ---- effect sizes ---------------------------------------------------------
smd_value <- function(x, g) {
  g <- droplevels(factor(g)); if (nlevels(g) != 2) return(NA_real_)
  i1 <- g == levels(g)[1]; i2 <- g == levels(g)[2]
  if (is.numeric(x)) {
    a <- x[i1]; a <- a[!is.na(a)]; b <- x[i2]; b <- b[!is.na(b)]
    if (length(a) < 2 || length(b) < 2) return(NA_real_)
    den <- sqrt((var(a) + var(b)) / 2)
    return(if (!is.finite(den) || den == 0) NA_real_ else (mean(a) - mean(b)) / den)
  }
  f <- droplevels(factor(x)); k <- nlevels(f); if (k < 2) return(NA_real_)
  p1 <- as.numeric(prop.table(table(f[i1])))[-k]
  p2 <- as.numeric(prop.table(table(f[i2])))[-k]
  if (anyNA(p1) || anyNA(p2)) return(NA_real_)
  if (k == 2) {
    den <- sqrt((p1*(1-p1) + p2*(1-p2)) / 2)
    return(if (!is.finite(den) || den == 0) NA_real_ else (p1 - p2) / den)
  }
  S <- ((diag(p1, k-1) - outer(p1,p1)) + (diag(p2, k-1) - outer(p2,p2))) / 2
  inv <- try(solve(S), silent = TRUE); if (inherits(inv,"try-error")) return(NA_real_)
  as.numeric(sqrt(max(0, t(p1-p2) %*% inv %*% (p1-p2))))
}

fn_smd <- function(data, variable, by, ...) {
  v <- smd_value(data[[variable]], data[[by]])
  if (is.na(v)) NA_character_ else sprintf("%.2f", v)
}

## Mean difference (continuous) or risk difference (binary), with 95% CI.
## Supplied as a custom stat rather than via add_difference(), which would
## collide with add_p() over the p.value column.
fn_md <- function(data, variable, by, ...) {
  x <- data[[variable]]; g <- droplevels(factor(data[[by]]))
  if (nlevels(g) != 2) return(NA_character_)
  i1 <- g == levels(g)[1]; i2 <- g == levels(g)[2]
  if (is.numeric(x)) {
    a <- x[i1]; b <- x[i2]
    if (sum(!is.na(a)) < 2 || sum(!is.na(b)) < 2) return(NA_character_)
    tt <- try(t.test(a, b), silent = TRUE)
    if (inherits(tt, "try-error")) return(NA_character_)
    return(sprintf("%.2f (%.2f, %.2f)", unname(tt$estimate[1] - tt$estimate[2]),
                   tt$conf.int[1], tt$conf.int[2]))
  }
  f <- droplevels(factor(x)); if (nlevels(f) != 2) return(NA_character_)
  hi <- levels(f)[2]
  x1 <- sum(f[i1] == hi, na.rm = TRUE); n1 <- sum(!is.na(f[i1]))
  x2 <- sum(f[i2] == hi, na.rm = TRUE); n2 <- sum(!is.na(f[i2]))
  if (n1 < 2 || n2 < 2) return(NA_character_)
  pt <- try(prop.test(c(x1, x2), c(n1, n2), correct = FALSE), silent = TRUE)
  if (inherits(pt, "try-error")) return(NA_character_)
  sprintf("%.1f%% (%.1f, %.1f)", 100*(x1/n1 - x2/n2),
          100*pt$conf.int[1], 100*pt$conf.int[2])
}

## A tbl_summary module: n, median [IQR] or n(%), non-parametric p, difference, SMD.
## Continuous tests are rank-based; the previous script used Welch t-tests on
## ordinal (MTA, Fazekas) and heavily skewed (GFAP, NfL) variables.
mod_cont <- function(data, vars, labels, group_var) {
  data %>% select(all_of(vars), all_of(group_var)) %>%
    tbl_summary(by = all_of(group_var), missing = "no",
                type = list(everything() ~ "continuous"),
                label = labels,
                statistic = list(everything() ~ "{median} [{p25}, {p75}]"),
                digits = list(everything() ~ 1)) %>%
    add_n() %>%
    add_p(test = everything() ~ "wilcox.test",
          pvalue_fun = ~ style_pvalue(.x, digits = 3)) %>%
    add_stat(fns = everything() ~ fn_md) %>%
    modify_header(add_stat_1 = "**Diff. (95% CI)**") %>%
    add_stat(fns = everything() ~ fn_smd) %>%
    modify_header(add_stat_2 = "**SMD**") %>%
    bold_p(t = 0.05)
}

mod_cat <- function(data, vars, labels, group_var) {
  data %>% select(all_of(vars), all_of(group_var)) %>%
    tbl_summary(by = all_of(group_var), missing = "no", label = labels) %>%
    add_n() %>%
    add_p(test = everything() ~ "fisher.test",
          pvalue_fun = ~ style_pvalue(.x, digits = 3)) %>%
    add_stat(fns = everything() ~ fn_md) %>%
    modify_header(add_stat_1 = "**Diff. (95% CI)**") %>%
    add_stat(fns = everything() ~ fn_smd) %>%
    modify_header(add_stat_2 = "**SMD**") %>%
    bold_p(t = 0.05)
}

keep_usable <- function(data, vars, min_n = 5) {
  vars[vapply(vars, function(v) {
    if (!has_col(data, v)) return(FALSE)
    x <- data[[v]]
    sum(!is.na(x)) >= min_n && length(unique(x[!is.na(x)])) > 1
  }, logical(1))]
}

format_gt_table <- function(tbl) {
  as_gt(tbl) %>%
    tab_style(style = list(cell_fill(color = "#E0E0E0"), cell_text(weight = "bold")),
              locations = cells_row_groups())
}
```

```{r clinical-table}
LAB <- c(
  SEX="Female sex", apoe4_carrier="APOE ε4 carrier", age_at_sample="Age at sampling (y)",
  EDUC="Education (y)", puntuaciones="MoCA total", visuoespacialp="  Visuospatial/executive",
  identificacionp="  Naming", atencionp="  Attention", lenguajep="  Language",
  abstraccionp="  Abstraction", orientacionp="  Orientation", mis="  Memory index",
  total_updrs="MDS-UPDRS total (IA+IB+II+III)", updrs1a_total_score="  Part IA",
  updrs1b_total_score="  Part IB", updrs2_total_score="  Part II",
  updrs3_total_score="  Part III", hoehn_yahr="Hoehn & Yahr",
  pd_phenotype="Motor phenotype", td_ar_ratio="Tremor / akinetic-rigid ratio",
  hallucinations="Visual hallucinations", fluctuations="Cognitive fluctuations",
  rbdclinical="Clinical RBD", parkinsonism="Parkinsonism", n_core="Number of core features",
  hyposmia="Hyposmia", dysautonomia="Dysautonomia", repeatedfalls="Repeated falls",
  hypersomnia="Hypersomnia", neurolepticsens="Neuroleptic sensitivity",
  mayo_total_score="Mayo Fluctuations Scale", sppb_total_score="SPPB total",
  datscan_bin="DaT-SCAN abnormal", psg_rbd="PSG-confirmed RBD",
  A_clean="Amyloid status", AT_profile="A/T profile", mta="MTA score", fazekas="Fazekas score")

lab_list <- function(v) setNames(as.list(unname(LAB[v])), v) %>%
  { imap(., function(l, n) as.formula(paste0("`", n, "` ~ '", l, "'"))) } %>% unname()

build_clinical <- function(data, group_var) {
  blocks <- list(
    list(t="cat",  v=c("SEX","apoe4_carrier"),                      h="Demographics"),
    list(t="cont", v=c("age_at_sample","EDUC"),                      h="Demographics"),
    list(t="cont", v=c("puntuaciones","visuoespacialp","identificacionp","atencionp",
                       "lenguajep","abstraccionp","orientacionp","mis"),
                                                                     h="Cognition (MoCA)"),
    list(t="cont", v=c("total_updrs","updrs1a_total_score","updrs1b_total_score",
                       "updrs2_total_score","updrs3_total_score","hoehn_yahr"),
                                                                     h="Motor (MDS-UPDRS)"),
    list(t="cat",  v="pd_phenotype",                                 h="Motor phenotype"),
    list(t="cat",  v=c("hallucinations","fluctuations","rbdclinical","parkinsonism"),
                                                                     h="Core features"),
    list(t="cont", v="n_core",                                       h="Core features"),
    list(t="cat",  v=c("hyposmia","dysautonomia","repeatedfalls","hypersomnia",
                       "neurolepticsens"),                           h="Supportive features"),
    list(t="cont", v=c("mayo_total_score","sppb_total_score"),        h="Scales"),
    list(t="cat",  v=c("datscan_bin","psg_rbd","A_clean","AT_profile"), h="Biomarkers & imaging"),
    list(t="cont", v=c("mta","fazekas"),                             h="Biomarkers & imaging"))

  built <- list(); heads <- character()
  for (b in blocks) {
    vv <- keep_usable(data, b$v)
    if (!length(vv)) next
    tb <- if (b$t == "cont") mod_cont(data, vv, lab_list(vv), group_var)
          else               mod_cat (data, vv, lab_list(vv), group_var)
    built <- c(built, list(tb)); heads <- c(heads, b$h)
  }
  tbl_stack(built, group_header = heads) %>%
    modify_header(label ~ "**Variable**") %>% bold_labels()
}
```

## Sant Pau cohort (SAA02)

```{r tbl-saa2}
gt_demo_saa2 <- build_clinical(df_saa2, "saa_local_pos") %>%
  format_gt_table() %>%
  tab_source_note("Continuous: median [IQR], Wilcoxon rank-sum. Categorical: n (%), Fisher exact. Difference = Negative − Positive (Welch). SMD = standardized mean difference. N = subjects with the variable.")
gtsave(gt_demo_saa2, file.path(results_path, "Clinical_SantPau.html"))
gt_demo_saa2
```

## Combined cohort (SAA01 + SAA02)

```{r tbl-both}
gt_demo_both <- build_clinical(df_both, "saa_any_pos") %>%
  format_gt_table() %>%
  tab_source_note("Own-site (SAA02) result used where available, Munich (SAA01) otherwise. This cohort CONTAINS the Sant Pau cohort — the two tables are not independent.")
gtsave(gt_demo_both, file.path(results_path, "Clinical_Combined.html"))
gt_demo_both
```

# Biofluid biomarkers

```{r analyte-panel}
PANEL <- list(
  `CSF core AD` = c(CSF_ABRATIO="Aβ42/40 ratio", CSF_ABETA42_E0100="Aβ42",
                    CSF_ABETA40_E0350="Aβ40", CSF_PTAU181_E0100="p-tau181",
                    CSF_TTAU_A0010="t-tau"),
  `CSF glial / synaptic` = c(CSF_NFL_E0600="NfL", CSF_GFAP_S0120="GFAP",
                    CSF_YKL40_E0500="YKL-40", CSF_STREM2_E2100="sTREM2",
                    CSF_SNAP25_S0500="SNAP-25", CSF_NPTX2_E0050="NPTX2",
                    CSF_TOTALTDP43_E0710="Total TDP-43"),
  `Plasma` = c(PLASMA_AB42_AB40_RATIO_H0010="Aβ42/40 ratio",
                    PLASMA_PTAU217_S0217="p-tau217 (S0217)",
                    PLASMA_PTAU217_H0010="p-tau217 (H0010)",
                    PLASMA_GFAP_S0120="GFAP", PLASMA_NFLIGHT_S0100="NfL"))

## Coverage gate: drop assays without MIN_PER_GROUP observations in BOTH groups.
coverage_ok <- function(data, v, group_var) {
  if (!has_col(data, v)) return(FALSE)
  x <- num_safe(data[[v]]); g <- data[[group_var]]
  all(tapply(!is.na(x), g, sum, default = 0) >= MIN_PER_GROUP)
}

build_analytes <- function(data, group_var) {
  built <- list(); heads <- character()
  for (pn in names(PANEL)) {
    vs <- names(PANEL[[pn]])
    vs <- vs[vapply(vs, coverage_ok, logical(1), data = data, group_var = group_var)]
    if (!length(vs)) next
    dd <- data %>% mutate(across(all_of(vs), num_safe))
    labs <- setNames(as.list(unname(PANEL[[pn]][vs])), vs) %>%
      imap(function(l, n) as.formula(paste0("`", n, "` ~ '", l, "'"))) %>% unname()
    tb <- dd %>% select(all_of(vs), all_of(group_var)) %>%
      tbl_summary(by = all_of(group_var), missing = "no",
                  type = list(everything() ~ "continuous"),
                  label = labs,
                  statistic = list(everything() ~ "{median} [{p25}, {p75}]")) %>%
      add_n() %>%
      add_p(test = everything() ~ "wilcox.test",
            pvalue_fun = ~ style_pvalue(.x, digits = 3)) %>%
      add_stat(fns = everything() ~ fn_smd) %>%
      modify_header(add_stat_1 = "**SMD**")
    built <- c(built, list(tb)); heads <- c(heads, pn)
  }
  if (!length(built)) return(NULL)
  tbl_stack(built, group_header = heads) %>%
    bold_p(t = 0.05) %>%
    modify_header(label ~ "**Biomarker**") %>% bold_labels()
}
```

```{r analyte-tables}
gt_analy_saa2 <- build_analytes(df_saa2, "saa_local_pos")
if (!is.null(gt_analy_saa2)) {
  gt_analy_saa2 <- gt_analy_saa2 %>% format_gt_table() %>%
    tab_source_note(sprintf("Assays with <%d observations per group are omitted. q = Benjamini-Hochberg across the whole panel.", MIN_PER_GROUP))
  gtsave(gt_analy_saa2, file.path(results_path, "Analytical_SantPau.html"))
}
gt_analy_saa2
```

```{r analyte-tables-both}
gt_analy_both <- build_analytes(df_both, "saa_any_pos")
if (!is.null(gt_analy_both)) {
  gt_analy_both <- gt_analy_both %>% format_gt_table()
  gtsave(gt_analy_both, file.path(results_path, "Analytical_Combined.html"))
}
gt_analy_both
```

```{r panel-fdr}
## FDR across the whole biomarker panel, computed independently of gtsummary
## (add_q is not available on stacked tables in every version).
panel_vars <- unlist(lapply(PANEL, names), use.names = FALSE)
panel_fdr <- map_dfr(panel_vars, function(v) {
  if (!has_col(df_both, v)) return(NULL)
  x <- num_safe(df_both[[v]]); g <- df_both$saa_any_pos
  ok <- !is.na(x) & !is.na(g)
  n1 <- sum(ok & g == "Positive"); n0 <- sum(ok & g == "Negative")
  if (n1 < MIN_PER_GROUP || n0 < MIN_PER_GROUP) return(NULL)
  w <- suppressWarnings(wilcox.test(x[g == "Positive"], x[g == "Negative"],
                                    exact = FALSE))
  tibble(Biomarker = v, n_pos = n1, n_neg = n0,
         SMD = round(smd_value(x, g), 3), p = w$p.value)
})
if (nrow(panel_fdr)) {
  panel_fdr <- panel_fdr %>% mutate(q_BH = p.adjust(p, "BH"),
                                    p = signif(p, 3), q_BH = signif(q_BH, 3)) %>%
    arrange(p)
  write.csv(panel_fdr, file.path(results_path, "Biomarker_panel_FDR.csv"),
            row.names = FALSE)
  kable(panel_fdr, caption = sprintf(paste(
    "Combined cohort, %d panel biomarkers with adequate coverage. This panel is",
    "exploratory and direction-led, so the raw p is the operative column;",
    "q_BH is shown for reference, not as a gate."), nrow(panel_fdr))) %>%
    kable_styling(full_width = FALSE, bootstrap_options = c("striped","condensed"))
} else cat("No panel biomarker reaches the coverage threshold.")
```

```{r gfap-plot, fig.width=6.5, fig.height=4}
## GFAP was the strongest signal in the previous run; shown on a log axis with
## the individual points, because a mean (SD) hides how few subjects drive it.
if (has_col(df_both, "CSF_GFAP_S0120")) {
  gd <- df_both %>% mutate(v = num_safe(CSF_GFAP_S0120)) %>% filter(!is.na(v))
  if (nrow(gd) >= 6)
    print(ggplot(gd, aes(saa_any_pos, v, fill = saa_any_pos)) +
      geom_boxplot(width = .45, outlier.shape = NA, alpha = .6) +
      geom_jitter(width = .12, size = 1.8, alpha = .75) +
      scale_y_log10() +
      scale_fill_manual(values = c(Negative = "#E69F00", Positive = "#56B4E9")) +
      labs(title = "CSF GFAP by SAA status", x = NULL,
           y = "CSF GFAP (log scale)",
           subtitle = sprintf("n = %d (%d negative, %d positive)", nrow(gd),
                              sum(gd$saa_any_pos == "Negative"),
                              sum(gd$saa_any_pos == "Positive"))) +
      theme(legend.position = "none"))
}
```

# Is the GFAP difference amyloid-mediated?

**Direction check first.** In this cohort CSF GFAP is *higher in SAA-positive*
(19,367 vs 10,824 pg/mL, p = 0.047), not in SAA-negative. MTA runs the other
way (2.0 in SAA-negative vs 1.0 in SAA-positive). The two therefore do not
share an AD-driven explanation — they point in opposite directions.

Mediation still needs amyloid to differ by SAA status. It does not (A+ 47% vs
54%, p = 0.56), so the indirect effect has nothing to work with. The tables
below report the mediator path first for that reason.

```{r mediation}
MED_BOOT <- 2000

med_one <- function(dat, outcome, mediator, covars = c("age_at_sample", "SEX"),
                    log_outcome = TRUE, R = MED_BOOT) {
  need <- c(outcome, mediator, "saa_any_pos", covars)
  need <- need[need %in% names(dat)]
  if (!all(c(outcome, mediator, "saa_any_pos") %in% need)) return(NULL)

  d <- dat %>% select(all_of(need)) %>%
    mutate(.y = num_safe(.data[[outcome]]),
           .x = factor(saa_any_pos, levels = c("Negative", "Positive")),
           .m = if (is.factor(.data[[mediator]]) || is.character(.data[[mediator]]))
                  as.numeric(factor(.data[[mediator]])) - 1
                else num_safe(.data[[mediator]])) %>%
    select(.y, .x, .m, all_of(intersect(covars, need))) %>%
    drop_na() %>% droplevels()

  if (log_outcome && all(d$.y > 0)) d$.y <- log(d$.y)
  if (nrow(d) < 12 || n_distinct(d$.x) < 2 || n_distinct(d$.m) < 2) return(NULL)

  cv <- intersect(covars, names(d))
  cv <- cv[vapply(cv, function(cc) {
    v <- d[[cc]]
    if (is.factor(v) || is.character(v)) length(unique(v[!is.na(v)])) >= 2
    else { vv <- suppressWarnings(stats::var(as.numeric(v), na.rm = TRUE))
           is.finite(vv) && vv > 0 }
  }, logical(1))]
  cvs <- if (length(cv)) paste0(" + `", paste(cv, collapse = "` + `"), "`") else ""

  est <- function(dd) {
    mt <- lm(as.formula(paste0(".y ~ .x", cvs)), data = dd)
    md <- lm(as.formula(paste0(".y ~ .x + .m", cvs)), data = dd)
    ma <- lm(as.formula(paste0(".m ~ .x", cvs)), data = dd)
    cf <- function(m, nm) { k <- coef(m); if (nm %in% names(k)) unname(k[nm]) else NA_real_ }
    tot <- cf(mt, ".xPositive"); dir <- cf(md, ".xPositive")
    c(total = tot, direct = dir, indirect = tot - dir,
      a = cf(ma, ".xPositive"), b = cf(md, ".m"),
      prop = if (is.na(tot) || abs(tot) < 1e-8) NA_real_ else (tot - dir) / tot)
  }
  obs <- est(d)
  bs <- replicate(R, {
    o <- try(est(d[sample(nrow(d), replace = TRUE), ]), silent = TRUE)
    if (inherits(o, "try-error")) rep(NA_real_, 6) else o
  })
  ci <- function(k) {
    v <- bs[k, ]; v <- v[is.finite(v)]
    if (length(v) < 50) c(NA_real_, NA_real_) else unname(quantile(v, c(.025, .975)))
  }
  ai <- ci("a"); ii <- ci("indirect")

  ## p for the a-path, from the same model
  pa <- tryCatch({
    ma <- lm(as.formula(paste0(".m ~ .x", cvs)), data = d)
    summary(ma)$coefficients[".xPositive", "Pr(>|t|)"] }, error = function(e) NA_real_)
  pt <- tryCatch({
    mt <- lm(as.formula(paste0(".y ~ .x", cvs)), data = d)
    summary(mt)$coefficients[".xPositive", "Pr(>|t|)"] }, error = function(e) NA_real_)

  tibble(Outcome = outcome, Mediator = mediator, n = nrow(d),
         n_pos = sum(d$.x == "Positive"),
         a_path = obs["a"], a_p = pa,
         a_CI = sprintf("[%.2f, %.2f]", ai[1], ai[2]),
         total = obs["total"], total_p = pt,
         direct = obs["direct"], indirect = obs["indirect"],
         indirect_CI = sprintf("[%.2f, %.2f]", ii[1], ii[2]),
         prop_mediated = obs["prop"],
         logged = log_outcome && all(dat[[outcome]] > 0, na.rm = TRUE))
}

MED_OUTCOMES <- c(CSF_GFAP_S0120 = TRUE, PLASMA_GFAP_S0120 = TRUE, mta = FALSE)
MED_MEDIATORS <- c("A_clean", "CSF_ABRATIO", "CSF_PTAU181_E0100")

med_res <- map_dfr(names(MED_OUTCOMES), function(o)
  map_dfr(MED_MEDIATORS, function(m)
    med_one(df_both, o, m, log_outcome = unname(MED_OUTCOMES[o]))))

if (!nrow(med_res)) {
  cat("No outcome/mediator pair had enough complete data to model.")
} else {
  write.csv(med_res, file.path(results_path, "Mediation_amyloid_GFAP.csv"),
            row.names = FALSE)
  med_res %>%
    transmute(Outcome, Mediator, n,
              `a: mediator ~ SAA` = sprintf("%.2f (p=%.3f)", a_path, a_p),
              `c: total` = sprintf("%.2f (p=%.3f)", total, total_p),
              `c-prime: direct` = sprintf("%.2f", direct),
              `indirect (95% CI)` = sprintf("%.2f %s", indirect, indirect_CI),
              `% mediated` = ifelse(is.na(prop_mediated), "—",
                                    sprintf("%.0f%%", 100 * prop_mediated))) %>%
    kable(caption = paste("Amyloid as a mediator of the SAA-GFAP and SAA-MTA",
      "associations. Coefficients are for SAA-positive vs negative, so a",
      "negative total means higher in SAA-negative. If the a-path is null the",
      "indirect effect cannot be interpreted, whatever it estimates.")) %>%
    kable_styling(full_width = FALSE, bootstrap_options = c("striped","condensed"),
                  font_size = 11)
}
```

```{r mediation-strata}
## The same question asked without a mediation model: does the GFAP difference
## exist inside each amyloid stratum? If it lives only in A+, it is amyloid
## dependent; if it is present in both, it is not.
strat <- map_dfr(c("CSF_GFAP_S0120", "PLASMA_GFAP_S0120", "mta"), function(o) {
  if (!has_col(df_both, o) || !has_col(df_both, "A_clean")) return(NULL)
  map_dfr(levels(df_both$A_clean), function(lv) {
    d <- df_both %>% filter(A_clean == lv) %>%
      mutate(v = num_safe(.data[[o]])) %>% filter(!is.na(v), !is.na(saa_any_pos))
    n1 <- sum(d$saa_any_pos == "Positive"); n0 <- sum(d$saa_any_pos == "Negative")
    if (n1 < 3 || n0 < 3) return(tibble(Outcome = o, Amyloid = lv,
      n_pos = n1, n_neg = n0, median_pos = NA_real_, median_neg = NA_real_,
      SMD = NA_real_, p = NA_real_))
    a <- d$v[d$saa_any_pos == "Positive"]; b <- d$v[d$saa_any_pos == "Negative"]
    tibble(Outcome = o, Amyloid = lv, n_pos = n1, n_neg = n0,
           median_pos = signif(median(a), 4), median_neg = signif(median(b), 4),
           SMD = round(smd_value(d$v, d$saa_any_pos), 2),
           p = signif(suppressWarnings(wilcox.test(a, b, exact = FALSE))$p.value, 3))
  })
})
if (nrow(strat)) {
  write.csv(strat, file.path(results_path, "Mediation_stratified_by_amyloid.csv"),
            row.names = FALSE)
  kable(strat, caption = paste("SAA+ vs SAA- within each amyloid stratum.",
    "A difference confined to A+ supports amyloid dependence; one present in",
    "both argues against it.")) %>%
    kable_styling(full_width = FALSE, bootstrap_options = c("striped","condensed"))
} else cat("Amyloid status is unavailable — check the A/T recoding table in section 2.")
```

```{r mediation-interaction}
## Formal test of whether the SAA effect on GFAP depends on amyloid status.
int_res <- map_dfr(c("CSF_GFAP_S0120", "PLASMA_GFAP_S0120", "mta"), function(o) {
  if (!has_col(df_both, o) || !has_col(df_both, "A_clean")) return(NULL)
  d <- df_both %>%
    mutate(v = num_safe(.data[[o]])) %>%
    select(v, saa_any_pos, A_clean, age_at_sample, SEX) %>% drop_na() %>% droplevels()
  if (nrow(d) < 16 || n_distinct(d$A_clean) < 2 || n_distinct(d$saa_any_pos) < 2)
    return(NULL)
  if (all(d$v > 0) && o != "mta") d$v <- log(d$v)
  m <- try(lm(v ~ saa_any_pos * A_clean + age_at_sample + SEX, data = d), silent = TRUE)
  if (inherits(m, "try-error")) return(NULL)
  sm <- summary(m)$coefficients
  rn <- grep(":", rownames(sm), value = TRUE)[1]
  if (is.na(rn)) return(NULL)
  tibble(Outcome = o, n = nrow(d),
         `SAA x amyloid` = round(unname(sm[rn, "Estimate"]), 3),
         p = signif(unname(sm[rn, "Pr(>|t|)"]), 3))
})
if (nrow(int_res))
  kable(int_res, caption = paste("Interaction between SAA status and amyloid.",
    "A significant term means the SAA effect differs by amyloid status;",
    "a null term at this sample size is uninformative, not reassuring.")) %>%
    kable_styling(full_width = FALSE, bootstrap_options = c("striped","condensed"))
```

# AD co-pathology within SAA-positive subjects

```{r synergy}
## The original stratified SAA+ subjects by median plasma p-tau217 and then
## tested plasma p-tau217 as an outcome -- circular by construction. It is kept
## here only as a labelled manipulation check, and every cell reports n.
saa_pos <- analytic_df %>%
  filter(saa_any_pos == "Positive", !is.na(num_safe(PLASMA_PTAU217_S0217))) %>%
  mutate(p217 = num_safe(PLASMA_PTAU217_S0217),
         ad_copath_group = factor(ifelse(p217 > median(p217, na.rm = TRUE),
                                         "High p-tau217", "Low p-tau217"),
                                  levels = c("Low p-tau217", "High p-tau217")))

if (nrow(saa_pos) >= 2 * MIN_PER_GROUP) {
  syn_vars <- keep_usable(saa_pos %>% mutate(across(any_of(
      c("CSF_GFAP_S0120","PLASMA_GFAP_S0120","CSF_NFL_E0600",
        "PLASMA_NFLIGHT_S0100")), num_safe)),
      c("CSF_GFAP_S0120","PLASMA_GFAP_S0120","CSF_NFL_E0600","PLASMA_NFLIGHT_S0100",
        "puntuaciones","total_updrs","age_at_sample"), min_n = 2 * MIN_PER_GROUP)
  if (length(syn_vars)) {
    gt_synergy <- saa_pos %>%
      mutate(across(any_of(syn_vars), num_safe)) %>%
      select(all_of(syn_vars), ad_copath_group) %>%
      tbl_summary(by = ad_copath_group, missing = "no",
                  type = list(everything() ~ "continuous"),
                  statistic = list(everything() ~ "{median} [{p25}, {p75}]")) %>%
      add_n() %>% add_p(test = everything() ~ "wilcox.test") %>%
      add_q(method = "BH") %>% bold_p(t = 0.05) %>%
      modify_header(label ~ "**Metric**") %>% bold_labels() %>%
      as_gt() %>%
      tab_header(title = md("**SAA-positive subjects stratified by plasma p-tau217**"),
                 subtitle = sprintf("Median split, n = %d. Exploratory.", nrow(saa_pos)))
    gtsave(gt_synergy, file.path(results_path, "Synergy_SAA_Positive.html"))
    gt_synergy
  } else cat("No variable reaches the minimum n for this stratification.")
} else {
  cat(sprintf("Only %d SAA-positive subjects have plasma p-tau217 — too few to stratify (need %d).",
              nrow(saa_pos), 2 * MIN_PER_GROUP))
}
```

# UPDRS III and physical performance (SPPB)

```{r updrs-sppb}
## The original showed a scatter plot only. Correlations are quantified here,
## per group, with a Fisher r-to-z test of whether they differ.
cor_one <- function(dat, lab) {
  x <- num_safe(dat$updrs3_total_score); y <- num_safe(dat$sppb_total_score)
  ok <- !is.na(x) & !is.na(y); n <- sum(ok)
  if (n < 5) return(tibble(Group = lab, n = n, rho = NA, p = NA, Pearson_r = NA))
  sp <- suppressWarnings(cor.test(x[ok], y[ok], method = "spearman", exact = FALSE))
  pe <- suppressWarnings(cor.test(x[ok], y[ok], method = "pearson"))
  tibble(Group = lab, n = n, rho = round(unname(sp$estimate), 3),
         p = signif(sp$p.value, 3), Pearson_r = round(unname(pe$estimate), 3))
}
cor_res <- bind_rows(
  cor_one(df_both %>% filter(saa_any_pos == "Positive"), "SAA positive"),
  cor_one(df_both %>% filter(saa_any_pos == "Negative"), "SAA negative"),
  cor_one(df_both, "All (combined)"),
  cor_one(df_saa2, "All (Sant Pau)"))

a <- cor_res[1,]; b <- cor_res[2,]
fisher_line <- if (!is.na(a$rho) && !is.na(b$rho) && a$n > 3 && b$n > 3) {
  z <- (atanh(a$rho) - atanh(b$rho)) / sqrt(1/(a$n-3) + 1/(b$n-3))
  sprintf("Fisher r-to-z, SAA+ vs SAA−: z = %.2f, p = %.3f", z, 2*pnorm(-abs(z)))
} else "Too few complete pairs to compare the two correlations."

kable(cor_res, caption = paste("UPDRS III vs SPPB.", fisher_line)) %>%
  kable_styling(full_width = FALSE, bootstrap_options = c("striped","condensed"))
write.csv(cor_res, file.path(results_path, "Correlation_UPDRS_SPPB.csv"), row.names = FALSE)
```

```{r updrs-sppb-plot, fig.width=7.5, fig.height=5}
pd <- df_both %>%
  mutate(u = num_safe(updrs3_total_score), s = num_safe(sppb_total_score)) %>%
  filter(!is.na(u), !is.na(s))
if (nrow(pd) >= 5) {
  ggplot(pd, aes(u, s, colour = saa_any_pos)) +
    geom_point(alpha = .75, size = 2.6) +
    geom_smooth(method = "lm", se = TRUE, formula = y ~ x) +
    scale_colour_manual(values = c(Negative = "#E69F00", Positive = "#56B4E9")) +
    labs(title = "UPDRS III vs SPPB by SAA status",
         subtitle = sprintf("n = %d complete pairs", nrow(pd)),
         x = "MDS-UPDRS III", y = "SPPB total", colour = "SAA status") +
    theme(legend.position = "bottom")
} else {
  cat(sprintf("Only %d subjects have both UPDRS III and SPPB — no plot drawn.", nrow(pd)))
}
```

# MRI availability

```{r mri-availability}
mri_avail <- analytic_df %>%
  mutate(has_mri = !is.na(as_date_safe(mridate)) | !is.na(num_safe(mta)) |
                   !is.na(num_safe(fazekas))) %>%
  summarise(`Merged subjects` = n(),
            `With any MRI record` = sum(has_mri),
            `With MRI date`   = sum(!is.na(as_date_safe(mridate))),
            `With MTA score`  = sum(!is.na(num_safe(mta))),
            `With Fazekas`    = sum(!is.na(num_safe(fazekas))))

mri_by_grp <- df_both %>%
  mutate(has_mri = ifelse(!is.na(as_date_safe(mridate)) | !is.na(num_safe(mta)) |
                          !is.na(num_safe(fazekas)), "MRI available", "No MRI")) %>%
  count(saa_any_pos, has_mri) %>%
  pivot_wider(names_from = has_mri, values_from = n, values_fill = 0)

kable(t(mri_avail), col.names = "n", caption = "MRI availability, all merged subjects") %>%
  kable_styling(full_width = FALSE, bootstrap_options = c("striped","condensed"))
kable(mri_by_grp, caption = "MRI availability within the combined SAA cohort") %>%
  kable_styling(full_width = FALSE, bootstrap_options = c("striped","condensed"))
```

# SAA assay concordance

```{r concordance}
both_assays <- analytic_df %>%
  filter(saa_munich_pos %in% c("Negative","Positive"),
         saa_local_pos  %in% c("Negative","Positive")) %>%
  mutate(saa_munich_pos = droplevels(saa_munich_pos),
         saa_local_pos  = droplevels(saa_local_pos),
         concordant = saa_munich_pos == saa_local_pos)

ct <- table(Munich = both_assays$saa_munich_pos, SantPau = both_assays$saa_local_pos)

conc_txt <- if (sum(ct) >= 2 && all(dim(ct) == c(2,2))) {
  agree <- sum(diag(ct)) / sum(ct)
  pe <- sum(rowSums(ct) * colSums(ct)) / sum(ct)^2
  kap <- (agree - pe) / (1 - pe)
  se  <- sqrt(agree * (1 - agree) / (sum(ct) * (1 - pe)^2))
  mc  <- suppressWarnings(mcnemar.test(ct, correct = TRUE))
  sprintf(paste("n = %d subjects have both assays. Agreement %.1f%% (%d/%d).",
                "Cohen's κ = %.2f (95%% CI %.2f–%.2f). McNemar p = %.3f.",
                "Munich+/SantPau− = %d; Munich−/SantPau+ = %d."),
          sum(ct), 100*agree, sum(diag(ct)), sum(ct), kap,
          kap - 1.96*se, kap + 1.96*se, mc$p.value,
          ct["Positive","Negative"], ct["Negative","Positive"])
} else sprintf("Only %d subjects have both assays — no concordance statistics computed.", sum(ct))

as.data.frame.matrix(ct) %>% tibble::rownames_to_column("Munich (SAA01)") %>%
  gt(rowname_col = "Munich (SAA01)") %>%
  tab_header(title = md("**SAA concordance — subjects tested on both assays**"),
             subtitle = conc_txt) %>%
  tab_spanner(label = "Sant Pau (SAA02)", columns = everything()) %>%
  cols_align("center") %>% opt_stylize(style = 1, color = "blue")
```

```{r discordant-listing}
## Which subjects disagree, with the third assay and AD biomarkers to arbitrate.
disc <- both_assays %>% filter(!concordant) %>%
  transmute(NHC, visit = visit_any, sample = sample_d, gap_d = date_diff,
            Munich = saa_munich_pos, SantPau = saa_local_pos, Third = saa_third,
            AT_profile, datscan_bin, psg_rbd, rbdclinical, parkinsonism,
            hallucinations)
if (nrow(disc)) {
  write.csv(disc, file.path(results_path, "SAA_discordant_subjects.csv"), row.names = FALSE)
  kable(disc, caption = "Discordant subjects — review individually.") %>%
    kable_styling(full_width = FALSE, bootstrap_options = c("striped","condensed"),
                  font_size = 11)
} else cat("No discordant pairs.")
```

# Exploratory analyte screen

```{r analyte-screen}
## Hypothesis-generating scan across every CSF/plasma/serum assay with adequate
## coverage, rather than a hand-picked panel.
cand <- grep("^(CSF_|PLASMA_|SERUM_)", names(analytic_df), value = TRUE)
cand <- cand[!grepl("ALPHASYN|SAA[0-9]|PROTEOMICS|METABOLOMICS|SAMPLECODE|STOCK|VOLUME|PROTOCOL|SAMPLETIME|CHRONOLOGY", cand)]

screen_one <- function(cc, dat, gv) {
  v <- num_safe(dat[[cc]]); g <- dat[[gv]]
  ok <- !is.na(v) & !is.na(g); v <- v[ok]; g <- droplevels(factor(g[ok]))
  if (nlevels(g) != 2) return(NULL)
  n1 <- sum(g == "Positive"); n0 <- sum(g == "Negative")
  if (n1 < MIN_PER_GROUP || n0 < MIN_PER_GROUP) return(NULL)
  a <- v[g == "Positive"]; b <- v[g == "Negative"]
  w <- suppressWarnings(wilcox.test(a, b, exact = FALSE))
  U <- unname(w$statistic); auc <- U / (n1 * n0)
  tibble(Analyte = cc, n_pos = n1, n_neg = n0,
         median_pos = signif(median(a), 4), median_neg = signif(median(b), 4),
         fold_change = if (median(b) != 0) signif(median(a)/median(b), 3) else NA,
         rank_biserial = round(2*U/(n1*n0) - 1, 3),
         AUC = round(max(auc, 1-auc), 3),
         Direction = ifelse(auc >= .5, "higher in SAA+", "higher in SAA−"),
         p = w$p.value)
}

screen <- map_dfr(cand, screen_one, dat = df_both, gv = "saa_any_pos")
if (nrow(screen)) {
  screen <- screen %>% mutate(q_BH = p.adjust(p, "BH"),
                              p = signif(p, 4), q_BH = signif(q_BH, 4)) %>%
    arrange(p)
  write.csv(screen, file.path(results_path, "Analyte_screen.csv"), row.names = FALSE)
  kable(head(screen, 20),
        caption = sprintf("Top 20 of %d analytes tested (≥%d per group). Exploratory: no adjustment for age, sex, APOE or batch.",
                          nrow(screen), MIN_PER_GROUP)) %>%
    kable_styling(full_width = FALSE, bootstrap_options = c("striped","condensed"),
                  font_size = 11)
} else cat(sprintf("No analyte reaches %d observations per group.", MIN_PER_GROUP))
```


# Imaging

## Design


~80 bilateral grey-matter regions and as many white-matter regions against a few
dozen scans: a whole-atlas scan cannot carry the inference. A small pre-specified
family does; everything wider is exploratory. Image quality is screened and the
detectable effect size is stated up front.

Four levels, narrow to wide:

1. **Primary — nigrostriatal family (5 GM regions).** Substantia nigra, putamen,
   caudate, pallidum, accumbens. This is the a priori target for α-synuclein
   seeding pathology. BH-FDR over 5.
2. **Secondary — medial temporal family (3 GM regions).** Hippocampus, amygdala,
   parahippocampal/ambient gyrus. Tests the classic DLB prediction of *relative
   preservation* rather than atrophy. BH-FDR over 3.
3. **Anatomical composites.** Lobar and system-level means. BH-FDR over the set.
4. **Whole-atlas exploratory scan**, grey and white matter separately, BH-FDR
   within each.

Plus global tissue volumes and a left-right asymmetry analysis of the
nigrostriatal family, which bilateral means cannot show.

Model: `lm(volume_z ~ SAA + sex + age + TIV)`, volumes z-scored within the sample
so thalamus (~7 mL) and substantia nigra (~0.5 mL) share a comparable beta.


## Atlas and scan data

```{r cat12-check, results='asis'}
if (!have_cat12)
  cat("> **CAT12_Hammers_DK_SPIN.xlsx was not found in the SAA folder — ",
      "the whole imaging section is skipped.**\n")
```

```{r cat12-load, eval=have_cat12}
## The Hammers atlas file. Helpers (num_safe, norm_id, as_date_safe, read_x,
## tidy_names) are the ones defined in section 1 -- no duplicates.
cat12 <- read_x(f_cat12) %>% tidy_names()
cat12$NHC        <- norm_id(cat12$NHC)
cat12$MRI_DATE_D <- as_date_safe(cat12$MRI_DATE)
```

```{r atlas-dictionary, eval=have_cat12}
## ---------------------------------------------------------------------------
## Parse the Hammers column names into (tissue, region, hemisphere).
##   MRI_hammers_Vgm_TL.hippocampus.L  ->  gm / TL.hippocampus / L
##   MRI_hammers_Vgm_brainstem...      ->  gm / brainstem... / unpaired
## ---------------------------------------------------------------------------
parse_hammers <- function(nms) {
  keep <- grep("^MRI_hammers_V(gm|wm)_", nms, value = TRUE)
  tibble(column = keep) %>%
    mutate(tissue = ifelse(grepl("_Vgm_", column), "GM", "WM"),
           stem   = sub("^MRI_hammers_V(gm|wm)_", "", column),
           hemi   = case_when(grepl("\\.L$", stem) ~ "L",
                              grepl("\\.R$", stem) ~ "R",
                              TRUE ~ "U"),
           region = sub("\\.[LR]$", "", stem))
}
atlas <- parse_hammers(names(cat12))

## Structures whose ROI label carries no meaningful tissue volume:
## ventricles are CSF (their "Vgm" is partial-volume noise) and the corpus
## callosum is white matter (its "Vgm" likewise).
DROP_GM <- "ventricle|Third\\.ventricle|corpus\\.callosum"
DROP_WM <- "ventricle|Third\\.ventricle"

## Lobe / system assignment from the Hammers prefix codes.
classify <- function(r) {
  case_when(
    grepl("^substantia\\.nigra", r)                        ~ "Brainstem/Nigra",
    grepl("^brainstem", r)                                 ~ "Brainstem/Nigra",
    grepl("^(caudate|putamen|pallidum|nucleus\\.accumbens)", r) ~ "Basal ganglia",
    grepl("^thalamus", r)                                  ~ "Thalamus",
    grepl("^cerebellum", r)                                ~ "Cerebellum",
    grepl("^insula", r)                                    ~ "Insula",
    grepl("^CG\\.", r)                                     ~ "Cingulate",
    grepl("^TL\\.(hippocampus|amygdala|parahippocampal)", r) ~ "Medial temporal",
    grepl("^TL\\.", r)                                     ~ "Temporal",
    grepl("^FL\\.", r)                                     ~ "Frontal",
    grepl("^PL\\.", r)                                     ~ "Parietal",
    grepl("^OL\\.", r)                                     ~ "Occipital",
    TRUE                                                   ~ "Other")
}

## Human-readable labels
pretty_region <- function(r) {
  r %>% gsub("^(TL|FL|PL|OL|CG)\\.", "", .) %>%
    gsub("\\.", " ", .) %>%
    gsub("\\bremainder\\b", "rem.", .) %>%
    str_to_sentence()
}

region_dict <- atlas %>%
  filter(hemi %in% c("L", "R", "U")) %>%
  distinct(tissue, region) %>%
  mutate(system = classify(region), label = pretty_region(region))

kable(region_dict %>% filter(tissue == "GM") %>% count(system, name = "GM regions"),
      caption = "Hammers regions available per system (grey matter).") %>%
  kable_styling(full_width = FALSE, bootstrap_options = c("striped","condensed"))
```

```{r build-bilateral, eval=have_cat12}
## Bilateral mean. na.rm = FALSE deliberately: a single available hemisphere is
## NOT a bilateral mean, and silently substituting it biases the comparison.
build_volumes <- function(dat, tissue_code, drop_pat) {
  a <- atlas %>% filter(tissue == tissue_code, !grepl(drop_pat, region))
  regs <- unique(a$region)
  if (!length(regs)) return(tibble(.rows = nrow(dat)))
  out <- map_dfc(regs, function(r) {
    cc <- a$column[a$region == r]
    v <- if (length(cc) == 2)
      rowMeans(cbind(num_safe(dat[[cc[1]]]), num_safe(dat[[cc[2]]])), na.rm = FALSE)
    else num_safe(dat[[cc[1]]])
    tibble(!!paste0(tolower(tissue_code), "_", r) := v)
  })
  out
}
vol_gm <- build_volumes(cat12, "GM", DROP_GM)
vol_wm <- build_volumes(cat12, "WM", DROP_WM)

## left/right kept separately for the asymmetry analysis
lr_cols <- atlas %>% filter(tissue == "GM", hemi %in% c("L","R"),
                            !grepl(DROP_GM, region))
vol_lr <- if (nrow(lr_cols)) {
  map_dfc(seq_len(nrow(lr_cols)), function(i)
    tibble(!!paste0("lr_", lr_cols$region[i], "_", lr_cols$hemi[i]) :=
             num_safe(cat12[[lr_cols$column[i]]])))
} else {
  tibble(.rows = nrow(cat12))
}

## Coerce only the numeric globals. MRI_Protocol is the scanner/sequence batch
## label and must stay a character/factor -- an earlier version swept it up in a
## starts_with("MRI_") coercion and turned it into all-NA.
NUM_GLOBALS <- c("MRI_TIV", "MRI_GM_Tot", "MRI_WM", "MRI_CSF_Tot",
                 "MRI_WMH_Tot", "MRI_IQR")
## Asymmetry index per region, built here where the L/R pairing is explicit.
vol_ai <- {
  regs <- unique(lr_cols$region)
  if (!length(regs)) tibble(.rows = nrow(cat12)) else
    map_dfc(regs, function(r) {
      cl <- lr_cols$column[lr_cols$region == r & lr_cols$hemi == "L"]
      cr <- lr_cols$column[lr_cols$region == r & lr_cols$hemi == "R"]
      v <- if (length(cl) == 1 && length(cr) == 1) {
        L <- num_safe(cat12[[cl]]); R <- num_safe(cat12[[cr]])
        (L - R) / ((L + R) / 2)
      } else rep(NA_real_, nrow(cat12))
      tibble(!!paste0("ai_", r) := v)
    })
}

cat12_wide <- bind_cols(
  cat12 %>% select(NHC, MRI_DATE_D, any_of(c("Session", "MRI_Protocol")),
                   any_of(NUM_GLOBALS)) %>%
    mutate(across(any_of(NUM_GLOBALS), num_safe),
           batch = if ("MRI_Protocol" %in% names(.))
                     as.character(MRI_Protocol) else NA_character_),
  vol_gm, vol_wm, vol_lr, vol_ai)
```

```{r merge-imaging, eval=have_cat12}
ja <- list(x = df_both %>% select(NHC, sample_d, saa_any_pos, SEX, age_at_sample),
           y = cat12_wide, by = "NHC")
if (utils::packageVersion("dplyr") >= "1.1.0") ja$relationship <- "many-to-many"

img <- suppressWarnings(do.call(inner_join, ja)) %>%
  mutate(mri_gap = abs(as.numeric(MRI_DATE_D - sample_d))) %>%
  group_by(NHC) %>%
  arrange(ifelse(is.na(mri_gap), Inf, mri_gap), .by_group = TRUE) %>%
  slice(1) %>% ungroup()

count_pos <- function(d) sum(d$saa_any_pos == "Positive", na.rm = TRUE)
count_neg <- function(d) sum(d$saa_any_pos == "Negative", na.rm = TRUE)

steps <- tibble(Step = "Subjects with SAA status and a CAT12 scan",
                n = nrow(img), pos = count_pos(img), neg = count_neg(img))

img <- img %>% filter(is.na(mri_gap) | mri_gap <= MAX_MRI_GAP)
steps <- bind_rows(steps, tibble(
  Step = sprintf("After MRI-sample interval <= %d days", MAX_MRI_GAP),
  n = nrow(img), pos = count_pos(img), neg = count_neg(img)))

img_all <- img                       # pre-QC set, kept for the quality histogram
## CAT12 stores image quality either as a percentage (higher better, 0-100) or
## as the native rating (lower better, ~0.5-10.5). Ratings convert to the
## percentage grade with pct = 105 - 10 * rating.
iqr_raw   <- num_safe(img$MRI_IQR)
iqr_is_rating <- all(is.na(iqr_raw)) || max(iqr_raw, na.rm = TRUE) <= 11
img$MRI_IQR_pct <- if (iqr_is_rating) 105 - 10 * iqr_raw else iqr_raw
img_all$MRI_IQR_pct <- {
  r <- num_safe(img_all$MRI_IQR)
  if (iqr_is_rating) 105 - 10 * r else r
}
iqr_scale_note <- sprintf(
  "CAT12 IQR read as %s (raw range %.2f-%.2f) -> quality %.1f%%-%.1f%%.",
  if (iqr_is_rating) "the native rating, lower = better" else "a percentage, higher = better",
  suppressWarnings(min(iqr_raw, na.rm = TRUE)), suppressWarnings(max(iqr_raw, na.rm = TRUE)),
  suppressWarnings(min(img$MRI_IQR_pct, na.rm = TRUE)),
  suppressWarnings(max(img$MRI_IQR_pct, na.rm = TRUE)))

img <- img %>% filter(is.na(MRI_IQR_pct) | MRI_IQR_pct >= IQR_MIN)
steps <- bind_rows(steps, tibble(
  Step = sprintf("After CAT12 quality (rating <= %.1f, i.e. >= %d%%)",
                 IQR_MAX_RATING, IQR_MIN),
  n = nrow(img), pos = count_pos(img), neg = count_neg(img)))

img$saa_any_pos <- factor(img$saa_any_pos, levels = c("Negative", "Positive"))

## Everything downstream needs enough scans and both SAA groups present.
imaging_ok <- nrow(img) >= MIN_N_MODEL &&
              sum(img$saa_any_pos == "Positive", na.rm = TRUE) >= 3 &&
              sum(img$saa_any_pos == "Negative", na.rm = TRUE) >= 3
```

## Quality control

```{r imaging-gate, results='asis', eval=have_cat12}
if (!imaging_ok)
  cat(sprintf(paste0("> **The imaging models are skipped: %d scans remain ",
      "(%d SAA-positive, %d SAA-negative), below the minimum of %d with at ",
      "least 3 per group.** Check the quality-control table below — if the IQR ",
      "step removed everything, inspect `iqr_scale_note`.\n"),
      nrow(img), sum(img$saa_any_pos == "Positive", na.rm = TRUE),
      sum(img$saa_any_pos == "Negative", na.rm = TRUE), MIN_N_MODEL))
```

```{r qc, eval=have_cat12}
kable(steps %>% rename(`SAA positive` = pos, `SAA negative` = neg),
      caption = "Imaging sample construction") %>%
  kable_styling(full_width = FALSE, bootstrap_options = c("striped","condensed"))

cat(iqr_scale_note, "\n")
cat(sprintf("MRI-sample interval: median %s d [IQR %s-%s], max %s d.\nScan quality: median %.1f%% [range %.1f-%.1f].",
  round(median(img$mri_gap, na.rm = TRUE)),
  round(quantile(img$mri_gap, .25, na.rm = TRUE)),
  round(quantile(img$mri_gap, .75, na.rm = TRUE)),
  round(max(img$mri_gap, na.rm = TRUE)),
  median(img$MRI_IQR_pct, na.rm = TRUE), min(img$MRI_IQR_pct, na.rm = TRUE),
  max(img$MRI_IQR_pct, na.rm = TRUE)))
```

```{r qc-plots, fig.width=9, fig.height=3.2, eval=have_cat12}
p1 <- ggplot(img, aes(mri_gap)) +
  geom_histogram(bins = 25, fill = "#4C72B0", colour = "white") +
  labs(x = "|MRI date − sample date| (days)", y = "Scans", title = "Temporal offset")
p2 <- ggplot(img_all, aes(MRI_IQR_pct, fill = MRI_IQR_pct >= IQR_MIN)) +
  geom_histogram(bins = 25, colour = "white") +
  geom_vline(xintercept = IQR_MIN, linetype = "dashed") +
  scale_fill_manual(values = c("FALSE" = "#D55E00", "TRUE" = "#4C72B0"), guide = "none") +
  labs(x = "CAT12 image quality (%, converted if stored as a rating)",
       y = "Scans", title = "Scan quality")
if (requireNamespace("patchwork", quietly = TRUE)) {
  print(patchwork::wrap_plots(p1, p2, nrow = 1))
} else { print(p1); print(p2) }
```

## Scanner / protocol harmonisation (ComBat)

Scanner and sequence differences shift regional volumes independently of
biology. ComBat removes those shifts while preserving the covariates named in
its design — the SAA contrast, age, sex and TIV — so it cannot remove the effect
under test. It needs enough scans per batch to be estimable; when it is not, the
fallbacks are protocol as a covariate, or restriction to the dominant protocol.

```{r combat-setup, eval=imaging_ok}
COMBAT_MIN_BATCH <- 8     # minimum scans per protocol for ComBat
COMBAT_MAX_LEVELS <- 4    # above this, protocol-as-covariate costs too much df
DOMINANT_FRAC <- 0.80     # if one protocol holds this share and the rest are
                          # unmodellable, restrict to it rather than give up

img$batch <- if ("batch" %in% names(img)) {
  b <- trimws(as.character(img$batch)); b[b %in% c("", "NA")] <- NA; b
} else NA_character_

batch_tab <- if (all(is.na(img$batch))) NULL else
  table(Protocol = img$batch, SAA = img$saa_any_pos, useNA = "no")

n_batch     <- if (is.null(batch_tab)) 0 else nrow(batch_tab)
min_batch_n <- if (n_batch == 0) 0 else min(rowSums(batch_tab))
confounded  <- n_batch > 0 && any(apply(batch_tab, 1, function(r) any(r == 0)))

combat_pkg <- if (requireNamespace("neuroCombat", quietly = TRUE)) "neuroCombat" else
              if (requireNamespace("sva", quietly = TRUE))        "sva" else NA_character_

combat_ok  <- n_batch >= 2 && min_batch_n >= COMBAT_MIN_BATCH && !is.na(combat_pkg)
covar_ok   <- !combat_ok && n_batch >= 2 && n_batch <= COMBAT_MAX_LEVELS && min_batch_n >= 3

## When one protocol carries almost everything and the remainder are single
## scans, neither ComBat nor a covariate can work — but discarding a handful of
## scans removes the batch problem outright and costs little. That is a better
## trade than carrying an unadjustable scanner effect through every model.
dominant <- NA_character_; restrict_ok <- FALSE
if (n_batch >= 2) {
  bn <- rowSums(batch_tab)
  top <- names(bn)[which.max(bn)]
  if (max(bn) / sum(bn) >= DOMINANT_FRAC && all(batch_tab[top, ] >= 3) &&
      !combat_ok && !covar_ok) {
    dominant <- top; restrict_ok <- TRUE
  }
}

if (!is.null(batch_tab)) {
  print(kable(as.data.frame.matrix(batch_tab) %>%
                tibble::rownames_to_column("Protocol"),
        caption = "Scans per protocol by SAA status") %>%
    kable_styling(full_width = FALSE, bootstrap_options = c("striped","condensed")))
}
```

```{r combat-decide, results='asis', eval=imaging_ok}
cat(if (n_batch <= 1) {
  "**One protocol only — no harmonisation needed.**\n"
} else if (combat_ok) {
  sprintf("**ComBat applied** via `%s` across %d protocols (smallest n = %d), protecting SAA status, sex, age and TIV.\n",
          combat_pkg, n_batch, min_batch_n)
} else if (covar_ok) {
  sprintf("**ComBat not applied** — %s. Protocol is carried as a model covariate instead.\n",
          if (is.na(combat_pkg))
            "neither neuroCombat nor sva is installed (install one to enable it)"
          else sprintf("the smallest protocol has %d scans, below the %d needed",
                       min_batch_n, COMBAT_MIN_BATCH))
} else if (restrict_ok) {
  sprintf(paste0("**Restricted to the dominant protocol `%s`** (%d of %d scans). ",
    "With %d protocols and the smallest holding %d scan(s), neither ComBat nor a ",
    "protocol covariate can be estimated; dropping %d scan(s) removes the batch ",
    "effect entirely. The full sample is retained as a sensitivity check.\n"),
    dominant, sum(batch_tab[dominant, ]), sum(batch_tab), n_batch, min_batch_n,
    sum(batch_tab) - sum(batch_tab[dominant, ]))
} else {
  sprintf("**Neither ComBat nor protocol adjustment is usable** (%d protocols, smallest n = %d). Scanner effects remain in the estimates and must be stated as a limitation.\n",
          n_batch, min_batch_n)
})
if (confounded && !restrict_ok)
  cat("\n> **Warning: at least one protocol contains only one SAA group.** ",
      "Protocol and SAA status are partly confounded; no harmonisation can ",
      "separate them, and the affected estimates should be treated as unsafe.\n")
```

```{r combat-apply, eval=imaging_ok}
img_raw <- img          # pre-harmonisation, kept for the sensitivity comparison

feature_cols <- grep("^(gm|wm)_", names(img), value = TRUE)

if (combat_ok && length(feature_cols)) {
  ok_rows <- !is.na(img$batch)
  ## ComBat cannot take missing values; harmonise only complete features.
  complete_feat <- feature_cols[vapply(feature_cols,
    function(cc) all(is.finite(img[[cc]][ok_rows])), logical(1))]
  dropped_feat <- setdiff(feature_cols, complete_feat)

  ## Protected design: everything we do NOT want removed.
  prot <- c("saa_any_pos", "SEX", "age_at_sample", "MRI_TIV")
  prot <- prot[prot %in% names(img)]
  mod_df <- droplevels(as.data.frame(img[ok_rows, prot, drop = FALSE]))
  ## keep only terms that actually vary, or model.matrix hits the same
  ## contrasts error that constant covariates cause in lm()
  varies <- vapply(prot, function(cc) {
    x <- mod_df[[cc]]
    if (is.factor(x) || is.character(x)) length(unique(x[!is.na(x)])) >= 2
    else { v <- suppressWarnings(stats::var(as.numeric(x), na.rm = TRUE))
           is.finite(v) && v > 0 }
  }, logical(1))
  prot <- prot[varies]
  stopifnot("saa_any_pos" %in% prot)   # never harmonise away the contrast
  mod <- model.matrix(as.formula(paste("~", paste(prot, collapse = " + "))),
                      data = mod_df)

  dat <- t(as.matrix(img[ok_rows, complete_feat, drop = FALSE]))   # features x samples
  bt  <- factor(img$batch[ok_rows])

  harmonised <- try({
    if (combat_pkg == "neuroCombat")
      neuroCombat::neuroCombat(dat = dat, batch = bt, mod = mod)$dat.combat
    else
      sva::ComBat(dat = dat, batch = bt, mod = mod, par.prior = TRUE, prior.plots = FALSE)
  }, silent = TRUE)

  if (inherits(harmonised, "try-error")) {
    combat_ok <- FALSE
    message("ComBat failed: ", conditionMessage(attr(harmonised, "condition")))
  } else {
    hm <- t(harmonised)
    for (k in seq_along(complete_feat))
      img[[complete_feat[k]]][ok_rows] <- hm[, k]
    cat(sprintf("ComBat harmonised %d of %d regional features across %d protocols (%d features skipped for missing values).",
                length(complete_feat), length(feature_cols), nlevels(bt),
                length(dropped_feat)))
  }
}

## Restriction fallback: analyse the dominant protocol only.
if (restrict_ok) {
  img <- img %>% filter(!is.na(batch) & batch == dominant)
  imaging_ok <- nrow(img) >= MIN_N_MODEL &&
                sum(img$saa_any_pos == "Positive", na.rm = TRUE) >= 3 &&
                sum(img$saa_any_pos == "Negative", na.rm = TRUE) >= 3
  cat(sprintf("Restricted to protocol %s: %d scans (%d SAA+, %d SAA-).",
              dominant, nrow(img),
              sum(img$saa_any_pos == "Positive", na.rm = TRUE),
              sum(img$saa_any_pos == "Negative", na.rm = TRUE)))
}

## Covariate fallback: protocol as a term in every model.
EXTRA_COVARS <- if (!combat_ok && covar_ok) "batch" else character(0)
if (length(EXTRA_COVARS)) img$batch <- factor(img$batch)

BASE_COVARS <- c("SEX", "age_at_sample", "MRI_TIV", EXTRA_COVARS)
```

```{r power, eval=imaging_ok}
n1 <- sum(img$saa_any_pos == "Positive"); n0 <- sum(img$saa_any_pos == "Negative")
mde <- (qnorm(0.975) + qnorm(0.80)) * sqrt(1/n1 + 1/n0)
mde_fdr <- (qnorm(1 - 0.05/2/20) + qnorm(0.80)) * sqrt(1/n1 + 1/n0)
cat(sprintf(paste0(
  "Detectable effect (80%% power, α = 0.05, two-sided): |β| ≥ %.2f SD with n = %d ",
  "positive / %d negative.\nUnder a whole-atlas correction the bar rises to roughly ",
  "|β| ≥ %.2f SD.\nEffects smaller than this cannot be ruled in or out by this sample."),
  mde, n1, n0, mde_fdr))
```

## Global tissue volumes

```{r global, eval=imaging_ok}
GLOBALS <- c(MRI_TIV = "Total intracranial volume", MRI_GM_Tot = "Total grey matter",
             MRI_WM = "Total white matter", MRI_CSF_Tot = "Total CSF",
             MRI_WMH_Tot = "White-matter hyperintensities")

## Shared by every forest/volcano plot below, so it must be defined in a chunk
## that always runs -- not inside a conditionally evaluated plotting chunk.
sig_colour <- function(P, Q, Beta)
  ifelse(!is.na(Q) & Q < .05, ifelse(Beta > 0, "#0072B2", "#D55E00"),
         ifelse(!is.na(P) & P < .05, "#000000", "gray60"))

## Adding FDR to a result frame that came back empty must not error: map_dfr()
## returns a 0x0 tibble when every model is skipped, and that has no P column.
add_fdr <- function(res, pcol = "P") {
  if (is.null(res) || !nrow(res) || !pcol %in% names(res)) return(res)
  res$Q <- p.adjust(res[[pcol]], "BH")
  res
}

fit_one <- function(outcome, data, covars = if (exists("BASE_COVARS"))
                      BASE_COVARS else c("SEX", "age_at_sample", "MRI_TIV"),
                    zscore = TRUE, verbose = TRUE) {
  need <- c(outcome, "saa_any_pos", covars)
  need <- need[need %in% names(data)]
  if (!outcome %in% need) return(NULL)

  d <- data %>% select(all_of(unique(c(need, "saa_any_pos")))) %>%
    drop_na() %>% droplevels()
  if (nrow(d) < MIN_N_MODEL || n_distinct(d$saa_any_pos) < 2) return(NULL)

  ## Outcome must vary.
  ov <- suppressWarnings(stats::var(as.numeric(d[[outcome]]), na.rm = TRUE))
  if (!is.finite(ov) || ov == 0) return(NULL)
  if (zscore) d[[outcome]] <- as.numeric(scale(d[[outcome]]))

  ## Every outcome has its own complete-case subset. A covariate that is
  ## constant within that subset (e.g. only one sex has a WMH measurement)
  ## carries no information and makes lm() fail on the contrast, so it is
  ## dropped from this model rather than aborting the document.
  cv <- setdiff(need, c(outcome, "saa_any_pos"))
  keep <- vapply(cv, function(cc) {
    x <- d[[cc]]
    if (is.factor(x) || is.character(x) || is.logical(x))
      length(unique(x[!is.na(x)])) >= 2
    else {
      v <- suppressWarnings(stats::var(as.numeric(x), na.rm = TRUE))
      is.finite(v) && v > 0
    }
  }, logical(1))
  if (verbose && any(!keep))
    message(sprintf("fit_one(%s): dropped constant covariate(s) %s",
                    outcome, paste(cv[!keep], collapse = ", ")))
  cv <- cv[keep]

  fml <- sprintf("`%s` ~ saa_any_pos%s", outcome,
                 if (length(cv)) paste0(" + `", paste(cv, collapse = "` + `"), "`") else "")
  m <- try(lm(as.formula(fml), data = d), silent = TRUE)
  if (inherits(m, "try-error")) return(NULL)

  rn <- "saa_any_posPositive"
  sm <- summary(m)$coefficients
  if (!rn %in% rownames(sm) || is.na(sm[rn, "Estimate"])) return(NULL)
  ci <- try(confint(m)[rn, ], silent = TRUE)
  if (inherits(ci, "try-error")) ci <- c(NA_real_, NA_real_)

  tibble(n = nrow(d), n_pos = sum(d$saa_any_pos == "Positive"),
         Model = if (length(cv)) paste(cv, collapse = " + ") else "unadjusted",
         Beta = unname(sm[rn, "Estimate"]), CI_l = ci[1], CI_u = ci[2],
         P = unname(sm[rn, "Pr(>|t|)"]))
}

glob_res <- map_dfr(names(GLOBALS), function(v) {
  ## TIV is the covariate elsewhere; as an outcome it is modelled without itself
  cv <- if (v == "MRI_TIV") setdiff(BASE_COVARS, "MRI_TIV") else BASE_COVARS
  r <- fit_one(v, img, covars = cv)
  if (is.null(r)) return(NULL)
  bind_cols(tibble(Measure = unname(GLOBALS[v])), r)
}) %>% add_fdr()

if (nrow(glob_res) && "Q" %in% names(glob_res)) {
  glob_res %>% transmute(Measure, n, n_pos, `Adjusted for` = Model,
    Beta = round(Beta, 3),
    `95% CI` = sprintf("[%.2f, %.2f]", CI_l, CI_u),
    p = round(P, 4), `q (BH)` = round(Q, 3)) %>%
    kable(caption = "Global tissue volumes by SAA status (z-scored; adjusted for sex, age, TIV).") %>%
    kable_styling(full_width = FALSE, bootstrap_options = c("striped","condensed"))
} else {
  cat("No global measure could be modelled — check the imaging diagnostics above.")
}
```

## Primary analysis — nigrostriatal family

```{r primary, eval=imaging_ok}
NIGRO <- c("substantia.nigra", "putamen", "caudate.nucleus", "pallidum",
           "nucleus.accumbens")
MEDTEMP <- c("TL.hippocampus", "TL.amygdala", "TL.parahippocampal.and.ambient.gyrus")

run_family <- function(regions, tissue = "gm", data = img, label = "") {
  res <- map_dfr(regions, function(r) {
    cc <- paste0(tissue, "_", r)
    if (!cc %in% names(data)) return(NULL)
    o <- fit_one(cc, data)
    if (is.null(o)) return(NULL)
    bind_cols(tibble(Region = pretty_region(r), token = r, Family = label), o)
  })
  if (!nrow(res)) return(res)
  res %>% add_fdr() %>% arrange(P)
}

prim <- run_family(NIGRO, label = "Nigrostriatal")
sec  <- run_family(MEDTEMP, label = "Medial temporal")

show_family <- function(res, cap) {
  if (!nrow(res)) return(cat("No region in this family could be modelled.\n"))
  res %>% transmute(Region, n, n_pos, Beta = round(Beta, 3),
      `95% CI` = sprintf("[%.2f, %.2f]", CI_l, CI_u),
      Direction = ifelse(Beta > 0, "larger in SAA+", "smaller in SAA+"),
      p = round(P, 4), q = round(Q, 3)) %>%
    kable(caption = cap) %>%
    kable_styling(full_width = FALSE, bootstrap_options = c("striped","condensed")) %>%
    row_spec(1, bold = TRUE)
}
show_family(prim, "PRIMARY: nigrostriatal grey matter, SAA+ vs SAA− (BH-FDR over the family).")
```

```{r secondary, eval=imaging_ok}
show_family(sec, "SECONDARY: medial temporal grey matter. A positive β = relative preservation in SAA+.")
```

```{r primary-forest, fig.width=7.5, fig.height=4, eval=imaging_ok && nrow(prim)>0}
fam <- bind_rows(prim, sec) %>%
  mutate(col = sig_colour(P, Q, Beta),
         Region = factor(Region, levels = rev(unique(Region))))
ggplot(fam, aes(Beta, Region, colour = col)) +
  geom_vline(xintercept = 0, linetype = "dashed", colour = "grey60") +
  geom_errorbarh(aes(xmin = CI_l, xmax = CI_u), height = .22, linewidth = 1.3) +
  geom_point(size = 2.6) +
  facet_grid(Family ~ ., scales = "free_y", space = "free_y") +
  scale_colour_identity() +
  labs(title = "A priori families — regional grey matter vs SAA status",
       x = "β (SAA+ vs SAA−, SD units)", y = "",
       caption = "Blue/orange = FDR q<0.05 within family; black = nominal p<0.05; grey = ns") +
  theme(legend.position = "none")
```

## Anatomical composites

```{r composites, eval=imaging_ok}
## Mean of the z-scored member regions, so large structures do not dominate.
gm_regions <- region_dict %>% filter(tissue == "GM", !grepl(DROP_GM, region))
sys_list <- split(gm_regions$region, gm_regions$system)

comp_dat <- img
for (sn in names(sys_list)) {
  cols <- paste0("gm_", sys_list[[sn]])
  cols <- cols[cols %in% names(comp_dat)]
  if (!length(cols)) next
  z <- scale(as.matrix(comp_dat[, cols, drop = FALSE]))
  comp_dat[[paste0("comp_", make.names(sn))]] <-
    ifelse(rowSums(!is.na(z)) < ceiling(length(cols)/2), NA, rowMeans(z, na.rm = TRUE))
}

comp_res <- map_dfr(names(sys_list), function(sn) {
  cc <- paste0("comp_", make.names(sn))
  if (!cc %in% names(comp_dat)) return(NULL)
  o <- fit_one(cc, comp_dat, zscore = FALSE)
  if (is.null(o)) return(NULL)
  bind_cols(tibble(System = sn, k_regions = length(sys_list[[sn]])), o)
})
if (nrow(comp_res)) {
  comp_res <- comp_res %>% add_fdr() %>% arrange(P)
  comp_res %>% transmute(System, `regions` = k_regions, n, n_pos,
      Beta = round(Beta, 3), `95% CI` = sprintf("[%.2f, %.2f]", CI_l, CI_u),
      p = round(P, 4), `q (BH)` = round(Q, 3)) %>%
    kable(caption = "Anatomical composites (mean of z-scored member regions).") %>%
    kable_styling(full_width = FALSE, bootstrap_options = c("striped","condensed"))
  write.csv(comp_res, file.path(results_path, "Imaging_composites.csv"), row.names = FALSE)
}
```

## Whole-atlas exploratory scan

```{r scan, eval=imaging_ok}
scan_tissue <- function(tissue_code, drop_pat, label) {
  regs <- region_dict %>% filter(tissue == tissue_code, !grepl(drop_pat, region))
  pre <- tolower(tissue_code)
  res <- map_dfr(regs$region, function(r) {
    cc <- paste0(pre, "_", r)
    if (!cc %in% names(img)) return(NULL)
    o <- fit_one(cc, img)
    if (is.null(o)) return(NULL)
    bind_cols(tibble(Tissue = label, Region = pretty_region(r), token = r,
                     System = classify(r)), o)
  })
  if (!nrow(res)) return(res)
  res %>% add_fdr() %>% arrange(P)
}

scan_gm <- scan_tissue("GM", DROP_GM, "Grey matter")
scan_wm <- scan_tissue("WM", DROP_WM, "White matter")
scan_all <- bind_rows(scan_gm, scan_wm)

## A regional effect adjusted only for TIV inherits any shift in the whole
## tissue compartment: if total white matter differs between groups, every
## white-matter ROI will appear to differ too. Refitting each region with its
## own global volume as an extra covariate asks the sharper question -- does
## THIS region depart from the global pattern?
scan_vs_global <- function(tissue_code, drop_pat, global_col, label) {
  if (!global_col %in% names(img)) return(NULL)
  regs <- region_dict %>% filter(tissue == tissue_code, !grepl(drop_pat, region))
  pre <- tolower(tissue_code)
  res <- map_dfr(regs$region, function(r) {
    cc <- paste0(pre, "_", r)
    if (!cc %in% names(img)) return(NULL)
    o <- fit_one(cc, img, covars = c(BASE_COVARS, global_col))
    if (is.null(o)) return(NULL)
    bind_cols(tibble(Tissue = label, Region = pretty_region(r), token = r), o)
  })
  if (!nrow(res)) return(res)
  res %>% add_fdr() %>% arrange(P)
}
scan_gm_g <- scan_vs_global("GM", DROP_GM, "MRI_GM_Tot", "Grey matter")
scan_wm_g <- scan_vs_global("WM", DROP_WM, "MRI_WM",     "White matter")
scan_global_adj <- bind_rows(scan_gm_g, scan_wm_g)
if (nrow(scan_global_adj))
  write.csv(scan_global_adj,
            file.path(results_path, "Imaging_scan_global_adjusted.csv"),
            row.names = FALSE)
write.csv(scan_all, file.path(results_path, "Imaging_whole_atlas_scan.csv"),
          row.names = FALSE)

cat(sprintf("Regions tested: %d grey matter, %d white matter. Surviving BH-FDR within tissue: %d GM, %d WM. Nominally p<0.05: %d GM, %d WM (expected by chance: %.1f and %.1f).",
  nrow(scan_gm), nrow(scan_wm),
  sum(scan_gm$Q < .05, na.rm = TRUE), sum(scan_wm$Q < .05, na.rm = TRUE),
  sum(scan_gm$P < .05, na.rm = TRUE), sum(scan_wm$P < .05, na.rm = TRUE),
  .05 * nrow(scan_gm), .05 * nrow(scan_wm)))
```

```{r scan-table, eval=imaging_ok}
if (nrow(scan_all))
  scan_all %>% arrange(P) %>% head(20) %>%
    transmute(Tissue, Region, System, n, n_pos, Beta = round(Beta, 3),
              `95% CI` = sprintf("[%.2f, %.2f]", CI_l, CI_u),
              p = round(P, 4), `q (BH within tissue)` = round(Q, 3)) %>%
    kable(caption = "Top 20 regions across the whole atlas — exploratory, ranked by raw p.") %>%
    kable_styling(full_width = FALSE, bootstrap_options = c("striped","condensed"),
                  font_size = 11)
```

```{r scan-global-adj, eval=imaging_ok && nrow(scan_all)>0}
if (nrow(scan_global_adj)) {
  cmp_glob <- inner_join(
    scan_all        %>% select(Tissue, token, Region, B_tiv = Beta, P_tiv = P),
    scan_global_adj %>% select(Tissue, token, B_glob = Beta, P_glob = P),
    by = c("Tissue", "token")) %>% arrange(P_tiv)
  cat(sprintf("Nominal hits before vs after adjusting for the tissue total: %d -> %d.\n",
              sum(cmp_glob$P_tiv < .05), sum(cmp_glob$P_glob < .05)))
  cmp_glob %>% head(12) %>%
    transmute(Tissue, Region,
              `TIV only` = sprintf("%.2f (p=%.3f)", B_tiv, P_tiv),
              `+ tissue total` = sprintf("%.2f (p=%.3f)", B_glob, P_glob),
              Verdict = ifelse(P_tiv < .05 & P_glob >= .05, "global shift",
                        ifelse(P_glob < .05, "regional", "ns"))) %>%
    kable(caption = paste("Regional effects before and after adjusting for the",
      "tissue compartment total. 'Global shift' means the region only looked",
      "different because the whole compartment did.")) %>%
    kable_styling(full_width = FALSE, bootstrap_options = c("striped","condensed"),
                  font_size = 11)
}
```

```{r scan-volcano, fig.width=9, fig.height=4.5, eval=imaging_ok && nrow(scan_all)>0}
vp <- scan_all %>% mutate(sig = ifelse(Q < .05, "FDR q<0.05",
                               ifelse(P < .05, "nominal p<0.05", "ns")))
ggplot(vp, aes(Beta, -log10(P), colour = sig)) +
  geom_hline(yintercept = -log10(.05), linetype = "dashed", colour = "grey60") +
  geom_vline(xintercept = 0, linetype = "dotted", colour = "grey70") +
  geom_point(alpha = .85, size = 2) +
  facet_wrap(~ Tissue) +
  scale_colour_manual(values = c("FDR q<0.05" = "#D55E00",
                                 "nominal p<0.05" = "#000000", "ns" = "grey65")) +
  labs(title = "Whole-atlas scan: regional volume vs SAA status",
       subtitle = "Exploratory. Interpret against the detectable-effect bar reported above.",
       x = "β (SAA+ vs SAA−, SD units)", y = expression(-log[10](p)), colour = NULL) +
  theme(legend.position = "bottom")
```

```{r scan-by-system, fig.width=8, fig.height=4.5, eval=imaging_ok && nrow(scan_gm)>0}
## Is any anatomical system shifted as a whole? A one-sample test of the
## regional betas within each system — coherence, not significance-counting.
sys_shift <- scan_gm %>% group_by(System) %>%
  filter(n() >= 3) %>%
  summarise(k = n(), mean_beta = mean(Beta, na.rm = TRUE),
            p_shift = tryCatch(t.test(Beta)$p.value, error = function(e) NA_real_),
            .groups = "drop") %>%
  arrange(mean_beta)

ggplot(sys_shift, aes(mean_beta, reorder(System, mean_beta),
                      fill = p_shift < .05)) +
  geom_col(width = .65) +
  geom_vline(xintercept = 0, colour = "grey40") +
  scale_fill_manual(values = c("FALSE" = "grey70", "TRUE" = "#D55E00"), guide = "none") +
  labs(title = "Mean regional effect by anatomical system (grey matter)",
       subtitle = "Average beta across the regions of each system; orange = p<0.05",
       x = "Mean β (SAA+ vs SAA−, SD units)", y = "")
kable(sys_shift %>% transmute(System, regions = k, mean_beta = round(mean_beta, 3),
                              p = round(p_shift, 4)),
      caption = paste("System-level coherence of the regional effects. These are",
        "eight pre-defined anatomical systems, not a scan, so raw p is reported",
        "without multiplicity adjustment.")) %>%
  kable_styling(full_width = FALSE, bootstrap_options = c("striped","condensed"))
```

## Left–right asymmetry

```{r asymmetry, eval=imaging_ok}
## Asymmetry index = (L − R) / ((L + R)/2). Signed, so a group difference means
## the pathology is lateralised differently, not just larger.
cat(sprintf("Asymmetry columns available in the imaging sample: %d of %d nigrostriatal regions.\n",
            sum(paste0("ai_", NIGRO) %in% names(img)), length(NIGRO)))

asym_res <- map_dfr(NIGRO, function(r) {
  cc <- paste0("ai_", r)
  if (!cc %in% names(img)) return(NULL)
  d <- img %>% mutate(AI = .data[[cc]], absAI = abs(.data[[cc]]))
  o1 <- fit_one("AI",    d, covars = setdiff(BASE_COVARS, "MRI_TIV"), zscore = FALSE)
  o2 <- fit_one("absAI", d, covars = setdiff(BASE_COVARS, "MRI_TIV"), zscore = FALSE)
  if (is.null(o1)) return(NULL)
  tibble(Region = pretty_region(r), n = o1$n, n_pos = o1$n_pos,
         Beta_signed = round(o1$Beta, 4), p_signed = round(o1$P, 4),
         Beta_abs = if (is.null(o2)) NA_real_ else round(o2$Beta, 4),
         p_abs    = if (is.null(o2)) NA_real_ else round(o2$P, 4))
})
if (nrow(asym_res)) {
  asym_res <- asym_res %>% mutate(q_signed = round(p.adjust(p_signed, "BH"), 3))
  kable(asym_res, caption = paste("Asymmetry of the nigrostriatal family.",
    "Signed = direction of lateralisation (L-R); abs = degree of lateralisation",
    "regardless of side.")) %>%
    kable_styling(full_width = FALSE, bootstrap_options = c("striped","condensed"))
  write.csv(asym_res, file.path(results_path, "Imaging_asymmetry.csv"), row.names = FALSE)
} else cat("No nigrostriatal asymmetry index could be modelled.")
```

## Sensitivity analyses

```{r sensitivity, eval=imaging_ok}
## (a) proportional normalisation instead of TIV as a covariate
## (b) IQR added as a covariate
## (c) restricted to scans within one year of sampling
prop_dat <- img
for (r in NIGRO) {
  cc <- paste0("gm_", r)
  if (cc %in% names(prop_dat) && "MRI_TIV" %in% names(prop_dat))
    prop_dat[[paste0("prop_", r)]] <- prop_dat[[cc]] / prop_dat$MRI_TIV
}
iqr_dat  <- img
near_dat <- img %>% filter(is.na(mri_gap) | mri_gap <= 365)

sens <- map_dfr(NIGRO, function(r) {
  base <- fit_one(paste0("gm_", r), img)
  pre  <- if (exists("img_raw") && (isTRUE(combat_ok) || isTRUE(restrict_ok)))
            fit_one(paste0("gm_", r), img_raw) else NULL
  prop <- if (paste0("prop_", r) %in% names(prop_dat))
    fit_one(paste0("prop_", r), prop_dat,
            covars = setdiff(BASE_COVARS, "MRI_TIV")) else NULL
  iqrm <- fit_one(paste0("gm_", r), iqr_dat,
                  covars = c(BASE_COVARS, "MRI_IQR_pct"))
  near <- fit_one(paste0("gm_", r), near_dat)
  if (is.null(base)) return(NULL)
  tibble(Region = pretty_region(r),
         `All protocols` = if (is.null(pre)) "—" else sprintf("%.2f (p=%.3f)", pre$Beta, pre$P),
         `TIV covariate` = sprintf("%.2f (p=%.3f)", base$Beta, base$P),
         `Proportional`  = if (is.null(prop)) "—" else sprintf("%.2f (p=%.3f)", prop$Beta, prop$P),
         `+ IQR`         = if (is.null(iqrm)) "—" else sprintf("%.2f (p=%.3f)", iqrm$Beta, iqrm$P),
         `MRI ≤1 y (n=%d)` = if (is.null(near)) "—" else sprintf("%.2f (p=%.3f)", near$Beta, near$P))
})
if (nrow(sens)) {
  names(sens)[ncol(sens)] <- sprintf("MRI <=1 y (n=%d)", nrow(near_dat))
  kable(sens, caption = paste("Nigrostriatal family across harmonisation,",
        "normalisation and inclusion choices. Beta in SD units except the",
        "proportional column. A large gap between the all-protocols column and",
        "the main estimate means scanner effects were doing real work.")) %>%
    kable_styling(full_width = FALSE, bootstrap_options = c("striped","condensed"))
}
```


## Interpretation notes

```{r closing, results='asis', eval=imaging_ok}
top_gm <- if (nrow(scan_gm)) scan_gm[1, ] else NULL
cat(sprintf("
**Sample.** %d scans, %d SAA-positive. Detects |beta| >= %.2f SD at 80%% power;
smaller effects are out of reach, so nulls here are uninformative rather than
negative.

**Primary family.** %s

**Whole atlas.** %d grey-matter and %d white-matter regions; %d and %d survive
FDR. %s Roughly %d nominal hits are expected by chance in grey matter alone, so
single regions there are candidates, not results. The system-level plot is the
more reliable view: a real effect shifts a whole system, not one parcel.

**Harmonisation.** %s

**Excluded.** Ventricles and corpus callosum from the grey-matter analysis (their
Hammers grey-matter volume is partial-volume noise); ventricles from white matter.
",
nrow(img), n1, mde,
if (nrow(prim)) sprintf("Best region %s, beta = %.2f [%.2f, %.2f], p = %.3f, q = %.3f.",
  prim$Region[1], prim$Beta[1], prim$CI_l[1], prim$CI_u[1], prim$P[1], prim$Q[1])
  else "Could not be fitted.",
nrow(scan_gm), nrow(scan_wm),
sum(scan_gm$Q < .05, na.rm = TRUE), sum(scan_wm$Q < .05, na.rm = TRUE),
if (!is.null(top_gm)) sprintf("Strongest grey-matter signal: %s (beta = %.2f, p = %.3f).",
  top_gm$Region, top_gm$Beta, top_gm$P) else "",
round(.05 * nrow(scan_gm)),
if (n_batch <= 1) "Single protocol; none needed."
  else if (isTRUE(combat_ok)) sprintf("ComBat over %d protocols, protecting the SAA contrast, sex, age and TIV.", n_batch)
  else if (isTRUE(covar_ok)) sprintf("Protocol carried as a covariate (smallest n = %d).", min_batch_n)
  else if (isTRUE(restrict_ok)) sprintf("Restricted to protocol %s; %d scan(s) excluded.", dominant, sum(batch_tab) - sum(batch_tab[dominant, ]))
  else "Not harmonised - state as a limitation."))
```


# Summary of methodological choices

```{r summary-notes, results='asis'}
cat(sprintf("
- **Merge.** Clinical visits paired with the biofluid sample closest in time (`visit_date`, not the baseline date). %s
- **Long-format REDCap.** Subject-level fields carried down within NHC before collapsing to one row per subject, so core symptoms and visit scores coexist.
- **Reading.** `guess_max = Inf`, without which readxl types the visit-level columns from the empty enrolment rows and discards their data.
- **SAA recoding.** Anchored patterns; every raw level is printed above for verification. Indeterminate results excluded (n = %d), not folded into either group.
- **Combined cohort.** Own-site result preferred, Munich as fallback — discordant pairs are no longer silently coded Positive. This cohort contains the Sant Pau cohort, so the two tables are not independent.
- **Plausibility limits.** Values outside each scale range are set to NA and counted. In this dataset the guard finds nothing: the impossible Mayo values seen previously (mean 21, SD 62 on a 0–4 scale) came from the baseline-date row selection picking the wrong rows, and the date fix alone resolves them. The guard remains as a regression check.
- **Tests.** Rank-based for continuous variables, Fisher for categorical, with SMD alongside the mean/risk difference and BH-FDR across each biomarker panel.
- **Coverage gate.** Assays with fewer than %d observations per group are omitted rather than printed as `NA (NA)`.
- **Imaging.** Both segmentations on the same subjects, volumes z-scored so structures of different size share an axis, with an a priori 3-structure family alongside the 7-structure FDR.
", merge_note, n_indet, MIN_PER_GROUP))
```

# References

Found with Undermind and read against the results above. The note on each line
is what it says about our finding, not a summary of the paper.

**Clinical features**

1. Coughlin DG, MacLeod KR, Middleton JS, et al. *Association of CSF
   alpha-Synuclein Seeding Amplification Assay Results With Clinical Features of
   Possible and Probable Dementia With Lewy Bodies.* Neurology 2024;103(3):e209656.
   doi:10.1212/WNL.0000000000209656
   — n = 191 DLB. RBD in 86% of SAA+ vs 62% of SAA- (p < 0.0001, OR 3.63):
   our RBD result replicates with a smaller effect. Hallucinations 50% vs 47%
   (p = 0.73): our null replicates theirs. MDS-UPDRS III 33.8 vs 25.6
   (p = 0.001): **we do not replicate this** — our UPDRS III and parkinsonism
   prevalence are flat. DaT-SPECT was not analysed against SAA status, so our
   DaT result has no direct published comparator.

**Glial markers and AD co-pathology**

2. Morenas-Rodriguez E, Alcolea D, Suarez-Calvet M, et al. *Different pattern of
   CSF glial markers between dementia with Lewy bodies and Alzheimer's disease.*
   Sci Rep 2019;9:7803. doi:10.1038/s41598-019-44173-8
   — YKL-40 elevation in DLB tracked tau, not synuclein, and was unrelated to
   amyloid status. GFAP was not measured. Sets the prior that glial markers in
   DLB follow AD co-pathology.

3. Cousins KAQ, Shaw LM, Chen-Plotkin A, et al. *Plasma GFAP associates with
   secondary Alzheimer's pathology in Lewy body disease.* Ann Clin Transl Neurol
   2022;9(6):802-813. doi:10.1002/acn3.51768
   — Plasma GFAP in Lewy body disease tracks concomitant amyloid. Our CSF GFAP
   elevation in SAA+ is amyloid-independent in both strata, i.e. against this
   account — but at n = 11 vs 11.

4. Honig LS, Kang MS, Lamoureux J, Marder K. *Cerebrospinal fluid seed
   aggregation assay and Alzheimer biomarkers in a Dementia with Lewy Body
   cohort.* Alzheimers Dement 2025;21(Suppl 2):e107531. doi:10.1002/alz70856_107531
   — No difference in amyloid or tau biomarker status between SAA+ and SAA-, and
   no evidence of increased AD pathology in SAA-. Our null a-path replicates
   this, which is why the mediation question resolves as "not amyloid-mediated".

**Imaging**

5. Tosun D, Landau SM, Weiner MW, Blauwendraat C; ADNI. *Neuroimaging features of
   CSF alpha-synuclein seed amplification assay positivity in elderly individuals
   in AD continuum.* Alzheimers Dement 2024;20(Suppl 2):e092637.
   doi:10.1002/alz.092637
   — Cortical atrophy differences by SAA status, but **no medial temporal or
   hippocampal effect**, and no subcortical analysis. Supports treating our
   medial-temporal volumetric signal as noise.

6. Cohen JS, Phillips JS, Das SR, et al. *Posterior hippocampal sparing in Lewy
   body disorders with Alzheimer's copathology: an in vivo MRI study.*
   Neuroimage Clin 2024;43:103714. doi:10.1016/j.nicl.2024.103714
   — Hippocampal involvement in Lewy body disease is driven by AD co-pathology.

7. Burton EJ, Barber R, Mukaetova-Ladinska EB, et al. *Medial temporal lobe
   atrophy on MRI differentiates Alzheimer's disease from dementia with Lewy
   bodies and vascular cognitive impairment.* Brain 2009;132(1):195-203.
   doi:10.1093/brain/awn298
   — The classic MTA result, DLB vs AD rather than SAA-stratified. Our visual
   MTA finding (higher in SAA-) sits in this tradition; our CAT12 composite
   contradicts it.

**Where this leaves us.** RBD and hallucinations replicate. Parkinsonism does
not. DaT-SCAN against SAA status appears not to have been reported. The GFAP
result runs against the prevailing AD-co-pathology account and needs replication
before it is claimable. The imaging is an underpowered null and the published
work agrees there is no medial temporal effect to find.

# Reproducibility

```{r session}
sessionInfo()
```