---
title: "DLB SAA — extended CAT12 / Hammers imaging analysis"
subtitle: "Regional grey and white matter volume by CSF α-synuclein SAA status"
author: "E. Bernardi"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: true
    code_folding: hide
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.align = "center", dpi = 150)
library(readxl); library(dplyr); library(tidyr); library(purrr); library(tibble)
library(stringr); library(ggplot2); library(knitr); library(kableExtra); library(gt)
theme_set(theme_minimal(base_size = 12))
set.seed(20260826)
```

```{r config}
desktop_path <- "~/Desktop/Ricerca/Lewy_R/SAA"
results_path <- file.path(desktop_path, "Results")
f_cat12      <- file.path(desktop_path, "CAT12_Hammers_DK_SPIN.xlsx")
dir.create(results_path, recursive = TRUE, showWarnings = FALSE)

MAX_MRI_GAP <- 730    # days between biofluid sampling and MRI
IQR_MIN     <- 70     # CAT12 image-quality rating, % (below ~70 = grade D)
MIN_N_MODEL <- 15     # minimum complete cases to fit a model
```

# Design

With `r "the sample sizes reported below"`, a whole-atlas scan cannot be the
primary analysis: the Hammers atlas gives ~80 bilateral grey-matter regions and
as many white-matter regions, against a few dozen scans. The structure below
follows the same logic as the cortical-thickness work on visual hallucinations —
**a small, pre-specified family carries the inference; everything wider is
explicitly exploratory** — with two additions the previous imaging section
lacked: image-quality control, and an honest statement of what effect size this
sample can actually detect.

Four levels, from narrow to wide:

1. **Primary — nigrostriatal family (5 GM regions).** Substantia nigra, putamen,
   caudate, pallidum, accumbens. This is the a priori target for α-synuclein
   seeding pathology. BH-FDR over 5.
2. **Secondary — medial temporal family (3 GM regions).** Hippocampus, amygdala,
   parahippocampal/ambient gyrus. Tests the classic DLB prediction of *relative
   preservation* rather than atrophy. BH-FDR over 3.
3. **Anatomical composites.** Lobar and system-level means. BH-FDR over the set.
4. **Whole-atlas exploratory scan**, grey and white matter separately, BH-FDR
   within each.

Plus global tissue volumes (GM, WM, CSF, white-matter hyperintensities) and a
left–right asymmetry analysis of the nigrostriatal family, since asymmetry is
biologically meaningful in synucleinopathy and is invisible to the bilateral
means used so far.

Model throughout: `lm(volume_z ~ SAA + sex + age_at_sample + TIV)`. Volumes are
z-scored within the analysis sample so that structures spanning two orders of
magnitude (thalamus ~7 mL, substantia nigra ~0.5 mL) share a comparable β scale.

# Data

```{r load}
## ---------------------------------------------------------------------------
## The merged clinical/biofluid frame is produced by SAA_revised.Rmd, which
## saves it to Results/analytic_df.rds. Run that first.
## ---------------------------------------------------------------------------
rds <- file.path(results_path, "analytic_df.rds")
if (!file.exists(rds))
  stop("Run SAA_revised.Rmd first — it writes Results/analytic_df.rds")
bundle  <- readRDS(rds)
df_both <- bundle$df_both

num_safe <- function(x) {
  if (is.numeric(x)) return(x)
  v <- gsub("\\s", "", gsub("[<>=]", "", trimws(as.character(x))))
  v <- ifelse(grepl(",", v) & !grepl("\\.", v), gsub(",", ".", v), gsub(",", "", v))
  suppressWarnings(as.numeric(v))
}
norm_id <- function(x) {
  x <- toupper(trimws(as.character(x)))
  x[x %in% c("", "NA", "NULL", "-", "0")] <- NA
  x <- gsub("[^0-9A-Z]", "", x)
  num <- !is.na(x) & grepl("^[0-9]+$", x)
  x[num] <- sub("^0+", "", x[num]); x[!is.na(x) & x == ""] <- NA
  x
}
as_date_safe <- function(x) {
  if (inherits(x, "Date")) return(x)
  if (inherits(x, "POSIXt")) return(as.Date(x))
  if (is.numeric(x)) {
    o <- as.Date(rep(NA, length(x))); ok <- !is.na(x) & x > 10000 & x < 60000
    o[ok] <- as.Date(x[ok], origin = "1899-12-30"); return(o)
  }
  x <- trimws(as.character(x)); x[x %in% c("", "NA", "NULL", "-")] <- NA
  o <- as.Date(rep(NA, length(x)))
  ser <- !is.na(x) & grepl("^[0-9]{5}(\\.[0-9]+)?$", x)
  o[ser] <- as.Date(as.numeric(x[ser]), origin = "1899-12-30")
  for (f in c("%Y-%m-%d","%d/%m/%Y","%d-%m-%Y","%Y/%m/%d","%d.%m.%Y","%Y%m%d")) {
    todo <- !is.na(x) & is.na(o); if (!any(todo)) break
    o[todo] <- as.Date(substr(x[todo], 1, 10), format = f)
  }
  o
}

cat12 <- read_excel(f_cat12, guess_max = Inf, .name_repair = "minimal")
cat12 <- cat12[, names(cat12) != "" & !is.na(names(cat12)), drop = FALSE]
cat12 <- cat12[, !duplicated(names(cat12)), drop = FALSE]
cat12$NHC        <- norm_id(cat12$NHC)
cat12$MRI_DATE_D <- as_date_safe(cat12$MRI_DATE)
```

```{r atlas-dictionary}
## ---------------------------------------------------------------------------
## Parse the Hammers column names into (tissue, region, hemisphere).
##   MRI_hammers_Vgm_TL.hippocampus.L  ->  gm / TL.hippocampus / L
##   MRI_hammers_Vgm_brainstem...      ->  gm / brainstem... / unpaired
## ---------------------------------------------------------------------------
parse_hammers <- function(nms) {
  keep <- grep("^MRI_hammers_V(gm|wm)_", nms, value = TRUE)
  tibble(column = keep) %>%
    mutate(tissue = ifelse(grepl("_Vgm_", column), "GM", "WM"),
           stem   = sub("^MRI_hammers_V(gm|wm)_", "", column),
           hemi   = case_when(grepl("\\.L$", stem) ~ "L",
                              grepl("\\.R$", stem) ~ "R",
                              TRUE ~ "U"),
           region = sub("\\.[LR]$", "", stem))
}
atlas <- parse_hammers(names(cat12))

## Structures whose ROI label carries no meaningful tissue volume:
## ventricles are CSF (their "Vgm" is partial-volume noise) and the corpus
## callosum is white matter (its "Vgm" likewise).
DROP_GM <- "ventricle|Third\\.ventricle|corpus\\.callosum"
DROP_WM <- "ventricle|Third\\.ventricle"

## Lobe / system assignment from the Hammers prefix codes.
classify <- function(r) {
  case_when(
    grepl("^substantia\\.nigra", r)                        ~ "Brainstem/Nigra",
    grepl("^brainstem", r)                                 ~ "Brainstem/Nigra",
    grepl("^(caudate|putamen|pallidum|nucleus\\.accumbens)", r) ~ "Basal ganglia",
    grepl("^thalamus", r)                                  ~ "Thalamus",
    grepl("^cerebellum", r)                                ~ "Cerebellum",
    grepl("^insula", r)                                    ~ "Insula",
    grepl("^CG\\.", r)                                     ~ "Cingulate",
    grepl("^TL\\.(hippocampus|amygdala|parahippocampal)", r) ~ "Medial temporal",
    grepl("^TL\\.", r)                                     ~ "Temporal",
    grepl("^FL\\.", r)                                     ~ "Frontal",
    grepl("^PL\\.", r)                                     ~ "Parietal",
    grepl("^OL\\.", r)                                     ~ "Occipital",
    TRUE                                                   ~ "Other")
}

## Human-readable labels
pretty_region <- function(r) {
  r %>% gsub("^(TL|FL|PL|OL|CG)\\.", "", .) %>%
    gsub("\\.", " ", .) %>%
    gsub("\\bremainder\\b", "rem.", .) %>%
    str_to_sentence()
}

region_dict <- atlas %>%
  filter(hemi %in% c("L", "R", "U")) %>%
  distinct(tissue, region) %>%
  mutate(system = classify(region), label = pretty_region(region))

kable(region_dict %>% filter(tissue == "GM") %>% count(system, name = "GM regions"),
      caption = "Hammers regions available per system (grey matter).") %>%
  kable_styling(full_width = FALSE, bootstrap_options = c("striped","condensed"))
```

```{r build-bilateral}
## Bilateral mean. na.rm = FALSE deliberately: a single available hemisphere is
## NOT a bilateral mean, and silently substituting it biases the comparison.
build_volumes <- function(dat, tissue_code, drop_pat) {
  a <- atlas %>% filter(tissue == tissue_code, !grepl(drop_pat, region))
  regs <- unique(a$region)
  if (!length(regs)) return(tibble(.rows = nrow(dat)))
  out <- map_dfc(regs, function(r) {
    cc <- a$column[a$region == r]
    v <- if (length(cc) == 2)
      rowMeans(cbind(num_safe(dat[[cc[1]]]), num_safe(dat[[cc[2]]])), na.rm = FALSE)
    else num_safe(dat[[cc[1]]])
    tibble(!!paste0(tolower(tissue_code), "_", r) := v)
  })
  out
}
vol_gm <- build_volumes(cat12, "GM", DROP_GM)
vol_wm <- build_volumes(cat12, "WM", DROP_WM)

## left/right kept separately for the asymmetry analysis
lr_cols <- atlas %>% filter(tissue == "GM", hemi %in% c("L","R"),
                            !grepl(DROP_GM, region))
vol_lr <- if (nrow(lr_cols))
  map_dfc(seq_len(nrow(lr_cols)), function(i)
    tibble(!!paste0("lr_", lr_cols$region[i], "_", lr_cols$hemi[i]) :=
             num_safe(cat12[[lr_cols$column[i]]])))
  else tibble(.rows = nrow(cat12))

cat12_wide <- bind_cols(
  cat12 %>% select(NHC, MRI_DATE_D, any_of(c("Session","MRI_Protocol")),
                   MRI_TIV, MRI_GM_Tot, MRI_WM, MRI_CSF_Tot, MRI_WMH_Tot, MRI_IQR) %>%
    mutate(across(starts_with("MRI_"), ~ if (is.character(.x)) num_safe(.x) else .x)),
  vol_gm, vol_wm, vol_lr)
```

```{r merge-imaging}
ja <- list(x = df_both %>% select(NHC, sample_d, saa_any_pos, SEX, age_at_sample),
           y = cat12_wide, by = "NHC")
if (utils::packageVersion("dplyr") >= "1.1.0") ja$relationship <- "many-to-many"

img <- suppressWarnings(do.call(inner_join, ja)) %>%
  mutate(mri_gap = abs(as.numeric(MRI_DATE_D - sample_d))) %>%
  group_by(NHC) %>%
  arrange(ifelse(is.na(mri_gap), Inf, mri_gap), .by_group = TRUE) %>%
  slice(1) %>% ungroup()

count_pos <- function(d) sum(d$saa_any_pos == "Positive", na.rm = TRUE)
count_neg <- function(d) sum(d$saa_any_pos == "Negative", na.rm = TRUE)

steps <- tibble(Step = "Subjects with SAA status and a CAT12 scan",
                n = nrow(img), pos = count_pos(img), neg = count_neg(img))

img <- img %>% filter(is.na(mri_gap) | mri_gap <= MAX_MRI_GAP)
steps <- bind_rows(steps, tibble(
  Step = sprintf("After MRI-sample interval <= %d days", MAX_MRI_GAP),
  n = nrow(img), pos = count_pos(img), neg = count_neg(img)))

img_all <- img                       # pre-QC set, kept for the quality histogram
img <- img %>% filter(is.na(MRI_IQR) | MRI_IQR >= IQR_MIN)
steps <- bind_rows(steps, tibble(
  Step = sprintf("After CAT12 IQR >= %d%%", IQR_MIN),
  n = nrow(img), pos = count_pos(img), neg = count_neg(img)))

img$saa_any_pos <- factor(img$saa_any_pos, levels = c("Negative", "Positive"))
```

# Quality control

```{r qc}
kable(steps %>% rename(`SAA positive` = pos, `SAA negative` = neg),
      caption = "Imaging sample construction") %>%
  kable_styling(full_width = FALSE, bootstrap_options = c("striped","condensed"))

cat(sprintf("MRI–sample interval: median %s d [IQR %s–%s], max %s d.\nCAT12 IQR: median %.1f%% [range %.1f–%.1f].",
  round(median(img$mri_gap, na.rm = TRUE)),
  round(quantile(img$mri_gap, .25, na.rm = TRUE)),
  round(quantile(img$mri_gap, .75, na.rm = TRUE)),
  round(max(img$mri_gap, na.rm = TRUE)),
  median(img$MRI_IQR, na.rm = TRUE), min(img$MRI_IQR, na.rm = TRUE),
  max(img$MRI_IQR, na.rm = TRUE)))
```

```{r qc-plots, fig.width=9, fig.height=3.2}
p1 <- ggplot(img, aes(mri_gap)) +
  geom_histogram(bins = 25, fill = "#4C72B0", colour = "white") +
  labs(x = "|MRI date − sample date| (days)", y = "Scans", title = "Temporal offset")
p2 <- ggplot(img_all, aes(MRI_IQR, fill = MRI_IQR >= IQR_MIN)) +
  geom_histogram(bins = 25, colour = "white") +
  geom_vline(xintercept = IQR_MIN, linetype = "dashed") +
  scale_fill_manual(values = c("FALSE" = "#D55E00", "TRUE" = "#4C72B0"), guide = "none") +
  labs(x = "CAT12 image quality rating (%)", y = "Scans", title = "Scan quality")
if (requireNamespace("patchwork", quietly = TRUE)) {
  print(patchwork::wrap_plots(p1, p2, nrow = 1))
} else { print(p1); print(p2) }
```

```{r power}
n1 <- sum(img$saa_any_pos == "Positive"); n0 <- sum(img$saa_any_pos == "Negative")
mde <- (qnorm(0.975) + qnorm(0.80)) * sqrt(1/n1 + 1/n0)
mde_fdr <- (qnorm(1 - 0.05/2/20) + qnorm(0.80)) * sqrt(1/n1 + 1/n0)
cat(sprintf(paste0(
  "Detectable effect (80%% power, α = 0.05, two-sided): |β| ≥ %.2f SD with n = %d ",
  "positive / %d negative.\nUnder a whole-atlas correction the bar rises to roughly ",
  "|β| ≥ %.2f SD.\nEffects smaller than this cannot be ruled in or out by this sample."),
  mde, n1, n0, mde_fdr))
```

# Global tissue volumes

```{r global}
GLOBALS <- c(MRI_TIV = "Total intracranial volume", MRI_GM_Tot = "Total grey matter",
             MRI_WM = "Total white matter", MRI_CSF_Tot = "Total CSF",
             MRI_WMH_Tot = "White-matter hyperintensities")

fit_one <- function(outcome, data, covars = c("SEX", "age_at_sample", "MRI_TIV"),
                    zscore = TRUE) {
  need <- c(outcome, "saa_any_pos", covars)
  need <- need[need %in% names(data)]
  if (!outcome %in% need) return(NULL)
  d <- data %>% select(all_of(unique(c(need, "saa_any_pos")))) %>% drop_na()
  if (nrow(d) < MIN_N_MODEL || n_distinct(d$saa_any_pos) < 2) return(NULL)
  if (zscore) d[[outcome]] <- as.numeric(scale(d[[outcome]]))
  cv <- setdiff(need, c(outcome, "saa_any_pos"))
  fml <- sprintf("`%s` ~ saa_any_pos%s", outcome,
                 if (length(cv)) paste0(" + `", paste(cv, collapse = "` + `"), "`") else "")
  m <- lm(as.formula(fml), data = d)
  rn <- "saa_any_posPositive"
  if (!rn %in% rownames(summary(m)$coefficients)) return(NULL)
  s <- summary(m)$coefficients[rn, ]; ci <- confint(m)[rn, ]
  tibble(n = nrow(d), n_pos = sum(d$saa_any_pos == "Positive"),
         Beta = unname(s["Estimate"]), CI_l = ci[1], CI_u = ci[2],
         P = unname(s["Pr(>|t|)"]))
}

glob_res <- map_dfr(names(GLOBALS), function(v) {
  ## TIV is the covariate elsewhere; as an outcome it is modelled without itself
  cv <- if (v == "MRI_TIV") c("SEX", "age_at_sample") else c("SEX", "age_at_sample", "MRI_TIV")
  r <- fit_one(v, img, covars = cv)
  if (is.null(r)) return(NULL)
  bind_cols(tibble(Measure = unname(GLOBALS[v])), r)
}) %>% mutate(Q = p.adjust(P, "BH"))

if (nrow(glob_res)) {
  glob_res %>% transmute(Measure, n, n_pos, Beta = round(Beta, 3),
    `95% CI` = sprintf("[%.2f, %.2f]", CI_l, CI_u),
    p = round(P, 4), `q (BH)` = round(Q, 3)) %>%
    kable(caption = "Global tissue volumes by SAA status (z-scored; adjusted for sex, age, TIV).") %>%
    kable_styling(full_width = FALSE, bootstrap_options = c("striped","condensed"))
}
```

# Primary analysis — nigrostriatal family

```{r primary}
NIGRO <- c("substantia.nigra", "putamen", "caudate.nucleus", "pallidum",
           "nucleus.accumbens")
MEDTEMP <- c("TL.hippocampus", "TL.amygdala", "TL.parahippocampal.and.ambient.gyrus")

run_family <- function(regions, tissue = "gm", data = img, label = "") {
  res <- map_dfr(regions, function(r) {
    cc <- paste0(tissue, "_", r)
    if (!cc %in% names(data)) return(NULL)
    o <- fit_one(cc, data)
    if (is.null(o)) return(NULL)
    bind_cols(tibble(Region = pretty_region(r), token = r, Family = label), o)
  })
  if (!nrow(res)) return(res)
  res %>% mutate(Q = p.adjust(P, "BH")) %>% arrange(P)
}

prim <- run_family(NIGRO, label = "Nigrostriatal")
sec  <- run_family(MEDTEMP, label = "Medial temporal")

show_family <- function(res, cap) {
  if (!nrow(res)) return(cat("No region in this family could be modelled.\n"))
  res %>% transmute(Region, n, n_pos, Beta = round(Beta, 3),
      `95% CI` = sprintf("[%.2f, %.2f]", CI_l, CI_u),
      Direction = ifelse(Beta > 0, "larger in SAA+", "smaller in SAA+"),
      p = round(P, 4), q = round(Q, 3)) %>%
    kable(caption = cap) %>%
    kable_styling(full_width = FALSE, bootstrap_options = c("striped","condensed")) %>%
    row_spec(1, bold = TRUE)
}
show_family(prim, "PRIMARY: nigrostriatal grey matter, SAA+ vs SAA− (BH-FDR over the family).")
```

```{r secondary}
show_family(sec, "SECONDARY: medial temporal grey matter. A positive β = relative preservation in SAA+.")
```

```{r primary-forest, fig.width=7.5, fig.height=4, eval=nrow(prim)>0}
sig_colour <- function(P, Q, Beta)
  ifelse(!is.na(Q) & Q < .05, ifelse(Beta > 0, "#0072B2", "#D55E00"),
         ifelse(!is.na(P) & P < .05, "#000000", "gray60"))
fam <- bind_rows(prim, sec) %>%
  mutate(col = sig_colour(P, Q, Beta),
         Region = factor(Region, levels = rev(unique(Region))))
ggplot(fam, aes(Beta, Region, colour = col)) +
  geom_vline(xintercept = 0, linetype = "dashed", colour = "grey60") +
  geom_errorbarh(aes(xmin = CI_l, xmax = CI_u), height = .22, linewidth = 1.3) +
  geom_point(size = 2.6) +
  facet_grid(Family ~ ., scales = "free_y", space = "free_y") +
  scale_colour_identity() +
  labs(title = "A priori families — regional grey matter vs SAA status",
       x = "β (SAA+ vs SAA−, SD units)", y = "",
       caption = "Blue/orange = FDR q<0.05 within family; black = nominal p<0.05; grey = ns") +
  theme(legend.position = "none")
```

# Anatomical composites

```{r composites}
## Mean of the z-scored member regions, so large structures do not dominate.
gm_regions <- region_dict %>% filter(tissue == "GM", !grepl(DROP_GM, region))
sys_list <- split(gm_regions$region, gm_regions$system)

comp_dat <- img
for (sn in names(sys_list)) {
  cols <- paste0("gm_", sys_list[[sn]])
  cols <- cols[cols %in% names(comp_dat)]
  if (!length(cols)) next
  z <- scale(as.matrix(comp_dat[, cols, drop = FALSE]))
  comp_dat[[paste0("comp_", make.names(sn))]] <-
    ifelse(rowSums(!is.na(z)) < ceiling(length(cols)/2), NA, rowMeans(z, na.rm = TRUE))
}

comp_res <- map_dfr(names(sys_list), function(sn) {
  cc <- paste0("comp_", make.names(sn))
  if (!cc %in% names(comp_dat)) return(NULL)
  o <- fit_one(cc, comp_dat, zscore = FALSE)
  if (is.null(o)) return(NULL)
  bind_cols(tibble(System = sn, k_regions = length(sys_list[[sn]])), o)
})
if (nrow(comp_res)) {
  comp_res <- comp_res %>% mutate(Q = p.adjust(P, "BH")) %>% arrange(P)
  comp_res %>% transmute(System, `regions` = k_regions, n, n_pos,
      Beta = round(Beta, 3), `95% CI` = sprintf("[%.2f, %.2f]", CI_l, CI_u),
      p = round(P, 4), `q (BH)` = round(Q, 3)) %>%
    kable(caption = "Anatomical composites (mean of z-scored member regions).") %>%
    kable_styling(full_width = FALSE, bootstrap_options = c("striped","condensed"))
  write.csv(comp_res, file.path(results_path, "Imaging_composites.csv"), row.names = FALSE)
}
```

# Whole-atlas exploratory scan

```{r scan}
scan_tissue <- function(tissue_code, drop_pat, label) {
  regs <- region_dict %>% filter(tissue == tissue_code, !grepl(drop_pat, region))
  pre <- tolower(tissue_code)
  res <- map_dfr(regs$region, function(r) {
    cc <- paste0(pre, "_", r)
    if (!cc %in% names(img)) return(NULL)
    o <- fit_one(cc, img)
    if (is.null(o)) return(NULL)
    bind_cols(tibble(Tissue = label, Region = pretty_region(r), token = r,
                     System = classify(r)), o)
  })
  if (!nrow(res)) return(res)
  res %>% mutate(Q = p.adjust(P, "BH")) %>% arrange(P)
}

scan_gm <- scan_tissue("GM", DROP_GM, "Grey matter")
scan_wm <- scan_tissue("WM", DROP_WM, "White matter")
scan_all <- bind_rows(scan_gm, scan_wm)
write.csv(scan_all, file.path(results_path, "Imaging_whole_atlas_scan.csv"),
          row.names = FALSE)

cat(sprintf("Regions tested: %d grey matter, %d white matter. Surviving BH-FDR within tissue: %d GM, %d WM. Nominally p<0.05: %d GM, %d WM (expected by chance: %.1f and %.1f).",
  nrow(scan_gm), nrow(scan_wm),
  sum(scan_gm$Q < .05, na.rm = TRUE), sum(scan_wm$Q < .05, na.rm = TRUE),
  sum(scan_gm$P < .05, na.rm = TRUE), sum(scan_wm$P < .05, na.rm = TRUE),
  .05 * nrow(scan_gm), .05 * nrow(scan_wm)))
```

```{r scan-table}
if (nrow(scan_all))
  scan_all %>% arrange(P) %>% head(20) %>%
    transmute(Tissue, Region, System, n, n_pos, Beta = round(Beta, 3),
              `95% CI` = sprintf("[%.2f, %.2f]", CI_l, CI_u),
              p = round(P, 4), `q (BH within tissue)` = round(Q, 3)) %>%
    kable(caption = "Top 20 regions across the whole atlas — exploratory, ranked by raw p.") %>%
    kable_styling(full_width = FALSE, bootstrap_options = c("striped","condensed"),
                  font_size = 11)
```

```{r scan-volcano, fig.width=9, fig.height=4.5, eval=nrow(scan_all)>0}
vp <- scan_all %>% mutate(sig = ifelse(Q < .05, "FDR q<0.05",
                               ifelse(P < .05, "nominal p<0.05", "ns")))
ggplot(vp, aes(Beta, -log10(P), colour = sig)) +
  geom_hline(yintercept = -log10(.05), linetype = "dashed", colour = "grey60") +
  geom_vline(xintercept = 0, linetype = "dotted", colour = "grey70") +
  geom_point(alpha = .85, size = 2) +
  facet_wrap(~ Tissue) +
  scale_colour_manual(values = c("FDR q<0.05" = "#D55E00",
                                 "nominal p<0.05" = "#000000", "ns" = "grey65")) +
  labs(title = "Whole-atlas scan: regional volume vs SAA status",
       subtitle = "Exploratory. Interpret against the detectable-effect bar reported above.",
       x = "β (SAA+ vs SAA−, SD units)", y = expression(-log[10](p)), colour = NULL) +
  theme(legend.position = "bottom")
```

```{r scan-by-system, fig.width=8, fig.height=4.5, eval=nrow(scan_gm)>0}
## Is any anatomical system shifted as a whole? A one-sample test of the
## regional betas within each system — coherence, not significance-counting.
sys_shift <- scan_gm %>% group_by(System) %>%
  filter(n() >= 3) %>%
  summarise(k = n(), mean_beta = mean(Beta, na.rm = TRUE),
            p_shift = tryCatch(t.test(Beta)$p.value, error = function(e) NA_real_),
            .groups = "drop") %>%
  mutate(q = p.adjust(p_shift, "BH")) %>% arrange(mean_beta)

ggplot(sys_shift, aes(mean_beta, reorder(System, mean_beta),
                      fill = q < .05)) +
  geom_col(width = .65) +
  geom_vline(xintercept = 0, colour = "grey40") +
  scale_fill_manual(values = c("FALSE" = "grey70", "TRUE" = "#D55E00"), guide = "none") +
  labs(title = "Mean regional effect by anatomical system (grey matter)",
       subtitle = "Average β across the regions of each system; orange = system-level shift at q<0.05",
       x = "Mean β (SAA+ vs SAA−, SD units)", y = "")
kable(sys_shift %>% transmute(System, regions = k, mean_beta = round(mean_beta, 3),
                              p = round(p_shift, 4), q = round(q, 3)),
      caption = "System-level coherence of the regional effects.") %>%
  kable_styling(full_width = FALSE, bootstrap_options = c("striped","condensed"))
```

# Left–right asymmetry

```{r asymmetry}
## Asymmetry index = (L − R) / ((L + R)/2). Signed, so a group difference means
## the pathology is lateralised differently, not just larger.
asym_res <- map_dfr(NIGRO, function(r) {
  l <- paste0("lr_", r, "_L"); rr <- paste0("lr_", r, "_R")
  if (!all(c(l, rr) %in% names(img))) return(NULL)
  ai <- (img[[l]] - img[[rr]]) / ((img[[l]] + img[[rr]]) / 2)
  d <- img %>% mutate(AI = ai, absAI = abs(ai))
  o1 <- fit_one("AI",    d, covars = c("SEX", "age_at_sample"), zscore = FALSE)
  o2 <- fit_one("absAI", d, covars = c("SEX", "age_at_sample"), zscore = FALSE)
  if (is.null(o1)) return(NULL)
  tibble(Region = pretty_region(r), n = o1$n, n_pos = o1$n_pos,
         Beta_signed = round(o1$Beta, 4), p_signed = round(o1$P, 4),
         Beta_abs = if (is.null(o2)) NA else round(o2$Beta, 4),
         p_abs    = if (is.null(o2)) NA else round(o2$P, 4))
})
if (nrow(asym_res)) {
  asym_res <- asym_res %>% mutate(q_signed = round(p.adjust(p_signed, "BH"), 3))
  kable(asym_res, caption = paste("Asymmetry of the nigrostriatal family.",
    "Signed = direction of lateralisation; abs = degree of lateralisation regardless of side.")) %>%
    kable_styling(full_width = FALSE, bootstrap_options = c("striped","condensed"))
  write.csv(asym_res, file.path(results_path, "Imaging_asymmetry.csv"), row.names = FALSE)
} else cat("Left/right columns not available for the nigrostriatal family.")
```

# Sensitivity analyses

```{r sensitivity}
## (a) proportional normalisation instead of TIV as a covariate
## (b) IQR added as a covariate
## (c) restricted to scans within one year of sampling
prop_dat <- img
for (r in NIGRO) {
  cc <- paste0("gm_", r)
  if (cc %in% names(prop_dat) && "MRI_TIV" %in% names(prop_dat))
    prop_dat[[paste0("prop_", r)]] <- prop_dat[[cc]] / prop_dat$MRI_TIV
}
iqr_dat  <- img
near_dat <- img %>% filter(is.na(mri_gap) | mri_gap <= 365)

sens <- map_dfr(NIGRO, function(r) {
  base <- fit_one(paste0("gm_", r), img)
  prop <- if (paste0("prop_", r) %in% names(prop_dat))
    fit_one(paste0("prop_", r), prop_dat, covars = c("SEX", "age_at_sample")) else NULL
  iqrm <- fit_one(paste0("gm_", r), iqr_dat,
                  covars = c("SEX", "age_at_sample", "MRI_TIV", "MRI_IQR"))
  near <- fit_one(paste0("gm_", r), near_dat)
  if (is.null(base)) return(NULL)
  tibble(Region = pretty_region(r),
         `TIV covariate` = sprintf("%.2f (p=%.3f)", base$Beta, base$P),
         `Proportional`  = if (is.null(prop)) "—" else sprintf("%.2f (p=%.3f)", prop$Beta, prop$P),
         `+ IQR`         = if (is.null(iqrm)) "—" else sprintf("%.2f (p=%.3f)", iqrm$Beta, iqrm$P),
         `MRI ≤1 y (n=%d)` = if (is.null(near)) "—" else sprintf("%.2f (p=%.3f)", near$Beta, near$P))
})
if (nrow(sens)) {
  names(sens)[5] <- sprintf("MRI ≤1 y (n=%d)", nrow(near_dat))
  kable(sens, caption = "Nigrostriatal family under four normalisation / inclusion choices. β in SD units except the proportional column.") %>%
    kable_styling(full_width = FALSE, bootstrap_options = c("striped","condensed"))
}
```

# Interpretation notes

```{r closing, results='asis'}
top_gm <- if (nrow(scan_gm)) scan_gm[1, ] else NULL
cat(sprintf("
- **Sample.** %d scans enter the models (%d SAA-positive). At 80%% power this
  detects |β| ≥ %.2f SD; anything subtler is simply beyond reach here, so a null
  result is *uninformative*, not evidence of absence.
- **Primary family.** %s
- **Whole-atlas scan.** %d grey-matter and %d white-matter regions tested;
  %d and %d survive BH-FDR. %s
- **Reading the exploratory sections.** With ~%d nominal hits expected by chance
  in the grey-matter scan alone, individual regions there are hypothesis
  candidates, not findings. The system-level coherence plot is the more
  trustworthy view: a genuine effect should shift a whole anatomical system, not
  one isolated parcel.
- **Ventricles and corpus callosum** are excluded from the grey-matter analysis
  (their Hammers grey-matter volume is partial-volume noise) and the ventricles
  from the white-matter analysis.
- **Still CAT12-only.** Drop `Z_ASEG_MERGED.xlsx` into the SAA folder and the
  cross-pipeline comparison in SAA_revised.Rmd becomes available on this same
  subject set — the check the visual-hallucination work needed and did not have.
",
nrow(img), n1, mde,
if (nrow(prim)) sprintf("Best region %s, β = %.2f [%.2f, %.2f], p = %.3f, q = %.3f.",
  prim$Region[1], prim$Beta[1], prim$CI_l[1], prim$CI_u[1], prim$P[1], prim$Q[1])
  else "Could not be fitted.",
nrow(scan_gm), nrow(scan_wm),
sum(scan_gm$Q < .05, na.rm = TRUE), sum(scan_wm$Q < .05, na.rm = TRUE),
if (!is.null(top_gm)) sprintf("Strongest grey-matter signal: %s (β = %.2f, p = %.3f).",
  top_gm$Region, top_gm$Beta, top_gm$P) else "",
round(.05 * nrow(scan_gm))))
```

# Reproducibility

```{r session}
sessionInfo()
```
