---
title: "DLB Cohort Analysis — CSF α-synuclein SAA"
subtitle: "Clinical, biofluid and imaging profile by seed-amplification assay status"
author: "E. Bernardi"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: true
    code_folding: hide
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.align = "center", dpi = 150)

library(readxl);  library(dplyr);   library(tidyr);   library(lubridate)
library(stringr); library(purrr);   library(tibble)
library(gtsummary); library(gt);    library(ggplot2); library(knitr)
library(kableExtra)

theme_set(theme_minimal(base_size = 12))
set.seed(20260826)
```

```{r config}
## ---------------------------------------------------------------------------
## PATHS
## ---------------------------------------------------------------------------
desktop_path <- "~/Desktop/Ricerca/Lewy_R/SAA"
results_path <- file.path(desktop_path, "Results")
dir.create(results_path, recursive = TRUE, showWarnings = FALSE)

f_clin  <- file.path(desktop_path, "datafreeze_lewy.xlsx")
f_bio   <- file.path(desktop_path, "BIOFLUIDS_CLINICAL.xlsx")
## Optional: FreeSurfer normative Z from the visual-hallucination pipeline.
## If present, the imaging section adds a cross-pipeline check on the same
## subjects; if absent, that subsection is skipped and everything else runs.


## ---------------------------------------------------------------------------
## MERGE OPTIONS
## ---------------------------------------------------------------------------
MAX_GAP_DAYS <- 365   # visit <-> sample; pairs beyond this are flagged

## ---------------------------------------------------------------------------
## DATA-QUALITY GUARDS
## Values outside these ranges are set to NA and reported in section 2.
## Kept as a regression check: an earlier run produced Mayo = 21 (SD 62) on a
## 0-4 scale, which turned out to come from the baseline-date row selection
## rather than from sentinel codes. If this table is ever non-empty, the source
## column is wrong.
## ---------------------------------------------------------------------------
SENTINELS <- c(-9, -99, -999, 888, 999, 9999, -1)

PLAUSIBLE <- tribble(
  ~var,                    ~lo, ~hi,
  "edad",                    18, 110,
  "EDUC",                     0,  30,
  "puntuaciones",             0,  30,   # MoCA total
  "visuoespacialp",           0,   5,
  "identificacionp",          0,   3,
  "atencionp",                0,   6,
  "lenguajep",                0,   3,
  "abstraccionp",             0,   2,
  "orientacionp",             0,   6,
  "mis",                      0,  15,   # MoCA memory index
  "mayo_total_score",         0,   4,   # 4 dichotomous items
  "updrs1a_total_score",      0,  24,
  "updrs1b_total_score",      0,  28,
  "updrs2_total_score",       0,  52,
  "updrs3_total_score",       0, 132,
  "hoehn_yahr",               0,   5,
  "sppb_total_score",         0,  12,
  "sppb_subtotal1",           0,   4,
  "sppb_subtotal2",           0,   4,
  "sppb_subtotal3",           0,   4,
  "mta",                      0,   4,
  "fazekas",                  0,   3,
  "MMSE",                     0,  30,
  "GDS",                      1,   7,
  "gds_visit",                1,   7
)

```

# Data loading and merge

```{r load}
## ---------------------------------------------------------------------------
## guess_max = Inf is essential. readxl types each column from the first 1000
## rows; in a long REDCap export the visit-level columns (UPDRS, MoCA, SPPB)
## are empty on the leading enrolment rows and would be typed <logical>,
## silently discarding every value below row 1000.
## ---------------------------------------------------------------------------
read_x <- function(p) read_excel(p, guess_max = Inf, .name_repair = "minimal")

clin_raw <- read_x(f_clin)
bio_raw  <- read_x(f_bio)

## drop blank / duplicated column names produced by Excel
tidy_names <- function(df) {
  df <- df[, names(df) != "" & !is.na(names(df)), drop = FALSE]
  df[, !duplicated(names(df)), drop = FALSE]
}
clin_raw <- tidy_names(clin_raw); bio_raw <- tidy_names(bio_raw)

## ---- helpers --------------------------------------------------------------
norm_id <- function(x) {                       # 0012345 == 12345
  x <- toupper(trimws(as.character(x)))
  x[x %in% c("", "NA", "NULL", "-", "0")] <- NA
  x <- gsub("[^0-9A-Z]", "", x)
  num <- !is.na(x) & grepl("^[0-9]+$", x)
  x[num] <- sub("^0+", "", x[num])
  x[!is.na(x) & x == ""] <- NA
  x
}

as_date_safe <- function(x) {                  # Date / POSIXct / serial / text
  if (inherits(x, "Date"))   return(x)
  if (inherits(x, "POSIXt")) return(as.Date(x))
  if (is.numeric(x)) {
    out <- as.Date(rep(NA, length(x)))
    ok <- !is.na(x) & x > 10000 & x < 60000
    out[ok] <- as.Date(x[ok], origin = "1899-12-30"); return(out)
  }
  x <- trimws(as.character(x))
  x[x %in% c("", "NA", "NaN", "NULL", "-", "#N/A")] <- NA
  out <- as.Date(rep(NA, length(x)))
  ser <- !is.na(x) & grepl("^[0-9]{5}(\\.[0-9]+)?$", x)
  out[ser] <- as.Date(as.numeric(x[ser]), origin = "1899-12-30")
  for (f in c("%Y-%m-%d","%d/%m/%Y","%d-%m-%Y","%Y/%m/%d","%d.%m.%Y","%Y%m%d")) {
    todo <- !is.na(x) & is.na(out)
    if (!any(todo)) break
    out[todo] <- as.Date(substr(x[todo], 1, 10), format = f)
  }
  out[!is.na(out) & (out < as.Date("1900-01-01") | out > Sys.Date() + 365)] <- NA
  out
}

num_safe <- function(x) {                      # "<200", "1,5", " 12 "
  if (is.numeric(x)) return(x)
  v <- gsub("\\s", "", gsub("[<>=]", "", trimws(as.character(x))))
  v <- ifelse(grepl(",", v) & !grepl("\\.", v), gsub(",", ".", v), gsub(",", "", v))
  suppressWarnings(as.numeric(v))
}

has_col <- function(df, x) x %in% names(df)

## First non-missing value across a list of candidate columns, coalesced
## row-wise. The join suffixes clinical/biofluid name clashes, so each candidate
## is also tried with _clin and _bio appended.
coalesce_cols <- function(df, candidates, numeric = FALSE) {
  nms <- unlist(lapply(candidates, function(x)
    c(x, paste0(x, "_clin"), paste0(x, "_bio"))))
  nms <- nms[nms %in% names(df)]
  if (!length(nms))
    return(if (numeric) rep(NA_real_, nrow(df)) else rep(NA_character_, nrow(df)))
  out <- if (numeric) rep(NA_real_, nrow(df)) else rep(NA_character_, nrow(df))
  for (nm in nms) {
    v <- if (numeric) num_safe(df[[nm]]) else as.character(df[[nm]])
    if (!numeric) v[!is.na(v) & trimws(v) == ""] <- NA
    out <- dplyr::coalesce(out, v)
  }
  out
}
```

```{r clinical-reshape}
## ---------------------------------------------------------------------------
## The REDCap export is LONG: subject-level fields sit on the enrolment event,
## visit-level fields on repeating instances. Collapsing without carrying the
## subject-level fields down loses either the core symptoms or the scores.
## ---------------------------------------------------------------------------
nhc_col <- grep("^nhc$", names(clin_raw), ignore.case = TRUE, value = TRUE)[1]
stopifnot(!is.na(nhc_col))

clin <- clin_raw %>%
  mutate(NHC = norm_id(.data[[nhc_col]])) %>%
  filter(!is.na(NHC))

## ---------------------------------------------------------------------------
## Per-visit date.
## visit_date is filled on only a minority of rows, so dating visits by it alone
## discards most assessments. Each instrument in this export carries its own
## date (fecha_moca, fechaupdrs, fecha_sppb ...), and those are used as
## fallbacks, with the baseline date last.
##
## NOT date fields, despite the names: `fecha`, `mes`, `any`, `dia_semana`,
## `lugar`, `localidad` are MoCA orientation ITEMS scored 0/1.
## ---------------------------------------------------------------------------
VISIT_DATE_FIELDS <- c("visit_date", "fecha_moca", "fechaupdrs", "fecha_updrs_ia",
                       "fecha_sppb", "fecha_eq5d", "fechahad", "fecha_zarit",
                       "fecha_seguimiento", "fecha_um", "visit_date_base_lewy")

date_coverage <- tibble(
  field = VISIT_DATE_FIELDS,
  present = VISIT_DATE_FIELDS %in% names(clin),
  n_dated = vapply(VISIT_DATE_FIELDS, function(f)
    if (f %in% names(clin)) sum(!is.na(as_date_safe(clin[[f]]))) else 0L, integer(1)))

clin$visit_any <- {
  out <- as.Date(rep(NA, nrow(clin)))
  for (f in VISIT_DATE_FIELDS)
    if (f %in% names(clin)) out <- coalesce(out, as_date_safe(clin[[f]]))
  out
}
clin$visit_d <- as_date_safe(if (has_col(clin, "visit_date")) clin$visit_date else NA)

## Columns that belong to a VISIT (never carried across rows)
visit_pat <- paste0("^(visit_date|redcap_event_instance|neurologist|visitdiagnosis|",
  "gds_visit|medication|hoehn_yahr|diagnostico_visita|updrs|sppb|mayo|zarit|",
  "lethargic|sleep_day|disorganized|staring|fecha|puntuaciones|resultado|mis$|",
  "visuoespacial|identificacion|atencion|lenguaje|abstraccion|orientacion|",
  "at_|len_|directos|inversos|letras|serie7|repeticion|fluidez|num_palab|",
  "rostro|seda|templo|clavel|rojo|speech|facial|rigidity|tapping|handmovements|",
  "pronation|legagility|arising|gait|freezing|postural|globalspont|kinetictremor|",
  "resttremor|constancy|dyskinesia|posture|movilidad|autocuidado|actividades|",
  "dolor_|ansiedad|depresion|termometro|a_[0-9]|d_[0-9]|resumen_|seguimiento)")
visit_cols   <- grep(visit_pat, names(clin), value = TRUE)
subject_cols <- setdiff(names(clin), c(visit_cols, "NHC", "visit_d",
                                       "visit_fallback", "visit_any"))

clin <- clin %>%
  group_by(NHC) %>%
  mutate(across(all_of(subject_cols),
                ~ if (all(is.na(.x))) .x else coalesce(.x, .x[which(!is.na(.x))[1]]))) %>%
  ungroup()

clin_visits <- clin %>% filter(!is.na(visit_any)) %>%
  distinct(NHC, visit_any, .keep_all = TRUE)
clin_nodate <- clin %>% filter(!NHC %in% clin_visits$NHC) %>%
  distinct(NHC, .keep_all = TRUE)
clin_use <- bind_rows(clin_visits, clin_nodate)
```

```{r biofluid-merge}
bio <- bio_raw %>%
  mutate(NHC  = norm_id(.data[["NHC"]]),
         NHCb = if (has_col(bio_raw, "NHC2")) norm_id(.data[["NHC2"]]) else NA_character_) %>%
  mutate(NHC = coalesce(NHC, NHCb)) %>%
  filter(!is.na(NHC)) %>%
  distinct(across(everything()), .keep_all = TRUE)

bio$sample_d <- as_date_safe(bio$SAMPLEDATE)

join_args <- list(x = clin_use, y = bio, by = "NHC", suffix = c("_clin", "_bio"))
if (utils::packageVersion("dplyr") >= "1.1.0")
  join_args$relationship <- "many-to-many"

merged_df <- suppressWarnings(do.call(inner_join, join_args)) %>%
  mutate(date_diff = abs(as.numeric(sample_d - visit_any)),
         .rank = ifelse(is.na(date_diff), Inf, date_diff)) %>%
  group_by(NHC) %>%
  arrange(.rank, visit_any, sample_d, .by_group = TRUE) %>%
  slice(1) %>% ungroup() %>%
  mutate(gap_over_limit = !is.na(date_diff) & date_diff > MAX_GAP_DAYS)

merge_note <- sprintf(
  paste("Clinical: %s subjects / %s dated visits. Biofluids: %s subjects / %s samples.",
        "Merged: %s subjects. Visit-sample gap: median %s d [IQR %s-%s], %s%% within 90 d,",
        "%s pair(s) beyond %s d."),
  n_distinct(clin$NHC), nrow(clin_visits), n_distinct(bio$NHC), nrow(bio),
  nrow(merged_df),
  round(median(merged_df$date_diff, na.rm = TRUE)),
  round(quantile(merged_df$date_diff, .25, na.rm = TRUE)),
  round(quantile(merged_df$date_diff, .75, na.rm = TRUE)),
  round(100 * mean(merged_df$date_diff <= 90, na.rm = TRUE)),
  sum(merged_df$gap_over_limit, na.rm = TRUE), MAX_GAP_DAYS)
cat(merge_note)
```

```{r date-coverage}
kable(date_coverage %>%
        mutate(`% of rows` = round(100 * n_dated / nrow(clin), 1)),
      caption = paste("Which date fields are actually populated in the clinical",
        "export. visit_date alone would date only a fraction of the rows;",
        "visit_any coalesces down this list in order.")) %>%
  kable_styling(full_width = FALSE, bootstrap_options = c("striped","condensed"))
cat(sprintf("Rows with any usable date: %d of %d (%.0f%%).",
            sum(!is.na(clin$visit_any)), nrow(clin),
            100 * mean(!is.na(clin$visit_any))))
```

```{r gap-plot, fig.width=6.5, fig.height=3}
ggplot(merged_df %>% filter(!is.na(date_diff)), aes(date_diff)) +
  geom_histogram(bins = 40, fill = "#4C72B0", colour = "white") +
  geom_vline(xintercept = c(90, MAX_GAP_DAYS), linetype = c("dashed", "dotted"),
             colour = c("#D55E00", "grey40")) +
  labs(title = "Interval between clinical visit and biofluid sampling",
       x = "|visit date − sample date| (days)", y = "Subjects",
       caption = "Dashed = 90 d; dotted = analysis limit")
```

# Data quality

```{r derive}
## ---- safe recoders --------------------------------------------------------
## ANCHORED patterns. The original used str_detect(x, "pos|1"), which matches
## any string containing a "1" (e.g. "10", "0.15", a plate code).
recode_saa <- function(x) {
  v <- tolower(trimws(as.character(x)))
  out <- rep(NA_character_, length(v))
  out[grepl("^(0|neg|negative|negativo|negativa)$", v)] <- "Negative"
  out[grepl("^(1|pos|positive|positivo|positiva)$", v)] <- "Positive"
  out[grepl("^(2|3|9|ind|indet|indeterminate|inconclusive|dudoso|equivocal)$", v)] <-
    "Indeterminate"
  factor(out, levels = c("Negative", "Positive", "Indeterminate"))
}

recode_yesno <- function(x) {
  v <- tolower(trimws(as.character(x)))
  v[v %in% c("", "na", "null", "-", "9", "99", "unknown", "desconocido", "nd")] <- NA
  out <- rep(NA_character_, length(v))
  out[grepl("^(0|no|n|absent|ausente|negative|negativo|normal|false)$", v)] <- "No"
  out[grepl("^(1|2|yes|y|si|s[ií]|present|presente|positive|positivo|abnormal|anormal|patologico|true)$", v)] <- "Yes"
  factor(out, levels = c("No", "Yes"))
}

## Free-text/multi-level fields that must be collapsed rather than shown raw.
## The previous run printed DaT-SCAN and PSG with their full level sets
## (including "Scheduled", "NA", "RBD+SAHS+PLM"), producing an uninterpretable
## chi-squared across administrative categories.
recode_datscan <- function(x) {
  v <- tolower(trimws(as.character(x)))
  out <- rep(NA_character_, length(v))
  out[grepl("normal", v) & !grepl("abnormal|anormal|patolog", v)] <- "Normal"
  out[grepl("abnormal|anormal|patolog|altered|alterado", v)] <- "Abnormal"
  factor(out, levels = c("Normal", "Abnormal"))
}
recode_psg <- function(x) {                 # RBD present on PSG, yes/no
  v <- tolower(trimws(as.character(x)))
  out <- rep(NA_character_, length(v))
  out[grepl("no rbd|sin rbd", v)] <- "No"
  out[grepl("rbd", v) & !grepl("no rbd|sin rbd", v)] <- "Yes"
  out[grepl("scheduled|pending|programad|^na$|^$", v)] <- NA
  factor(out, levels = c("No", "Yes"))
}

## A/T status. The previous anchored version silently matched nothing on this
## export -- A_clean and AT_profile then vanished from every table because
## keep_usable() drops all-missing variables without comment. Substring matching
## is used instead, and the raw-to-recoded mapping is printed below so the
## result can be checked rather than assumed.
recode_status <- function(x, letter) {
  v <- gsub("\\s+", "", tolower(trimws(as.character(x))))
  v[v %in% c("", "na", "nan", "null", "nd", "unknown", "desconocido", "?")] <- NA
  out <- rep(NA_character_, length(v))
  lo <- tolower(letter)
  neg <- !is.na(v) & (grepl("neg", v) | v %in% c("0", "false", "-", paste0(lo, "-")))
  pos <- !is.na(v) & (grepl("pos", v) | v %in% c("1", "true", "+", paste0(lo, "+")))
  out[neg] <- paste0(letter, "-")
  out[pos] <- paste0(letter, "+")      # positive wins if both somehow match
  factor(out, levels = paste0(letter, c("-", "+")))
}

clamp <- function(x, lo, hi, label) {
  v <- num_safe(x)
  v[v %in% SENTINELS] <- NA
  bad <- !is.na(v) & (v < lo | v > hi)
  if (any(bad)) qc_log[[label]] <<- sum(bad)
  v[bad] <- NA
  v
}

qc_log <- list()

analytic_df <- merged_df
analytic_df$saa3_raw <- if (has_col(analytic_df, "CSF_ALPHASYN_SAA03"))
  recode_saa(analytic_df$CSF_ALPHASYN_SAA03) else factor(rep(NA, nrow(analytic_df)),
                                                         levels = c("Negative","Positive"))

## apply plausibility limits
for (i in seq_len(nrow(PLAUSIBLE))) {
  vn <- PLAUSIBLE$var[i]
  if (has_col(analytic_df, vn))
    analytic_df[[vn]] <- clamp(analytic_df[[vn]], PLAUSIBLE$lo[i],
                               PLAUSIBLE$hi[i], vn)
}

analytic_df <- analytic_df %>%
  mutate(
    SEX = factor(ifelse(grepl("^(f|female|mujer|femenino)$", tolower(trimws(SEX))),
                        "Female",
                 ifelse(grepl("^(m|male|hombre|var|masculino)", tolower(trimws(SEX))),
                        "Male", NA)), levels = c("Female", "Male")),

    saa_local_pos  = recode_saa(CSF_ALPHASYN_SAA02),
    saa_munich_pos = recode_saa(CSF_ALPHASYN_SAA01),
    saa_third      = saa3_raw,

    ## Own-site result preferred, Munich as fallback. The original rule coded
    ## ANY positive as Positive, which silently resolved every discordant pair
    ## in favour of Positive -- incompatible with the concordance analysis.
    saa_any_pos = factor(coalesce(as.character(saa_local_pos),
                                  as.character(saa_munich_pos)),
                         levels = c("Negative", "Positive", "Indeterminate")),

    A_clean = recode_status(ASTATUS, "A"),
    T_clean = recode_status(TSTATUS, "T"),
    AT_profile = factor(if_else(is.na(A_clean) | is.na(T_clean), NA_character_,
                                paste0(A_clean, " / ", T_clean)),
                        levels = c("A- / T-", "A- / T+", "A+ / T-", "A+ / T+")),

    ## APOE: genotype string -> carrier. Guarded against a dosage-coded column.
    apoe4_carrier = {
      g <- gsub("[^0-9]", "", as.character(APOE))
      factor(ifelse(is.na(g) | nchar(g) < 2, NA,
                    ifelse(grepl("4", g), "Yes", "No")), levels = c("No", "Yes"))
    },

    hallucinations = recode_yesno(hallucinations),
    fluctuations   = recode_yesno(fluctuations),
    rbdclinical    = recode_yesno(rbdclinical),
    parkinsonism   = recode_yesno(parkinsonism),
    hyposmia       = recode_yesno(hyposmia),
    dysautonomia   = recode_yesno(dysautonomia),
    repeatedfalls  = recode_yesno(repeatedfalls),
    hypersomnia    = recode_yesno(hypersomnia),
    neurolepticsens= recode_yesno(neurolepticsens),
    datscan_bin    = recode_datscan(datscan),
    psg_rbd        = recode_psg(psg),

    ## age at sampling beats a stored age from an unrelated timepoint
    age_at_sample = {
      a <- as.numeric(sample_d - as_date_safe(fnac)) / 365.25
      a[!is.na(a) & (a < 18 | a > 110)] <- NA
      coalesce(a, num_safe(edad), num_safe(AGE))
    },

    n_core = rowSums(cbind(hallucinations == "Yes", fluctuations == "Yes",
                           rbdclinical == "Yes", parkinsonism == "Yes"), na.rm = TRUE),
    n_core = ifelse(rowSums(!is.na(cbind(hallucinations, fluctuations,
                                         rbdclinical, parkinsonism))) == 0, NA, n_core),

    ## UPDRS total: NA unless all four parts are present (partial sums are not
    ## comparable across subjects)
    total_updrs = ifelse(
      is.na(updrs1a_total_score) | is.na(updrs1b_total_score) |
      is.na(updrs2_total_score)  | is.na(updrs3_total_score), NA,
      updrs1a_total_score + updrs1b_total_score +
      updrs2_total_score + updrs3_total_score)
  )
```

```{r dlb-and-combinations}
## ---------------------------------------------------------------------------
## Added on Inigo's feedback: DLB diagnostic certainty, GDS, and combined
## criteria counts.
## ---------------------------------------------------------------------------

## Diagnostic certainty. Raw levels are printed below -- check them before
## trusting the recode, the wording varies between fields.
dx_raw <- coalesce_cols(analytic_df, c("dxlewy_last", "dx_estudio_last_visit",
                                       "diagnostico_visita", "visitdiagnosis",
                                       "dxbaselinelewy"))

analytic_df$dlb_certainty <- {
  v <- tolower(trimws(dx_raw))
  out <- rep(NA_character_, length(v))
  out[grepl("probab", v)]              <- "Probable DLB"
  out[grepl("posib|possib", v)]        <- "Possible DLB"
  out[grepl("mci|dcl|prodrom", v)]     <- "MCI-LB / prodromal"
  factor(out, levels = c("MCI-LB / prodromal", "Possible DLB", "Probable DLB"))
}
analytic_df$dx_raw <- dx_raw

## GDS-FAST
analytic_df$gds_fast <- {
  g <- coalesce_cols(analytic_df, c("gds_visit", "GDS", "gds_fast_last_visit"),
                     numeric = TRUE)
  g[!is.na(g) & (g < 1 | g > 7)] <- NA
  g
}

## MIBG, needed for the biomarker count
analytic_df$mibg <- recode_yesno(coalesce_cols(analytic_df, c("mibg")))

## ---- combined categories --------------------------------------------------
yes <- function(x) !is.na(x) & x == "Yes"
abn <- function(x) !is.na(x) & x == "Abnormal"
seen <- function(...) rowSums(!is.na(cbind(...))) > 0

analytic_df <- analytic_df %>%
  mutate(
    ## RBD: clinical, PSG-confirmed, either, both
    rbd_any  = factor(ifelse(seen(rbdclinical, psg_rbd),
                 ifelse(yes(rbdclinical) | yes(psg_rbd), "Yes", "No"), NA),
                 levels = c("No", "Yes")),
    rbd_both = factor(ifelse(!is.na(rbdclinical) & !is.na(psg_rbd),
                 ifelse(yes(rbdclinical) & yes(psg_rbd), "Yes", "No"), NA),
                 levels = c("No", "Yes")),
    ## parkinsonism with dopaminergic imaging
    park_dat_any = factor(ifelse(seen(parkinsonism, datscan_bin),
                 ifelse(yes(parkinsonism) | abn(datscan_bin), "Yes", "No"), NA),
                 levels = c("No", "Yes")),
    park_dat_both = factor(ifelse(!is.na(parkinsonism) & !is.na(datscan_bin),
                 ifelse(yes(parkinsonism) & abn(datscan_bin), "Yes", "No"), NA),
                 levels = c("No", "Yes")),
    ## indicative biomarkers present (DaT, PSG, MIBG), 0-3
    n_indicative = ifelse(seen(datscan_bin, psg_rbd, mibg),
      rowSums(cbind(abn(datscan_bin), yes(psg_rbd), yes(mibg)), na.rm = TRUE), NA_real_),
    ## core clinical + indicative biomarkers, 0-7
    n_core_plus = ifelse(is.na(n_core) & is.na(n_indicative), NA_real_,
      rowSums(cbind(n_core, n_indicative), na.rm = TRUE)))

cat(sprintf("Diagnostic certainty recoded for %d of %d subjects. GDS available in %d. MIBG in %d.\n",
  sum(!is.na(analytic_df$dlb_certainty)), nrow(analytic_df),
  sum(!is.na(analytic_df$gds_fast)), sum(!is.na(analytic_df$mibg))))

dx_check <- tibble(Raw = dx_raw,
                   Recoded = as.character(analytic_df$dlb_certainty)) %>%
  filter(!is.na(Raw)) %>% count(Raw, Recoded, sort = TRUE) %>% head(25)
kable(dx_check, caption = paste("Diagnosis field: raw value to recoded certainty.",
  "Rows with Recoded = NA are not counted as possible or probable DLB.")) %>%
  kable_styling(full_width = FALSE, bootstrap_options = c("striped","condensed"))
```

```{r pd-phenotype}
## ---------------------------------------------------------------------------
## Motor phenotype (Jankovic): mean tremor / mean akinetic-rigid, from the
## MDS-UPDRS III items. Absent from the original script.
## ---------------------------------------------------------------------------
TREM <- c("posturaltremorr","posturaltremorl","kinetictremorr","kinetictremorl",
          "resttremorrue","resttremorlue","resttremorrle","resttremorlle",
          "resttremorlipjaw","constancy")
AKRI <- c("speech","facialexpression","rigidityneck","rigidityrue","rigiditylue",
          "rigidityrle","rigiditylle","tappingrue","tappinglue","handmovementsr",
          "handmovementsl","pronationsupinationr","pronationsupinationl",
          "tappingrle","tappinglle","legagilityr","legagilityl","arising","gait",
          "freezing","posturalstability","posture","globalspontaneityl")

item_matrix <- function(df, items) {
  got <- intersect(items, names(df))
  if (!length(got)) return(list(m = NULL, used = character()))
  m <- sapply(got, function(cc) {
    v <- num_safe(df[[cc]]); v[!is.na(v) & (v < 0 | v > 4)] <- NA; v
  })
  list(m = as.matrix(m), used = got)
}
tr <- item_matrix(analytic_df, TREM); ak <- item_matrix(analytic_df, AKRI)

mean_min <- function(m, k) {
  if (is.null(m)) return(rep(NA_real_, nrow(analytic_df)))
  out <- rowMeans(m, na.rm = TRUE); out[rowSums(!is.na(m)) < k] <- NA_real_; out
}
analytic_df$tremor_score  <- mean_min(tr$m, max(3, length(tr$used) %/% 2))
analytic_df$akrigid_score <- mean_min(ak$m, max(5, length(ak$used) %/% 2))
analytic_df <- analytic_df %>%
  mutate(td_ar_ratio = ifelse(!is.na(akrigid_score) & akrigid_score > 0,
                              tremor_score / akrigid_score, NA_real_),
         pd_phenotype = factor(case_when(
           is.na(td_ar_ratio)   ~ NA_character_,
           td_ar_ratio >= 1.0   ~ "Tremor-dominant",
           td_ar_ratio <= 0.8   ~ "Akinetic-rigid",
           TRUE                 ~ "Mixed"),
           levels = c("Akinetic-rigid", "Mixed", "Tremor-dominant")))

cat(sprintf("UPDRS-III items found: %d/%d tremor, %d/%d akinetic-rigid. Phenotype classifiable in %d subjects.",
            length(tr$used), length(TREM), length(ak$used), length(AKRI),
            sum(!is.na(analytic_df$pd_phenotype))))
```

```{r qc-report}
qc_tbl <- tibble(Variable = names(qc_log),
                 `Implausible values set to NA` = unlist(qc_log))
if (nrow(qc_tbl)) {
  kable(qc_tbl, caption = paste("Out-of-range values removed. Any non-zero row",
        "means the source column contains sentinel codes or the wrong field.")) %>%
    kable_styling(full_width = FALSE, bootstrap_options = c("striped","condensed"))
} else cat("No out-of-range values detected.")

saa_raw_check <- bind_rows(
  tibble(Assay = "SAA02 (Sant Pau)", Raw = as.character(merged_df$CSF_ALPHASYN_SAA02)),
  tibble(Assay = "SAA01 (Munich)",   Raw = as.character(merged_df$CSF_ALPHASYN_SAA01))) %>%
  filter(!is.na(Raw)) %>% count(Assay, Raw, sort = TRUE)
kable(saa_raw_check, caption = "Raw SAA values before recoding — verify every level maps as intended.") %>%
  kable_styling(full_width = FALSE, bootstrap_options = c("striped","condensed"))

at_raw_check <- bind_rows(
  if (has_col(merged_df, "ASTATUS"))
    tibble(Field = "ASTATUS", Raw = as.character(merged_df$ASTATUS),
           Recoded = as.character(recode_status(merged_df$ASTATUS, "A"))) else NULL,
  if (has_col(merged_df, "TSTATUS"))
    tibble(Field = "TSTATUS", Raw = as.character(merged_df$TSTATUS),
           Recoded = as.character(recode_status(merged_df$TSTATUS, "T"))) else NULL) %>%
  filter(!is.na(Raw)) %>% count(Field, Raw, Recoded, sort = TRUE)
if (nrow(at_raw_check))
  kable(at_raw_check, caption = paste("Amyloid / tau status: raw value to recoded",
    "value. Any row with Recoded = NA is being lost.")) %>%
    kable_styling(full_width = FALSE, bootstrap_options = c("striped","condensed"))
```

```{r cohorts}
## Indeterminate results are excluded from group comparisons, not silently merged.
n_indet <- sum(analytic_df$saa_any_pos == "Indeterminate", na.rm = TRUE)

df_saa2 <- analytic_df %>% filter(saa_local_pos %in% c("Negative","Positive")) %>%
  mutate(saa_local_pos = droplevels(saa_local_pos))
## Iñigo: Sant Pau only. saa_any_pos is kept so the SAA1/SAA2 concordance
## section can still use it, but no analysis is grouped by it.

saveRDS(list(analytic_df = analytic_df, df_saa2 = df_saa2),
        file.path(results_path, "analytic_df.rds"))

cat(sprintf("Sant Pau (SAA02): n = %d (%d SAA+ / %d SAA-). Indeterminate excluded: %d.",
  nrow(df_saa2), sum(df_saa2$saa_local_pos == "Positive"),
  sum(df_saa2$saa_local_pos == "Negative"), n_indet))
```

# Clinical demographics

```{r table-helpers}
## ---- effect sizes ---------------------------------------------------------
smd_value <- function(x, g) {
  g <- droplevels(factor(g)); if (nlevels(g) != 2) return(NA_real_)
  i1 <- g == levels(g)[1]; i2 <- g == levels(g)[2]
  if (is.numeric(x)) {
    a <- x[i1]; a <- a[!is.na(a)]; b <- x[i2]; b <- b[!is.na(b)]
    if (length(a) < 2 || length(b) < 2) return(NA_real_)
    den <- sqrt((var(a) + var(b)) / 2)
    return(if (!is.finite(den) || den == 0) NA_real_ else (mean(a) - mean(b)) / den)
  }
  f <- droplevels(factor(x)); k <- nlevels(f); if (k < 2) return(NA_real_)
  p1 <- as.numeric(prop.table(table(f[i1])))[-k]
  p2 <- as.numeric(prop.table(table(f[i2])))[-k]
  if (anyNA(p1) || anyNA(p2)) return(NA_real_)
  if (k == 2) {
    den <- sqrt((p1*(1-p1) + p2*(1-p2)) / 2)
    return(if (!is.finite(den) || den == 0) NA_real_ else (p1 - p2) / den)
  }
  S <- ((diag(p1, k-1) - outer(p1,p1)) + (diag(p2, k-1) - outer(p2,p2))) / 2
  inv <- try(solve(S), silent = TRUE); if (inherits(inv,"try-error")) return(NA_real_)
  as.numeric(sqrt(max(0, t(p1-p2) %*% inv %*% (p1-p2))))
}

fn_smd <- function(data, variable, by, ...) {
  v <- smd_value(data[[variable]], data[[by]])
  if (is.na(v)) NA_character_ else sprintf("%.2f", v)
}

## Mean difference (continuous) or risk difference (binary), with 95% CI.
## Supplied as a custom stat rather than via add_difference(), which would
## collide with add_p() over the p.value column.
fn_md <- function(data, variable, by, ...) {
  x <- data[[variable]]; g <- droplevels(factor(data[[by]]))
  if (nlevels(g) != 2) return(NA_character_)
  i1 <- g == levels(g)[1]; i2 <- g == levels(g)[2]
  if (is.numeric(x)) {
    a <- x[i1]; b <- x[i2]
    if (sum(!is.na(a)) < 2 || sum(!is.na(b)) < 2) return(NA_character_)
    tt <- try(t.test(a, b), silent = TRUE)
    if (inherits(tt, "try-error")) return(NA_character_)
    return(sprintf("%.2f (%.2f, %.2f)", unname(tt$estimate[1] - tt$estimate[2]),
                   tt$conf.int[1], tt$conf.int[2]))
  }
  f <- droplevels(factor(x)); if (nlevels(f) != 2) return(NA_character_)
  hi <- levels(f)[2]
  x1 <- sum(f[i1] == hi, na.rm = TRUE); n1 <- sum(!is.na(f[i1]))
  x2 <- sum(f[i2] == hi, na.rm = TRUE); n2 <- sum(!is.na(f[i2]))
  if (n1 < 2 || n2 < 2) return(NA_character_)
  pt <- try(prop.test(c(x1, x2), c(n1, n2), correct = FALSE), silent = TRUE)
  if (inherits(pt, "try-error")) return(NA_character_)
  sprintf("%.1f%% (%.1f, %.1f)", 100*(x1/n1 - x2/n2),
          100*pt$conf.int[1], 100*pt$conf.int[2])
}

## A tbl_summary module: n, median [IQR] or n(%), non-parametric p, difference, SMD.
## Continuous tests are rank-based; the previous script used Welch t-tests on
## ordinal (MTA, Fazekas) and heavily skewed (GFAP, NfL) variables.
mod_cont <- function(data, vars, labels, group_var) {
  data %>% select(all_of(vars), all_of(group_var)) %>%
    tbl_summary(by = all_of(group_var), missing = "no",
                type = list(everything() ~ "continuous"),
                label = labels,
                statistic = list(everything() ~ "{median} [{p25}, {p75}]"),
                digits = list(everything() ~ 1)) %>%
    add_n() %>%
    add_p(test = everything() ~ "wilcox.test",
          pvalue_fun = ~ style_pvalue(.x, digits = 3)) %>%
    add_stat(fns = everything() ~ fn_md) %>%
    modify_header(add_stat_1 = "**Diff. (95% CI)**") %>%
    add_stat(fns = everything() ~ fn_smd) %>%
    modify_header(add_stat_2 = "**SMD**") %>%
    bold_p(t = 0.05)
}

mod_cat <- function(data, vars, labels, group_var) {
  data %>% select(all_of(vars), all_of(group_var)) %>%
    tbl_summary(by = all_of(group_var), missing = "no", label = labels) %>%
    add_n() %>%
    add_p(test = everything() ~ "fisher.test",
          pvalue_fun = ~ style_pvalue(.x, digits = 3)) %>%
    add_stat(fns = everything() ~ fn_md) %>%
    modify_header(add_stat_1 = "**Diff. (95% CI)**") %>%
    add_stat(fns = everything() ~ fn_smd) %>%
    modify_header(add_stat_2 = "**SMD**") %>%
    bold_p(t = 0.05)
}

keep_usable <- function(data, vars, min_n = 5) {
  vars[vapply(vars, function(v) {
    if (!has_col(data, v)) return(FALSE)
    x <- data[[v]]
    sum(!is.na(x)) >= min_n && length(unique(x[!is.na(x)])) > 1
  }, logical(1))]
}

format_gt_table <- function(tbl) {
  as_gt(tbl) %>%
    tab_style(style = list(cell_fill(color = "#E0E0E0"), cell_text(weight = "bold")),
              locations = cells_row_groups())
}
```

```{r clinical-table}
LAB <- c(
  SEX="Female sex", apoe4_carrier="APOE ε4 carrier", age_at_sample="Age at sampling (y)",
  EDUC="Education (y)", puntuaciones="MoCA total", visuoespacialp="  Visuospatial/executive",
  identificacionp="  Naming", atencionp="  Attention", lenguajep="  Language",
  abstraccionp="  Abstraction", orientacionp="  Orientation", mis="  Memory index",
  total_updrs="MDS-UPDRS total (IA+IB+II+III)", updrs1a_total_score="  Part IA",
  updrs1b_total_score="  Part IB", updrs2_total_score="  Part II",
  updrs3_total_score="  Part III", hoehn_yahr="Hoehn & Yahr",
  pd_phenotype="Motor phenotype", td_ar_ratio="Tremor / akinetic-rigid ratio",
  hallucinations="Visual hallucinations", fluctuations="Cognitive fluctuations",
  rbdclinical="Clinical RBD", parkinsonism="Parkinsonism", n_core="Number of core features",
  hyposmia="Hyposmia", dysautonomia="Dysautonomia", repeatedfalls="Repeated falls",
  hypersomnia="Hypersomnia", neurolepticsens="Neuroleptic sensitivity",
  mayo_total_score="Mayo Fluctuations Scale", sppb_total_score="SPPB total",
  datscan_bin="DaT-SCAN abnormal", psg_rbd="PSG-confirmed RBD",
  A_clean="Amyloid status", AT_profile="A/T profile", mta="MTA score", fazekas="Fazekas score",
  dlb_certainty="Diagnosis", gds_fast="GDS-FAST stage", mibg="MIBG abnormal",
  rbd_any="RBD (clinical or PSG)", rbd_both="RBD (clinical and PSG)",
  park_dat_any="Parkinsonism or abnormal DaT", park_dat_both="Parkinsonism and abnormal DaT",
  n_indicative="Indicative biomarkers (0-3)", n_core_plus="Core + biomarkers (0-7)")

lab_list <- function(v) setNames(as.list(unname(LAB[v])), v) %>%
  { imap(., function(l, n) as.formula(paste0("`", n, "` ~ '", l, "'"))) } %>% unname()

build_clinical <- function(data, group_var) {
  blocks <- list(
    list(t="cat",  v=c("SEX","apoe4_carrier"),                      h="Demographics"),
    list(t="cont", v=c("age_at_sample","EDUC"),                      h="Demographics"),
    list(t="cat",  v="dlb_certainty",                                h="Diagnosis"),
    list(t="cont", v="gds_fast",                                     h="Diagnosis"),
    list(t="cont", v=c("puntuaciones","visuoespacialp","identificacionp","atencionp",
                       "lenguajep","abstraccionp","orientacionp","mis"),
                                                                     h="Cognition (MoCA)"),
    list(t="cont", v=c("total_updrs","updrs1a_total_score","updrs1b_total_score",
                       "updrs2_total_score","updrs3_total_score","hoehn_yahr"),
                                                                     h="Motor (MDS-UPDRS)"),
    list(t="cat",  v="pd_phenotype",                                 h="Motor phenotype"),
    list(t="cat",  v=c("hallucinations","fluctuations","rbdclinical","parkinsonism"),
                                                                     h="Core features"),
    list(t="cont", v="n_core",                                       h="Core features"),
    list(t="cat",  v=c("hyposmia","dysautonomia","repeatedfalls","hypersomnia",
                       "neurolepticsens"),                           h="Supportive features"),
    list(t="cont", v=c("mayo_total_score","sppb_total_score"),        h="Scales"),
    list(t="cat",  v=c("datscan_bin","psg_rbd","mibg","A_clean","AT_profile"), h="Biomarkers & imaging"),
    list(t="cat",  v=c("rbd_any","rbd_both","park_dat_any","park_dat_both"),
                                                                     h="Combined categories"),
    list(t="cont", v=c("n_indicative","n_core_plus"),                h="Combined counts"),
    list(t="cont", v=c("mta","fazekas"),                             h="Biomarkers & imaging"))

  built <- list(); heads <- character()
  for (b in blocks) {
    vv <- keep_usable(data, b$v)
    if (!length(vv)) next
    tb <- if (b$t == "cont") mod_cont(data, vv, lab_list(vv), group_var)
          else               mod_cat (data, vv, lab_list(vv), group_var)
    built <- c(built, list(tb)); heads <- c(heads, b$h)
  }
  tbl_stack(built, group_header = heads) %>%
    modify_header(label ~ "**Variable**") %>% bold_labels()
}
```

## Sant Pau cohort (SAA02)

```{r tbl-saa2}
gt_demo_saa2 <- build_clinical(df_saa2, "saa_local_pos") %>%
  format_gt_table() %>%
  tab_source_note("Continuous: median [IQR], Wilcoxon rank-sum. Categorical: n (%), Fisher exact. Difference = Negative − Positive (Welch). SMD = standardized mean difference. N = subjects with the variable.")
gtsave(gt_demo_saa2, file.path(results_path, "Clinical_SantPau.html"))
gt_demo_saa2
```

# Cognitive progression (MMSE and MoCA)

Longitudinal scores, so this uses **all** visits rather than the single row per
subject the rest of the report is built on. MoCA comes from the REDCap visits,
MMSE from the biofluids follow-ups.

```{r progression-build}
## Each instrument is dated by ITS OWN date field. Dating MoCA by visit_date
## discarded most assessments, which is why the first run found only 8 subjects
## with two MoCAs.
saa_lookup <- df_saa2 %>% select(NHC, saa_local_pos, SEX, age_at_sample)

score_history <- function(df, score_col, date_cols, label, lo, hi) {
  if (!has_col(df, score_col))
    return(tibble(NHC = character(), date = as.Date(character()),
                  score = numeric(), test = character()))
  d <- as.Date(rep(NA, nrow(df)))
  for (f in date_cols) if (f %in% names(df)) d <- coalesce(d, as_date_safe(df[[f]]))
  tibble(NHC = df$NHC, date = d, score = num_safe(df[[score_col]]), test = label) %>%
    filter(!is.na(NHC), !is.na(date), !is.na(score), score >= lo, score <= hi)
}

moca_long  <- score_history(clin, "puntuaciones",
                            c("fecha_moca", "visit_date", "visit_date_base_lewy"),
                            "MoCA", 0, 30)
updrs_long <- score_history(clin, "updrs3_total_score",
                            c("fechaupdrs", "visit_date"), "MDS-UPDRS III", 0, 132)
mmse_long  <- score_history(bio, "MMSE",
                            c("FUPDATE", "DATA_CAPTURE", "SAMPLEDATE"), "MMSE", 0, 30)

## What each source contributed, before any modelling.
src <- bind_rows(
  tibble(test = "MoCA",          rows = nrow(moca_long),  subjects = n_distinct(moca_long$NHC)),
  tibble(test = "MDS-UPDRS III", rows = nrow(updrs_long), subjects = n_distinct(updrs_long$NHC)),
  tibble(test = "MMSE",          rows = nrow(mmse_long),  subjects = n_distinct(mmse_long$NHC)))
kable(src, caption = "Dated assessments found in the whole export, before restricting to the SAA cohort.") %>%
  kable_styling(full_width = FALSE, bootstrap_options = c("striped","condensed"))

long <- bind_rows(moca_long, updrs_long, mmse_long) %>%
  distinct(NHC, test, date, .keep_all = TRUE) %>%
  inner_join(saa_lookup, by = "NHC") %>%
  group_by(NHC, test) %>%
  arrange(date, .by_group = TRUE) %>%
  mutate(t_years = as.numeric(date - min(date)) / 365.25,
         n_obs = dplyr::n(),
         span = max(t_years)) %>%
  ungroup()

avail <- long %>% group_by(test) %>%
  summarise(subjects = n_distinct(NHC),
            `with >=2 visits` = n_distinct(NHC[n_obs >= 2]),
            `SAA+ with >=2` = n_distinct(NHC[n_obs >= 2 & saa_local_pos == "Positive"]),
            `SAA- with >=2` = n_distinct(NHC[n_obs >= 2 & saa_local_pos == "Negative"]),
            `median visits` = median(n_obs),
            `median follow-up (y)` = round(median(span[n_obs >= 2]), 2),
            .groups = "drop")
kable(avail, caption = "Longitudinal data available in the Sant Pau SAA cohort.") %>%
  kable_styling(full_width = FALSE, bootstrap_options = c("striped","condensed"))
```

```{r progression-model}
## Two estimates, because neither alone is convincing at this size:
## a per-subject slope (simple, robust, uses only subjects with >=2 visits) and
## a mixed model (uses everyone, borrows strength, needs lme4).
prog_slopes <- long %>% filter(n_obs >= 2) %>%
  group_by(NHC, test, saa_local_pos) %>%
  summarise(slope = tryCatch(unname(coef(lm(score ~ t_years))[2]),
                             error = function(e) NA_real_),
            .groups = "drop") %>%
  filter(is.finite(slope))

if (nrow(prog_slopes)) {
  slope_tab <- prog_slopes %>% group_by(test, saa_local_pos) %>%
    summarise(n = dplyr::n(),
              `median slope (pts/y)` = round(median(slope), 2),
              IQR = sprintf("[%.2f, %.2f]", quantile(slope, .25), quantile(slope, .75)),
              .groups = "drop")
  slope_p <- prog_slopes %>% group_by(test) %>%
    summarise(p = tryCatch(signif(wilcox.test(slope ~ saa_local_pos,
                                              exact = FALSE)$p.value, 3),
                           error = function(e) NA_real_),
              SMD = round(smd_value(slope, saa_local_pos), 2), .groups = "drop")
  kable(left_join(slope_tab, slope_p, by = "test"),
        caption = paste("Per-subject annual change. Negative = decline.",
                        "Only subjects with at least two timepoints.")) %>%
    kable_styling(full_width = FALSE, bootstrap_options = c("striped","condensed"))
  write.csv(prog_slopes, file.path(results_path, "Progression_slopes.csv"),
            row.names = FALSE)
} else cat("No subject has two or more timepoints for either test.")
```

```{r progression-lmm}
if (!requireNamespace("lme4", quietly = TRUE)) {
  cat("lme4 is not installed, so the mixed model is skipped.",
      "install.packages('lme4') to enable it.")
} else {
  lmm_res <- map_dfr(unique(long$test), function(tt) {
    d <- long %>% filter(test == tt) %>% drop_na(score, t_years, saa_local_pos,
                                                 SEX, age_at_sample)
    if (n_distinct(d$NHC) < 10 || sum(d$n_obs >= 2) < 6) return(NULL)
    m <- try(lme4::lmer(score ~ t_years * saa_local_pos + age_at_sample + SEX +
                          (1 | NHC), data = d, REML = TRUE), silent = TRUE)
    if (inherits(m, "try-error")) return(NULL)
    sm <- summary(m)$coefficients
    rn <- grep("^t_years:saa_local_pos", rownames(sm), value = TRUE)[1]
    if (is.na(rn)) return(NULL)
    tibble(Test = tt, subjects = n_distinct(d$NHC), observations = nrow(d),
           `slope difference (pts/y)` = round(unname(sm[rn, "Estimate"]), 3),
           SE = round(unname(sm[rn, "Std. Error"]), 3),
           t = round(unname(sm[rn, "t value"]), 2))
  })
  if (nrow(lmm_res))
    kable(lmm_res, caption = paste("Mixed model, score ~ time x SAA + age + sex",
      "+ (1|subject). The term shown is the difference in slope, SAA+ minus",
      "SAA-. |t| > 2 is roughly p < 0.05.")) %>%
      kable_styling(full_width = FALSE, bootstrap_options = c("striped","condensed"))
  else cat("Not enough longitudinal data to fit the mixed model.")
}
```

```{r progression-plot, fig.width=9, fig.height=4}
pd <- long %>% filter(n_obs >= 2)
if (nrow(pd) >= 10) {
  ggplot(pd, aes(t_years, score, group = NHC, colour = saa_local_pos)) +
    geom_line(alpha = .45) + geom_point(size = 1.2, alpha = .7) +
    geom_smooth(aes(group = saa_local_pos), method = "lm", formula = y ~ x,
                se = TRUE, linewidth = 1.2) +
    facet_wrap(~ test, scales = "free_y") +
    scale_colour_manual(values = c(Negative = "#E69F00", Positive = "#56B4E9")) +
    labs(x = "Years from first assessment", y = "Score", colour = "SAA",
         title = "Cognitive trajectories by SAA status") +
    theme(legend.position = "bottom")
} else cat("Too few repeated assessments to plot.")
```

# UPDRS III and physical performance (SPPB)

```{r updrs-sppb}
## The original showed a scatter plot only. Correlations are quantified here,
## per group, with a Fisher r-to-z test of whether they differ.
cor_one <- function(dat, lab) {
  x <- num_safe(dat$updrs3_total_score); y <- num_safe(dat$sppb_total_score)
  ok <- !is.na(x) & !is.na(y); n <- sum(ok)
  if (n < 5) return(tibble(Group = lab, n = n, rho = NA, p = NA, Pearson_r = NA))
  sp <- suppressWarnings(cor.test(x[ok], y[ok], method = "spearman", exact = FALSE))
  pe <- suppressWarnings(cor.test(x[ok], y[ok], method = "pearson"))
  tibble(Group = lab, n = n, rho = round(unname(sp$estimate), 3),
         p = signif(sp$p.value, 3), Pearson_r = round(unname(pe$estimate), 3))
}
cor_res <- bind_rows(
  cor_one(df_saa2 %>% filter(saa_local_pos == "Positive"), "SAA positive"),
  cor_one(df_saa2 %>% filter(saa_local_pos == "Negative"), "SAA negative"),
  cor_one(df_saa2, "All (combined)"),
  cor_one(df_saa2, "All (Sant Pau)"))

a <- cor_res[1,]; b <- cor_res[2,]
fisher_line <- if (!is.na(a$rho) && !is.na(b$rho) && a$n > 3 && b$n > 3) {
  z <- (atanh(a$rho) - atanh(b$rho)) / sqrt(1/(a$n-3) + 1/(b$n-3))
  sprintf("Fisher r-to-z, SAA+ vs SAA−: z = %.2f, p = %.3f", z, 2*pnorm(-abs(z)))
} else "Too few complete pairs to compare the two correlations."

kable(cor_res, caption = paste("UPDRS III vs SPPB.", fisher_line)) %>%
  kable_styling(full_width = FALSE, bootstrap_options = c("striped","condensed"))
write.csv(cor_res, file.path(results_path, "Correlation_UPDRS_SPPB.csv"), row.names = FALSE)
```

```{r updrs-sppb-plot, fig.width=7.5, fig.height=5}
pd <- df_saa2 %>%
  mutate(u = num_safe(updrs3_total_score), s = num_safe(sppb_total_score)) %>%
  filter(!is.na(u), !is.na(s))
if (nrow(pd) >= 5) {
  ggplot(pd, aes(u, s, colour = saa_local_pos)) +
    geom_point(alpha = .75, size = 2.6) +
    geom_smooth(method = "lm", se = TRUE, formula = y ~ x) +
    scale_colour_manual(values = c(Negative = "#E69F00", Positive = "#56B4E9")) +
    labs(title = "UPDRS III vs SPPB by SAA status",
         subtitle = sprintf("n = %d complete pairs", nrow(pd)),
         x = "MDS-UPDRS III", y = "SPPB total", colour = "SAA status") +
    theme(legend.position = "bottom")
} else {
  cat(sprintf("Only %d subjects have both UPDRS III and SPPB — no plot drawn.", nrow(pd)))
}
```

# MRI availability

```{r mri-availability}
mri_avail <- analytic_df %>%
  mutate(has_mri = !is.na(as_date_safe(mridate)) | !is.na(num_safe(mta)) |
                   !is.na(num_safe(fazekas))) %>%
  summarise(`Merged subjects` = n(),
            `With any MRI record` = sum(has_mri),
            `With MRI date`   = sum(!is.na(as_date_safe(mridate))),
            `With MTA score`  = sum(!is.na(num_safe(mta))),
            `With Fazekas`    = sum(!is.na(num_safe(fazekas))))

mri_by_grp <- df_saa2 %>%
  mutate(has_mri = ifelse(!is.na(as_date_safe(mridate)) | !is.na(num_safe(mta)) |
                          !is.na(num_safe(fazekas)), "MRI available", "No MRI")) %>%
  count(saa_local_pos, has_mri) %>%
  pivot_wider(names_from = has_mri, values_from = n, values_fill = 0)

kable(t(mri_avail), col.names = "n", caption = "MRI availability, all merged subjects") %>%
  kable_styling(full_width = FALSE, bootstrap_options = c("striped","condensed"))
kable(mri_by_grp, caption = "MRI availability within the combined SAA cohort") %>%
  kable_styling(full_width = FALSE, bootstrap_options = c("striped","condensed"))
```

# SAA assay concordance

```{r concordance}
both_assays <- analytic_df %>%
  filter(saa_munich_pos %in% c("Negative","Positive"),
         saa_local_pos  %in% c("Negative","Positive")) %>%
  mutate(saa_munich_pos = droplevels(saa_munich_pos),
         saa_local_pos  = droplevels(saa_local_pos),
         concordant = saa_munich_pos == saa_local_pos)

ct <- table(Munich = both_assays$saa_munich_pos, SantPau = both_assays$saa_local_pos)

conc_txt <- if (sum(ct) >= 2 && all(dim(ct) == c(2,2))) {
  agree <- sum(diag(ct)) / sum(ct)
  pe <- sum(rowSums(ct) * colSums(ct)) / sum(ct)^2
  kap <- (agree - pe) / (1 - pe)
  se  <- sqrt(agree * (1 - agree) / (sum(ct) * (1 - pe)^2))
  mc  <- suppressWarnings(mcnemar.test(ct, correct = TRUE))
  sprintf(paste("n = %d subjects have both assays. Agreement %.1f%% (%d/%d).",
                "Cohen's κ = %.2f (95%% CI %.2f–%.2f). McNemar p = %.3f.",
                "Munich+/SantPau− = %d; Munich−/SantPau+ = %d."),
          sum(ct), 100*agree, sum(diag(ct)), sum(ct), kap,
          kap - 1.96*se, kap + 1.96*se, mc$p.value,
          ct["Positive","Negative"], ct["Negative","Positive"])
} else sprintf("Only %d subjects have both assays — no concordance statistics computed.", sum(ct))

as.data.frame.matrix(ct) %>% tibble::rownames_to_column("Munich (SAA01)") %>%
  gt(rowname_col = "Munich (SAA01)") %>%
  tab_header(title = md("**SAA concordance — subjects tested on both assays**"),
             subtitle = conc_txt) %>%
  tab_spanner(label = "Sant Pau (SAA02)", columns = everything()) %>%
  cols_align("center") %>% opt_stylize(style = 1, color = "blue")
```

```{r discordant-listing}
## Which subjects disagree, with the third assay and AD biomarkers to arbitrate.
disc <- both_assays %>% filter(!concordant) %>%
  transmute(NHC, visit = visit_any, sample = sample_d, gap_d = date_diff,
            Munich = saa_munich_pos, SantPau = saa_local_pos, Third = saa_third,
            AT_profile, datscan_bin, psg_rbd, rbdclinical, parkinsonism,
            hallucinations)
if (nrow(disc)) {
  write.csv(disc, file.path(results_path, "SAA_discordant_subjects.csv"), row.names = FALSE)
  kable(disc, caption = "Discordant subjects — review individually.") %>%
    kable_styling(full_width = FALSE, bootstrap_options = c("striped","condensed"),
                  font_size = 11)
} else cat("No discordant pairs.")
```

# Summary of methodological choices

```{r summary-notes, results='asis'}
cat(sprintf("
- **Scope.** Sant Pau (SAA02) only, clinical variables and A/T status. Other
  biofluid biomarkers and the MRI analyses are parked; they are recoverable from
  the git history of this file.
- **Merge.** Each clinical visit paired with the biofluid sample closest in time
  (`visit_date`, not the baseline date). %s
- **Long-format REDCap.** Subject-level fields carried down within NHC before
  collapsing to one row per subject, so core symptoms and visit scores coexist.
  The progression section deliberately uses all visits instead.
- **Reading.** `guess_max = Inf`, without which readxl types the visit-level
  columns from the empty enrolment rows and discards their data.
- **Recoding.** Anchored SAA patterns and substring A/T patterns; every raw level
  is printed above for checking. Indeterminate SAA excluded (n = %d).
- **Plausibility limits.** Values outside each scale range are set to NA and
  counted; the guard finds nothing in this dataset and remains as a regression
  check.
- **Tests.** Rank-based for continuous, Fisher for categorical, with SMD next to
  the mean or risk difference. No multiplicity adjustment: the comparisons are
  pre-specified and direction-led.
", merge_note, n_indet))
```

# References

Found with Undermind and read against the results above. The note on each line
is what it says about our finding, not a summary of the paper.

**Clinical features**

1. Coughlin DG, MacLeod KR, Middleton JS, et al. *Association of CSF
   alpha-Synuclein Seeding Amplification Assay Results With Clinical Features of
   Possible and Probable Dementia With Lewy Bodies.* Neurology 2024;103(3):e209656.
   doi:10.1212/WNL.0000000000209656
   — n = 191 DLB. RBD in 86% of SAA+ vs 62% of SAA- (p < 0.0001, OR 3.63):
   our RBD result replicates with a smaller effect. Hallucinations 50% vs 47%
   (p = 0.73): our null replicates theirs. MDS-UPDRS III 33.8 vs 25.6
   (p = 0.001): **we do not replicate this** — our UPDRS III and parkinsonism
   prevalence are flat. DaT-SPECT was not analysed against SAA status, so our
   DaT result has no direct published comparator.

**AD co-pathology**

2. Honig LS, Kang MS, Lamoureux J, Marder K. *Cerebrospinal fluid seed
   aggregation assay and Alzheimer biomarkers in a Dementia with Lewy Body
   cohort.* Alzheimers Dement 2025;21(Suppl 2):e107531. doi:10.1002/alz70856_107531
   — No difference in amyloid or tau biomarker status between SAA+ and SAA-.
   Our A/T null replicates this.

3. Burton EJ, Barber R, Mukaetova-Ladinska EB, et al. *Medial temporal lobe
   atrophy on MRI differentiates Alzheimer's disease from dementia with Lewy
   bodies and vascular cognitive impairment.* Brain 2009;132(1):195-203.
   doi:10.1093/brain/awn298
   — The classic MTA result, DLB vs AD rather than SAA-stratified. Our visual
   MTA finding (higher in SAA-) sits in this tradition.

**Where this leaves us.** RBD and hallucinations replicate. Parkinsonism does
not. DaT-SCAN against SAA status appears not to have been reported, and is the
most novel of the clinical results.

# Reproducibility

```{r session}
sessionInfo()
```